---
title: "Figure 2 mock-up: ensembles vs. single maps"
date: last-modified
number-sections: true
---

A backcast on a synthetic 30 × 30 landscape. Domain and classes follow the evoland-plus vignette
`stochastic-allocation-sensitivity.qmd`; unlike the vignette, the initial map is structured
(urban clustered where accessible, forest on better sites) and change between periods follows a known process driven by two latent drivers and the neighbourhood,
so that there is something to learn. Calibrate on the transition from period 1 to 2, allocate
period 3 from the observed period 2 `n_realisations` times with CLUMPY, and validate every
realisation against the observed period 3, which calibration never sees.

Keeping the backcast honest:

- Periods 1 and 2 are observed, period 3 is flagged as extrapolated. Model fitting and
  allocation-parameter estimation therefore only see the transition 1 → 2.
- The observed period 3 sits in a run of its own (`id_run_observed`), a child of the base run.
  The ensemble runs are siblings of it and cannot read it; the figure-of-merit view reads it
  as the reference.
- The demand for period 3 is the observed quantity of each transition from 2 to 3, as in the
  comparison protocol: the figure is about *where* change is placed, not how much.

```{r}
#| label: setup
#| output: false
library(evoland)
library(data.table)
library(terra)
library(ggplot2)
library(patchwork)
library(mlr3learners)

out_dir <- "2026-09-paper-figures/figures"
db_path <- "2026-09-paper-figures/fig2-ensembles.evolanddb"
unlink(db_path, recursive = TRUE)

n_realisations <- 100L
id_run_observed <- 1000L
```

# Database, domain and synthetic landscape

```{r}
#| label: create-db
db <- evoland_db$new(path = db_path)

db$lulc_meta_t <- create_lulc_meta_t(list(
  forest = list(pretty_name = "Forest"),
  arable = list(pretty_name = "Arable Land"),
  urban = list(pretty_name = "Urban Areas"),
  static = list(pretty_name = "Immutable")
))

template_rast <- terra::rast(
  crs = "EPSG:2056",
  extent = terra::ext(c(
    xmin = 2697000, xmax = 2697000 + 30 * 100,
    ymin = 1252000, ymax = 1252000 + 30 * 100
  )),
  resolution = 100
)
n_cells <- terra::ncell(template_rast)

db$coords_t <- create_coords_t_square(
  epsg = 2056L,
  extent = terra::ext(template_rast),
  resolution = terra::res(template_rast)[1]
)

# periods 1, 2 observed; period 3 held out as "extrapolated"
db$periods_t <- create_periods_t(
  period_length_str = "P10Y",
  start_observed = "2000-01-01",
  end_observed = "2010-01-01",
  end_extrapolated = "2020-01-01"
)
db$periods_t
```

```{r}
#| label: drivers
set.seed(321)
scale01 <- function(x) {
  rng <- terra::global(x, c("min", "max"), na.rm = TRUE)
  (x - rng[1, 1]) / (rng[1, 2] - rng[1, 1])
}
smooth_field <- function(template, sd = 1, w = 7) {
  terra::setValues(template, rnorm(terra::ncell(template), sd = sd)) |>
    terra::focal(w = w, fun = mean, na.rm = TRUE) |>
    scale01()
}

xy <- terra::crds(template_rast, df = TRUE)
x_grad <- terra::setValues(template_rast, (xy$x - min(xy$x)) / (max(xy$x) - min(xy$x)))
y_grad <- terra::setValues(template_rast, (xy$y - min(xy$y)) / (max(xy$y) - min(xy$y)))
accessibility <- scale01(0.55 * (1 - x_grad) + 0.25 * (1 - y_grad) + 0.20 * smooth_field(template_rast, w = 9))
site_quality <- scale01(0.50 * y_grad + 0.35 * smooth_field(template_rast, w = 5) + 0.15 * x_grad)
random_nuisance <- smooth_field(template_rast, sd = 1, w = 3)
```

The true process. Per period, each cell of an anterior class draws at most one transition with
these probabilities, where `share_k` is the share of class k among the 5 × 5 neighbourhood:

| transition | logit of the per-period probability |
|---|---|
| arable → urban | −5.5 + 7 · share_urban + 3 · accessibility |
| forest → urban | −7 + 6 · share_urban + 3 · accessibility |
| forest → arable | −5 + 4 · (1 − site_quality) + 3 · share_arable |
| arable → forest | −6 + 4 · site_quality · (1 − accessibility) + 3 · share_forest |

```{r}
#| label: synthesize-lulc
#| fig-asp: 0.3
set.seed(123)
# initial landscape: urban clustered where accessible, a small immutable lake, forest on the
# better sites, arable on the rest
urban_score <- scale01(0.7 * accessibility + 0.3 * smooth_field(template_rast, w = 3))
lake_score <- smooth_field(template_rast, w = 7)
forest_score <- scale01(0.6 * smooth_field(template_rast, w = 5) + 0.4 * site_quality)
q <- function(r, p) stats::quantile(terra::values(r), p, na.rm = TRUE)
initial <- terra::ifel(
  urban_score > q(urban_score, 0.88), 3,
  terra::ifel(
    lake_score > q(lake_score, 0.97), 4,
    terra::ifel(forest_score > q(forest_score, 0.45), 1, 2)
  )
)

plogis_rast <- function(x) 1 / (1 + exp(-x))
neighbour_share <- function(map, class) {
  terra::focal(map == class, w = 5, fun = mean, na.rm = TRUE)
}

step_process <- function(map) {
  share_urban <- neighbour_share(map, 3)
  share_arable <- neighbour_share(map, 2)
  share_forest <- neighbour_share(map, 1)
  probs <- list(
    list(from = 2, to = 3, p = plogis_rast(-5.5 + 7 * share_urban + 3 * accessibility)),
    list(from = 1, to = 3, p = plogis_rast(-7 + 6 * share_urban + 3 * accessibility)),
    list(from = 1, to = 2, p = plogis_rast(-5 + 4 * (1 - site_quality) + 3 * share_arable)),
    list(from = 2, to = 1, p = plogis_rast(-6 + 4 * site_quality * (1 - accessibility) + 3 * share_forest))
  )
  anterior <- terra::values(map, mat = FALSE)
  posterior <- anterior
  draw <- runif(length(anterior))
  cumulative <- numeric(length(anterior))
  for (tr in probs) {
    p <- terra::values(tr$p, mat = FALSE)
    at_risk <- anterior == tr$from
    flips <- at_risk & draw >= cumulative & draw < cumulative + p & posterior == anterior
    posterior[flips] <- tr$to
    cumulative[at_risk] <- cumulative[at_risk] + p[at_risk]
  }
  terra::setValues(map, posterior)
}

lulc_2 <- step_process(initial)
lulc_3 <- step_process(lulc_2)
synthetic_lulc <- c(initial, lulc_2, lulc_3) |> setNames(paste0("id_period=", 1:3))

plot(
  synthetic_lulc,
  nc = 3,
  col = data.frame(value = 1:4, color = c("#91B690", "#EB9486", "#F3DE8A", "#CBC6D2"))
)

lulc_long <- extract_using_coords_t(synthetic_lulc, db$coords_t)[,
  .(id_coord, id_period = as.integer(sub(".*[^0-9]", "", layer)), id_lulc = value)
]
lulc_long[, .N, by = .(id_period, id_lulc)][order(id_period, id_lulc)]

db$lulc_data_t <- as_lulc_data_t(lulc_long[id_period <= 2L, .(id_run = 0L, id_coord, id_period, id_lulc)])
```

# Predictors

The model sees the two drivers, a nuisance field, and evoland's neighbourhood predictors
(computed from the land use map), but not the functional form above.

```{r}
#| label: predictors
db$pred_meta_t <- create_pred_meta_t(list(
  accessibility = list(description = "synthetic driver", data_type = "float"),
  site_quality = list(description = "synthetic driver", data_type = "float"),
  random_nuisance = list(description = "synthetic nuisance", data_type = "float")
))

db$pred_data_t <-
  extract_using_coords_t(
    c(accessibility, site_quality, random_nuisance) |>
      setNames(c("accessibility", "site_quality", "random_nuisance")),
    db$coords_minimal
  )[, layer := as.character(layer)][,
    .(
      id_coord,
      id_run = 0L,
      id_period = 0L,
      id_pred = fcase(
        layer == "accessibility", 1L,
        layer == "site_quality", 2L,
        layer == "random_nuisance", 3L
      ),
      value
    )
  ] |>
  as_pred_data_t()

db$set_neighbors(max_distance = 1000, distance_breaks = c(0, 300, 1000), quiet = TRUE)
db$generate_neighbor_predictors()
```

# Calibration on 1 → 2

```{r}
#| label: calibrate
set.seed(666)
db$trans_meta_t <- create_trans_meta_t(
  db$trans_v,
  min_cardinality_abs = 10,
  exclude_anterior = 4
)
db$set_full_trans_preds()

trans_pred_scored <- db$get_pred_filter_score(
  filter = mlr3filters::FilterImportance$new(learner = mlr3::lrn("classif.rpart"))
)
db$commit(
  trans_pred_scored[order(-importance)][, head(.SD, 4), by = id_trans],
  "trans_preds_t",
  method = "overwrite"
)

trans_models <- db$fit_full_models(learner = mlr3::lrn("classif.ranger", num.trees = 200))
modeled_trans <- unique(
  trans_models$id_trans[!vapply(trans_models$learner_full, is.null, logical(1L))]
)
db$trans_models_t <- trans_models
trans_meta <- db$trans_meta_t
trans_meta[, is_viable := is_viable & id_trans %in% modeled_trans]
db$trans_meta_t <- trans_meta

db$alloc_params_t <- db$create_alloc_params_t()
db$trans_meta_t[is_viable == TRUE]
```

# Demand: observed quantities 2 → 3

```{r}
#| label: demand
observed_2_3 <-
  lulc_long[id_period == 2L, .(id_coord, id_lulc_anterior = id_lulc)][
    lulc_long[id_period == 3L, .(id_coord, id_lulc_posterior = id_lulc)],
    on = "id_coord"
  ]
anterior_totals <- observed_2_3[, .(n_anterior = .N), by = id_lulc_anterior]
observed_counts <- observed_2_3[
  id_lulc_anterior != id_lulc_posterior,
  .(count = .N),
  by = .(id_lulc_anterior, id_lulc_posterior)
]

rates_3 <-
  db$trans_meta_t[is_viable == TRUE, .(id_trans, id_lulc_anterior, id_lulc_posterior)][
    observed_counts,
    on = .(id_lulc_anterior, id_lulc_posterior),
    nomatch = NULL
  ][anterior_totals, on = "id_lulc_anterior", nomatch = NULL][,
    .(id_run = 0L, id_period = 3L, id_trans, count, rate = count / n_anterior)
  ]
db$trans_rates_t <- as_trans_rates_t(rates_3)

# change the model cannot produce: transitions not viable after 1 -> 2
unmodelled_change <- observed_counts[, sum(count)] - rates_3[, sum(count)]
rates_3
unmodelled_change
```

# Runs: the held-out observation and the ensemble

```{r}
#| label: runs
runs <- data.table(
  id_run = c(0L, id_run_observed, seq_len(n_realisations)),
  parent_id_run = c(NA_integer_, 0L, rep(0L, n_realisations)),
  description = c(
    "calibration base",
    "observed period 3 (validation only)",
    paste("CLUMPY realisation", seq_len(n_realisations))
  ),
  seed = c(NA_integer_, NA_integer_, 1000L + seq_len(n_realisations))
)
db$commit(as_runs_t(runs), "runs_t", method = "overwrite")

db$lulc_data_t <- as_lulc_data_t(
  lulc_long[id_period == 3L, .(id_run = id_run_observed, id_coord, id_period, id_lulc)]
)
```

```{r}
#| label: allocate
#| output: false
for (id in seq_len(n_realisations)) {
  db$id_run <- id
  set.seed(runs[id_run == id, seed])
  db$alloc_clumpy(
    id_periods = 3,
    select_score = "no.crossval",
    select_maximize = TRUE,
    use_parent_trans_pot = TRUE,
    update_neighbors = FALSE
  )
}
```

# A deterministic comparator on the same potentials

Tools without stochastic allocation return one map. As a stand-in until the real
comparators run (`2026-09-model-comparison/`), allocate the same demand deterministically on
the same adjusted potentials: greedily, highest potential first, one class per cell. This
isolates the allocator's contribution; it says nothing about how Dinamica or LCM would score.

```{r}
#| label: deterministic
db$id_run <- 0L
adjusted <- db$adjusted_trans_pot_v(3L)
str(adjusted)
```

```{r}
#| label: deterministic-alloc
pot_col <- intersect(c("value", "trans_pot", "potential", "prob"), names(adjusted))[1]
stopifnot("unknown adjusted potential column" = !is.na(pot_col))

candidates <-
  adjusted[, .(id_coord, id_trans, potential = get(pot_col))][
    db$trans_meta_t[, .(id_trans, id_lulc_posterior)],
    on = "id_trans",
    nomatch = NULL
  ][rates_3[, .(id_trans, quota = count)], on = "id_trans", nomatch = NULL][order(-potential)]

remaining <- setNames(candidates[, quota[1], by = id_trans]$V1, candidates[, unique(id_trans)])
assigned <- new.env()
greedy <- candidates[, {
  keep <- logical(.N)
  for (i in seq_len(.N)) {
    key <- as.character(id_coord[i])
    tr <- as.character(id_trans[i])
    if (remaining[[tr]] > 0 && is.null(assigned[[key]])) {
      assigned[[key]] <- TRUE
      remaining[[tr]] <- remaining[[tr]] - 1L
      keep[i] <- TRUE
    }
  }
  .(id_coord = id_coord[keep], id_lulc = id_lulc_posterior[keep])
}]

deterministic_map <- lulc_long[id_period == 2L, .(id_coord, id_lulc)]
deterministic_map[greedy, on = "id_coord", id_lulc := i.id_lulc]
```

# Validation

```{r}
#| label: fom
db$id_run <- NULL
fom <- db$figure_of_merit_v(
  id_period_anterior = 2L,
  id_period_post = 3L,
  id_run_reference = id_run_observed,
  id_run_simulated = seq_len(n_realisations)
)

# expected counts of the ensemble: ratio of mean counts, not mean of ratios
fom_ensemble <- fom[, lapply(.SD, mean), .SDcols = c("hits", "wrong_hits", "misses", "false_alarms")][,
  figure_of_merit := hits / (hits + wrong_hits + misses + false_alarms)
]

fom_counts <- function(anterior, observed, simulated) {
  obs_change <- anterior != observed
  sim_change <- anterior != simulated
  hits <- sum(obs_change & sim_change & simulated == observed)
  wrong <- sum(obs_change & sim_change & simulated != observed)
  misses <- sum(obs_change & !sim_change)
  false_alarms <- sum(!obs_change & sim_change)
  hits / (hits + wrong + misses + false_alarms)
}
maps <- lulc_long[id_period == 2L, .(id_coord, anterior = id_lulc)][
  lulc_long[id_period == 3L, .(id_coord, observed = id_lulc)],
  on = "id_coord"
][deterministic_map[, .(id_coord, deterministic = id_lulc)], on = "id_coord"]
fom_deterministic <- maps[, fom_counts(anterior, observed, deterministic)]

# the ensemble as a probabilistic forecast of change, against single maps (Brier score, lower
# is better); needs the realisations, so fetched here rather than in the figure section
simulated_all <- db$fetch("lulc_data_t", where = glue::glue("id_run between 1 and {n_realisations} and id_period = 3"))[,
  .(id_run, id_coord, simulated = id_lulc)
]
brier <- maps[, .(id_coord, anterior, observed, deterministic)][
  simulated_all[, .(id_run, id_coord, simulated)], on = "id_coord"
][, .(
  p_ensemble = mean(simulated != anterior),
  y = as.numeric(observed[1] != anterior[1]),
  deterministic = as.numeric(deterministic[1] != anterior[1])
), by = id_coord]
brier_draws <- maps[, .(id_coord, anterior, observed)][simulated_all, on = "id_coord"][,
  .(brier = mean((as.numeric(simulated != anterior) - as.numeric(observed != anterior))^2)),
  by = id_run
]

summary_tab <- data.table(
  quantity = c(
    "realisations: median FoM", "realisations: 5-95 %",
    "ensemble expected FoM", "deterministic greedy", "random-allocation null",
    "Brier: ensemble change frequency", "Brier: single realisations (median)", "Brier: deterministic"
  ),
  value = c(
    sprintf("%.3f", median(fom$figure_of_merit)),
    paste(sprintf("%.3f", quantile(fom$figure_of_merit, c(0.05, 0.95))), collapse = " to "),
    sprintf("%.3f", fom_ensemble$figure_of_merit),
    sprintf("%.3f", fom_deterministic),
    sprintf("%.3f", mean(fom$figure_of_merit_null)),
    sprintf("%.4f", brier[, mean((p_ensemble - y)^2)]),
    sprintf("%.4f", median(brier_draws$brier)),
    sprintf("%.4f", brier[, mean((deterministic - y)^2)])
  )
)
knitr::kable(summary_tab)
```

# Figure

Map panels are coloured by the figure-of-merit outcome rather than land-use class: the figure
is about where change goes, and the outcome classes are what the metric counts.

```{r}
#| label: figure-data
simulated <- maps[, .(id_coord, anterior, observed)][simulated_all, on = "id_coord"]

coords_xy <- db$coords_t[, .(id_coord, x = lon, y = lat)]

outcome <- function(anterior, observed, simulated) {
  obs_change <- anterior != observed
  sim_change <- anterior != simulated
  fcase(
    obs_change & sim_change & simulated == observed, "hit",
    obs_change & sim_change, "wrong hit",
    obs_change, "miss",
    sim_change, "false alarm",
    default = "persistence"
  )
}

# the realisation shown is the one closest to the ensemble median
shown_run <- fom[order(abs(figure_of_merit - median(figure_of_merit)))][1, id_run]

panel_obs <- maps[, .(id_coord, status = fifelse(anterior != observed, "observed change", "persistence"))]
panel_draw <- simulated[id_run == shown_run, .(id_coord, status = outcome(anterior, observed, simulated))]
panel_det <- maps[, .(id_coord, status = outcome(anterior, observed, deterministic))]
panel_freq <- simulated[, .(p_change = mean(simulated != anterior)), by = id_coord]
observed_change_cells <- maps[anterior != observed, .(id_coord)]
```

```{r}
#| label: figure
#| fig-width: 7.2
#| fig-height: 5.6
col_surface <- "#f1f0ed"
outcome_cols <- c(
  "persistence" = col_surface,
  "observed change" = "#184f95",
  "hit" = "#2a78d6",
  "miss" = "#eb6834",
  "false alarm" = "#1baf7a",
  "wrong hit" = "#3d3d3d"
)
seq_blue <- c("#fcfcfb", "#cde2fb", "#86b6ef", "#3987e5", "#1c5cab", "#0d366b")

theme_map <- theme_void(base_size = 9) +
  theme(
    plot.title = element_text(size = 9, face = "bold", hjust = 0, margin = margin(b = 3)),
    plot.subtitle = element_text(size = 8, colour = "#555555", margin = margin(b = 4)),
    plot.margin = margin(2, 8, 2, 8),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key.size = unit(8, "pt"),
    legend.text = element_text(size = 7.5)
  )

map_panel <- function(d, title, subtitle = NULL) {
  ggplot(coords_xy[d, on = "id_coord"], aes(x, y, fill = status)) +
    geom_tile(colour = "white", linewidth = 0.15) +
    scale_fill_manual(values = outcome_cols, breaks = intersect(names(outcome_cols)[-1], unique(d$status))) +
    coord_equal(expand = FALSE) +
    labs(title = title, subtitle = subtitle) +
    theme_map
}

p_a <- map_panel(panel_obs, "(a) Observed change", "period 2 to 3, held out")
p_b <- map_panel(
  panel_draw,
  "(b) One realisation",
  sprintf("median draw, FoM = %.2f", fom[id_run == shown_run, figure_of_merit])
)
p_c <- ggplot(coords_xy[panel_freq, on = "id_coord"], aes(x, y)) +
  geom_tile(aes(fill = p_change), colour = "white", linewidth = 0.15) +
  geom_tile(
    data = coords_xy[observed_change_cells, on = "id_coord"],
    fill = NA, colour = "#1a1a1a", linewidth = 0.45, width = 70, height = 70
  ) +
  scale_fill_gradientn(
    colours = seq_blue, limits = c(0, 1), breaks = c(0, 0.5, 1),
    name = "share of realisations",
    guide = guide_colourbar(title.position = "top", title.hjust = 0.5)
  ) +
  coord_equal(expand = FALSE) +
  labs(
    title = "(c) Change frequency",
    subtitle = "outline: observed change"
  ) +
  theme_map +
  theme(
    legend.title = element_text(size = 7.5),
    legend.key.width = unit(22, "pt"),
    legend.key.height = unit(6, "pt")
  )

fom_null <- mean(fom$figure_of_merit_null)
refs <- data.table(
  label = c("persistence", "random allocation", "deterministic, same potentials", "ensemble expected"),
  value = c(0, fom_null, fom_deterministic, fom_ensemble$figure_of_merit)
)
p_d <- ggplot(fom, aes(x = figure_of_merit, y = 0)) +
  geom_vline(xintercept = 0, colour = "#9a9a9a", linewidth = 0.4) +
  geom_vline(xintercept = fom_null, colour = "#9a9a9a", linewidth = 0.4, linetype = "22") +
  geom_violin(fill = "#cde2fb", colour = NA, width = 0.7) +
  geom_jitter(height = 0.18, width = 0, size = 1.1, colour = "#2a78d6", alpha = 0.7) +
  geom_point(data = refs[3], aes(x = value, y = 0), shape = 23, size = 3, fill = "#eb6834", colour = "white", stroke = 0.8) +
  geom_point(data = refs[4], aes(x = value, y = 0), shape = 21, size = 3, fill = "#0d366b", colour = "white", stroke = 0.8) +
  annotate("text", x = 0, y = 0.66, label = " persistence", hjust = 0, size = 2.5, colour = "#555555") +
  annotate("text", x = fom_null, y = 0.66, label = " random allocation", hjust = 0, size = 2.5, colour = "#555555") +
  annotate("segment", x = refs[4, value], xend = refs[4, value], y = -0.08, yend = -0.36, colour = "#0d366b", linewidth = 0.3) +
  annotate("text", x = refs[4, value], y = -0.42, label = "ensemble expectation ", hjust = 1, size = 2.5, colour = "#0d366b") +
  annotate("segment", x = refs[3, value], xend = refs[3, value], y = -0.08, yend = -0.5, colour = "#eb6834", linewidth = 0.3) +
  annotate("text", x = refs[3, value], y = -0.56, label = "deterministic allocation, same potentials ", hjust = 1, size = 2.5, colour = "#b8461c") +
  scale_y_continuous(limits = c(-0.62, 0.72), breaks = NULL) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0.02, 0.08))) +
  labs(
    title = "(d) Figure of merit against the held-out observation",
    subtitle = sprintf("one dot per realisation (n = %d)", n_realisations),
    x = "figure of merit", y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(
    plot.title = element_text(size = 9, face = "bold", margin = margin(b = 3)),
    plot.subtitle = element_text(size = 8, colour = "#555555", margin = margin(b = 4)),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(colour = "#e6e6e6", linewidth = 0.3)
  )

maps_row <- (p_a | p_b) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
fig2 <- (maps_row | p_c) / p_d + plot_layout(heights = c(1.3, 1), widths = c(2, 1))
ggsave(file.path(out_dir, "fig2-ensembles.pdf"), fig2, width = 7.2, height = 5.6, device = cairo_pdf)
ggsave(file.path(out_dir, "fig2-ensembles.png"), fig2, width = 7.2, height = 5.6, dpi = 200, bg = "white")
fig2
```
