---
title: "00 — Setup DB"
---

Sets up the SSP-CH experiment database: an `evoland_db` store aimed at Arealstatistik land-use plus
the Swiss predictor stack. This step defines the three axes that later steps key against:

- **coordinate grid** (`id_coord`)
- **periods** (`id_period`)
- **runs** (`id_run`)


```{r}
#| label: setup-db-create
library(evoland)

db <- evoland_db$new(
  path = "ssp-ch.evolanddb",
  report_name = "ssp-ch",
  report_name_pretty = "Replication of SSP-CH in new evoland"
)
```

## Coordinate grid

A square 100 m grid in **CH1903+ / LV95 (EPSG:2056)** over the full Swiss bounding box.

```{r}
#| label: setup-db-coords
db$coords_t <- create_coords_t_square(
  epsg = 2056,
  extent = terra::ext(c(
    xmin = 2480000,
    xmax = 2840000,
    ymin = 1070000,
    ymax = 1300000
  )),
  resolution = 100
)
```

## Periods

Observed Arealstatistik survey years followed by extrapolated steps out to 2100.

```{r}
#| label: setup-db-periods
db$periods_t <- as_periods_t(data.table::rowwiseDT(
  id_period = , start_date = , end_date = , is_extrapolated = ,
  0, "2020-01-01", "2020-01-01", FALSE,
  1, "1979-01-01", "1985-12-31", FALSE,
  2, "1992-01-01", "1992-12-31", FALSE,
  3, "2004-01-01", "2009-12-31", FALSE,
  4, "2013-01-01", "2018-12-31", FALSE,
  5, "2020-01-01", "2029-12-31", TRUE,
  6, "2030-01-01", "2039-12-31", TRUE,
  7, "2040-01-01", "2049-12-31", TRUE,
  8, "2050-01-01", "2059-12-31", TRUE,
  9, "2060-01-01", "2069-12-31", TRUE,
  10, "2070-01-01", "2079-12-31", TRUE,
  11, "2080-01-01", "2089-12-31", TRUE,
  12, "2090-01-01", "2099-12-31", TRUE
))
```

This yields 11 decadal periods plus the static `id_period 0` for static predictors (DEM, EIVs, distances, climate baseline).

## Runs. Setting up scenarios.

`runs_t` is the third axis, alongside coordinates and periods. Runs form a **hierarchy**:
`db$get_read_expr()` resolves each data slice to the closest ancestor run that has it, so a
run only needs to store what it overrides. Everything else (calibration, observed LULC,
static predictors) is inherited from the base run `0`.

Because the tree replicates whatever sits **below** a branch point, the ordering of the two
scenario axes is a storage decision.
The technical ordering may not match the storytelling. A run's `description` reads
"SSP3 under …", because that is how results get discussed.

```
0  base (observed, calibration)
└── climate trajectory  (GWL schedule × uncertainty quantile)   ← expensive: 4.1 M cells × 4 periods × N predictors
    └── SSP                (demand curves, STATENT, interventions)   ← cheap: ~10²–10³ rows
```

> **Scenario scope.** SSP0/1/3/4/5; SSP2 is deliberately excluded (see `README.md`). The
> per-SSP demand targets live in `NCCS-SSP-scenarios/Tools/NCCS_simulation_LULC_areas.xlsx`,
> keyed by these same labels.

### GWL trajectories

CH2025 publishes exactly three levels (GWL1.5 / GWL2.0 / GWL3.0) as 30-year climatologies. The four
schedules below are a defensible reading of IPCC AR6 WG1 SPM.8a. See
`02-ingest-preds-ch2025-3-gwl.qmd` which consumes them.
`fast30` is **capped** — AR6 puts SSP5-8.5 above > 3 °C well before the 2060s, but CH2025 publishes
nothing higher.

```{r}
#| label: setup-db-runs-spec

# Trajectory order fixes the id_run digit; default_for_ssp is reporting metadata only.
trajectory_specs <- data.table::rowwiseDT(
  gwl_trajectory = , traj_idx = , default_for_ssp = , traj_description = ,
  "stab15", 1L, "SSP0", "Held at 1.5 C throughout",
  "stab20", 2L, "SSP1", "1.5 C, stabilising at 2.0 C",
  "rise30", 3L, "SSP3,SSP4", "1.5 C rising to 3.0 C",
  "fast30", 4L, "SSP5", "2.0 C reaching 3.0 C by mid-century (capped)"
)

ssp_specs <- data.table::rowwiseDT(
  ssp = , ssp_idx = , ssp_description = ,
  "SSP0", 1L, "Positive normative visioning",
  "SSP1", 2L, "Sustainability",
  "SSP3", 3L, "Regional rivalry",
  "SSP4", 4L, "Inequality",
  "SSP5", 5L, "Fossil-fuelled development"
)

# CH2025 uncertainty arms to register. q50 alone by default; add "q5"/"q95" for the
# uncertainty sensitivity -- each one multiplies stored climate volume.
gwl_quantiles <- data.table::data.table(gwl_quantile = "q50", q_idx = 2L)
```

The NCCS-SSP framing consciously kept SSP and RCP/GWL orthogonal, so steady climate runs are also included. 

```{r}
#| label: setup-db-runs
climate_runs <- rbind(
  data.table::data.table(
    id_run = 100L,
    parent_id_run = 0L,
    description = "Current climatology (no projection; inherits the id_period 0 baseline)",
    climate_framing = "current",
    gwl_trajectory = NA_character_,
    gwl_quantile = NA_character_
  ),
  trajectory_specs[
    ,
    .(gwl_quantile = gwl_quantiles$gwl_quantile, q_idx = gwl_quantiles$q_idx),
    by = .(gwl_trajectory, traj_idx, traj_description)
  ][, .(
    id_run = 1000L + 100L * traj_idx + q_idx,
    parent_id_run = 0L,
    description = paste0(
      "CH2025 GWL trajectory '", gwl_trajectory, "' (", traj_description, "), ", gwl_quantile
    ),
    climate_framing = "gwl",
    gwl_trajectory,
    gwl_quantile
  )]
)

# SSP leaves: the full cross of SSP x climate run. Leaves are cheap -- only the pairings
# actually allocated cost anything downstream.
ssp_runs <-
  climate_runs[
    ,
    .(ssp_idx = ssp_specs$ssp_idx),
    by = .(climate_id_run = id_run)
  ][
    climate_runs,
    on = c(climate_id_run = "id_run"),
    nomatch = NULL
  ][
    ssp_specs,
    on = "ssp_idx",
    nomatch = NULL
  ][, .(
    id_run = climate_id_run * 10L + ssp_idx,
    parent_id_run = climate_id_run,
    description = paste0(ssp, " (", ssp_description, ") under ", description),
    ssp,
    climate_framing,
    gwl_trajectory,
    gwl_quantile
  )]

db$commit(
  as_runs_t(rbind(
    db$runs_t, # the base run 0 created on init
    climate_runs,
    ssp_runs,
    fill = TRUE
  )), 
  "runs_t",
  "overwrite" # need this because we're adding columns
)
```
