---
title: "00 — Setup DB"
---

Sets up the SSP-CH experiment database: an `evoland_db` (DuckDB) backing store aimed at
Arealstatistik land-use plus the Swiss predictor stack. This step defines the two axes
that every later step keys against — the **coordinate grid** (`id_coord`) and the
**periods** (`id_period`) — so it must run first.

```{r}
#| label: setup-db-create
library(evoland)

db <- evoland_db$new(
  path = "ssp-ch.evolanddb",
  report_name = "ssp-ch",
  report_name_pretty = "Replication of SSP-CH in new evoland"
)
```

## Coordinate grid

A square 100 m grid in **CH1903+ / LV95 (EPSG:2056)** over the full Swiss bounding box.
The commented-out inner extent is a small central-Switzerland window kept for quick local
test runs; the active extent is the whole country.

```{r}
#| label: setup-db-coords
db$coords_t <- create_coords_t_square(
  epsg = 2056,
  extent = terra::ext(c(
    # somewhere in the middle of the country
    # xmin = 2697000,
    # xmax = 2698000,
    # ymin = 1252000,
    # ymax = 1253000

    # full extent
    xmin = 2480000,
    xmax = 2840000,
    ymin = 1070000,
    ymax = 1300000
  )),
  resolution = 100
)
```

## Periods

Decadal steps (`P10Y`): observed 1985–2020 (the Arealstatistik survey years) followed by
extrapolated steps out to 2060.

```{r}
#| label: setup-db-periods
db$periods_t <- create_periods_t(
  period_length_str = "P10Y",
  start_observed = "1985-01-01",
  end_observed = "2020-01-01",
  end_extrapolated = "2060-01-01"
)
```

This yields eight decadal periods plus the static `id_period 0`:

| `id_period` | Span | Role |
| --- | --- | --- |
| 0 | 2020-01-01 (instant) | static predictors (DEM, EIVs, distances, climate baseline) |
| 1–4 | 1985–1994 … 2015–2024 | observed; carry the Arealstatistik surveys 1985/1997/2009/2018 |
| 5–8 | 2025–2034 … 2055–2064 | extrapolated |

**Note the last period ends 2064, not 2060.** Decadal steps from 1985 cannot land on 2060,
so the SSP demand targets (stated for 2060) fall *inside* period 8. Step `07` resamples the
5-yearly demand curves onto this decadal grid and must therefore treat period 8's target as
an interpolation, not an endpoint. This also depends on the `create_periods_t` fix that makes
the final period a full decade — see the pin note in `rproject.toml`; before that fix period 8
was a half-decade (2055–2060) and any rate applied to it was effectively doubled.

## Runs — the scenario axis

`runs_t` is the third axis, alongside coordinates and periods. Runs form a **hierarchy**:
`db$get_read_expr()` resolves each data slice to the closest ancestor run that has it, so a
scenario run only needs to store what it *overrides* — everything else (calibration,
observed LULC, static predictors) is inherited from the base run `0`.

We use three levels:

1. **Base (`0`)** — observed data and everything fitted from it.
2. **Socioeconomic pathway** — one run per SSP. Carries what the SSP narrative changes:
   demand curves (`07`), projected STATENT employment, and interventions.
3. **Climate framing** — children of an SSP run, keeping socioeconomic and climatological
   pathways orthogonal as in the original SSP-CH work:
   - `current` — no climate override at all; every period inherits the observed
     climatology at `id_period 0`. This is the "what if only society changes" arm.
   - `gwl` — CH2025 projections at the mapped global warming level, one run per
     uncertainty quantile (`q5`/`q50`/`q95`).

`id_run` is `1000 + 100 × ssp_num`, `+1` for the current-climatology arm and `+11/12/13`
for the GWL quantiles — so `1312` reads as "SSP3, GWL framing, q50". Downstream steps should
nonetheless **look runs up by the `ssp`/`climate_framing`/`gwl_quantile` columns** rather than
hard-coding integers. `runs_t` only mandates `id_run`/`parent_id_run`/`description`; extra
columns are carried through, which is what makes that lookup possible.

> **Scenario scope.** SSP0/1/3/4/5; SSP2 is deliberately excluded (see `README.md`). Note
> that the per-SSP demand targets live in `NCCS-SSP-scenarios/Tools/NCCS_simulation_LULC_areas.xlsx`,
> keyed by these same SSP labels. The `EI_Nature`/`EI_Culture`/`EI_Society`/`Growth and Extinction`
> names in `Transition_Tables.xlsx` are an **earlier vintage** and do *not* map cleanly onto
> SSP0/1/3/4/5 — matching the two on their 2060 class areas leaves residuals of 3–7 × 10⁵ ha and
> assigns two SSPs to the same name. Don't build a crosswalk between them.

```{r}
#| label: setup-db-runs
library(data.table)

ssp_specs <- data.table::data.table(
  ssp = c("SSP0", "SSP1", "SSP3", "SSP4", "SSP5"),
  ssp_num = c(0L, 1L, 3L, 4L, 5L),
  ssp_description = c(
    "Positive normative visioning; newly added, no counterpart in the original SSP-CH runs",
    "Sustainability",
    "Regional rivalry",
    "Inequality",
    "Fossil-fuelled development"
  )
)

gwl_quantiles <- c("q5", "q50", "q95") # CH2025 uncertainty levels

runs_scenario <-
  rbind(
    # level 2: socioeconomic pathway
    ssp_specs[, .(
      id_run = 1000L + 100L * ssp_num,
      parent_id_run = 0L,
      description = paste0(ssp, " — ", ssp_description),
      ssp,
      climate_framing = NA_character_,
      gwl_quantile = NA_character_
    )],
    # level 3a: current climatology (inherits the id_period 0 baseline; stores no climate)
    ssp_specs[, .(
      id_run = 1000L + 100L * ssp_num + 1L,
      parent_id_run = 1000L + 100L * ssp_num,
      description = paste0(ssp, " × current climatology"),
      ssp,
      climate_framing = "current",
      gwl_quantile = NA_character_
    )],
    # level 3b: CH2025 GWL projections, one run per uncertainty quantile
    ssp_specs[,
      .(gwl_quantile = gwl_quantiles, q_idx = seq_along(gwl_quantiles)),
      by = .(ssp, ssp_num)
    ][, .(
      id_run = 1000L + 100L * ssp_num + 10L + q_idx,
      parent_id_run = 1000L + 100L * ssp_num,
      description = paste0(ssp, " × CH2025 GWL (", gwl_quantile, ")"),
      ssp,
      climate_framing = "gwl",
      gwl_quantile
    )]
  )

db$runs_t <- as_runs_t(rbind(
  db$runs_t, # keep the base run 0 created on init
  runs_scenario,
  fill = TRUE
))

db$runs_t
```
