---
title: "00 — Setup DB"
---

Sets up the SSP-CH experiment database: an `evoland_db` (DuckDB) backing store aimed at
Arealstatistik land-use plus the Swiss predictor stack. This step defines the two axes
that every later step keys against — the **coordinate grid** (`id_coord`) and the
**periods** (`id_period`) — so it must run first.

```{r}
#| label: setup-db-create
library(evoland)

db <- evoland_db$new(
  path = "ssp-ch.evolanddb",
  report_name = "ssp-ch",
  report_name_pretty = "Replication of SSP-CH in new evoland"
)
```

## Coordinate grid

A square 100 m grid in **CH1903+ / LV95 (EPSG:2056)** over the full Swiss bounding box.
The commented-out inner extent is a small central-Switzerland window kept for quick local
test runs; the active extent is the whole country.

```{r}
#| label: setup-db-coords
db$coords_t <- create_coords_t_square(
  epsg = 2056,
  extent = terra::ext(c(
    # somewhere in the middle of the country
    # xmin = 2697000,
    # xmax = 2698000,
    # ymin = 1252000,
    # ymax = 1253000

    # full extent
    xmin = 2480000,
    xmax = 2840000,
    ymin = 1070000,
    ymax = 1300000
  )),
  resolution = 100
)
```

## Periods

Decadal steps (`P10Y`): observed 1985–2020 (the Arealstatistik survey years) followed by
extrapolated steps out to 2060.

```{r}
#| label: setup-db-periods
# FIXME we can now construct this using rowwiseDT to reflect irregular period lengths
db$periods_t <- create_periods_t(
  period_length_str = "P10Y",
  start_observed = "1985-01-01",
  end_observed = "2020-01-01",
  end_extrapolated = "2060-01-01"
)
```

This yields eight decadal periods plus the static `id_period 0`:

| `id_period` | Span | Role |
| --- | --- | --- |
| 0 | 2020-01-01 (instant) | static predictors (DEM, EIVs, distances, climate baseline) |
| 1–4 | 1985–1994 … 2015–2024 | observed; carry the Arealstatistik surveys 1985/1997/2009/2018 |
| 5–8 | 2025–2034 … 2055–2064 | extrapolated |

**Note the last period ends 2064, not 2060.** Decadal steps from 1985 cannot land on 2060,
so the SSP demand targets (stated for 2060) fall *inside* period 8. Step `07` resamples the
5-yearly demand curves onto this decadal grid and must therefore treat period 8's target as
an interpolation, not an endpoint. This also depends on the `create_periods_t` fix that makes
the final period a full decade — see the pin note in `rproject.toml`; before that fix period 8
was a half-decade (2055–2060) and any rate applied to it was effectively doubled.

## Runs — the scenario axis

`runs_t` is the third axis, alongside coordinates and periods. Runs form a **hierarchy**:
`db$get_read_expr()` resolves each data slice to the closest ancestor run that has it, so a
run only needs to store what it *overrides* — everything else (calibration, observed LULC,
static predictors) is inherited from the base run `0`.

Because the tree replicates whatever sits **below** a branch point, the ordering of the two
scenario axes is a storage decision, not just a narrative one:

```
0  base (observed, calibration)
└── climate trajectory  (GWL schedule × uncertainty quantile)   ← expensive: 4.1 M cells × 4 periods × N predictors
    └── SSP                (demand curves, STATENT, interventions)   ← cheap: ~10²–10³ rows
```

Climate goes **above** SSP, because it is the heavier per-run payload. Putting SSP on top —
the obvious reading of the narrative, where a scenario "has" a climate — would store the
climate rasters once per SSP instead.

Be precise about the size of that win, though: projected climate is not the *only* per-run,
per-period predictor family. Projected employment (`02-ingest-preds-statent-ssp.qmd`) is one
too, and since SSP is the leaf it gets replicated once per (SSP × climate) pairing:

| | rows per arm | copies | total |
| --- | --- | --- | --- |
| climate (18 yearly indicators × 4 periods × 4.1 M) | 295 M | 4 trajectories | 1.18 B |
| employment (3 sectors × 4 periods × 4.1 M) | 49 M | 5 SSP × 4 traj | 0.98 B |

That is ~2.2 B rows against ~6.2 B for SSP-on-top — **a factor 2.8**. Climate alone would
suggest a factor 5; employment replication eats the difference. The ordering is still clearly
right (climate is ~6× heavier per arm), but "the SSP payload is tiny" is only true of
`trans_rates_t` and `intrv_meta_t`, not of the projected predictors.

This ordering also expresses the design intent better. The original SSP-CH work kept
climatological and socioeconomic change **orthogonal**, and that only works if a climate
trajectory is a first-class object several SSPs can be realised against, rather than a
property of one SSP. So the GWL schedules below are defined **independently of the SSPs**;
which trajectory is a given SSP's "own" realisation is reporting metadata
(`default_for_ssp`), not structure.

> **The technical ordering deliberately does not match the storytelling.** A run's
> `description` reads "SSP3 under …", because that is how results get discussed; its
> *position* in the tree is chosen to avoid replicating rasters. Read lineage as
> "which climate, then which society", and prose as "which society, under which climate".

The saving is only real because the trajectories are shared. Realising 5 SSPs against 4
trajectories × 3 quantiles costs **12** stored climate copies with this ordering versus
**60** with SSP on top. If instead every SSP had its own bespoke GWL schedule, the two
orderings would cost the same — the replication comes from the schedule being SSP-specific,
not from the tree direction.

`id_run` layout — climate runs are `100` (current climatology) and
`1000 + 100 × traj_idx + q_idx` for GWL; an SSP leaf is `climate_id_run × 10 + ssp_idx`. So
`1302` is "third trajectory (`rise30`), q50" and `13023` is "SSP3 under it". Downstream steps should nonetheless **look runs up by the
`ssp` / `climate_framing` / `gwl_trajectory` / `gwl_quantile` columns** rather than
hard-coding integers. `runs_t` only mandates `id_run`/`parent_id_run`/`description`; extra
columns are carried through, which is what makes that lookup possible.

> **Scenario scope.** SSP0/1/3/4/5; SSP2 is deliberately excluded (see `README.md`). The
> per-SSP demand targets live in `NCCS-SSP-scenarios/Tools/NCCS_simulation_LULC_areas.xlsx`,
> keyed by these same labels. The `EI_Nature`/`EI_Culture`/`EI_Society`/`Growth and Extinction`
> names in `Transition_Tables.xlsx` are an **earlier vintage** and do *not* map cleanly onto
> SSP0/1/3/4/5 — matching the two on their 2060 class areas leaves residuals of 3–7 × 10⁵ ha
> and assigns two SSPs to the same name. Don't build a crosswalk between them.

### GWL trajectories

> **PROVISIONAL — the open question in `TODO.md` ("SSP→GWL / CO₂ mapping").** CH2025 publishes
> exactly three levels (GWL1.5 / GWL2.0 / GWL3.0) as 30-year climatologies. The four schedules
> below are a defensible reading of IPCC AR6 WG1 SPM.8a, not a settled decision, and are
> isolated in `02-ingest-preds-ch2025-3-gwl.qmd` (which consumes them) so they change in one
place. `fast30` is **capped** — AR6 puts SSP5-8.5 above
> 3 °C well before the 2060s, but CH2025 publishes nothing higher.

```{r}
#| label: setup-db-runs-spec
library(data.table)

# Trajectory order fixes the id_run digit; default_for_ssp is reporting metadata only.
trajectory_specs <- data.table::rowwiseDT(
  gwl_trajectory =, traj_idx =, default_for_ssp =, traj_description        =,
  "stable15",       1L,         "SSP0",           "Held at 1.5 C throughout",
  "stab20",         2L,         "SSP1",           "1.5 C, stabilising at 2.0 C",
  "rise30",         3L,         "SSP3,SSP4",      "1.5 C rising to 3.0 C",
  "fast30",         4L,         "SSP5",           "2.0 C reaching 3.0 C by mid-century (capped)"
)

ssp_specs <- data.table::rowwiseDT(
  ssp   =, ssp_idx =, ssp_description                                                       =,
  "SSP0", 1L,         "Positive normative visioning; newly added, no original counterpart",
  "SSP1", 2L,         "Sustainability",
  "SSP3", 3L,         "Regional rivalry",
  "SSP4", 4L,         "Inequality",
  "SSP5", 5L,         "Fossil-fuelled development"
)

# CH2025 uncertainty arms to register. q50 alone by default; add "q5"/"q95" for the
# uncertainty sensitivity -- each one multiplies stored climate volume.
gwl_quantiles <- data.table::data.table(gwl_quantile = "q50", q_idx = 2L)
```

The `current` framing stores **no climate at all**: its SSP children inherit the observed
climatology at `id_period 0`, which is exactly what "only society changes" means. It is
therefore free, and is always registered.

```{r}
#| label: setup-db-runs
climate_runs <- rbind(
  data.table::data.table(
    id_run = 100L,
    parent_id_run = 0L,
    description = "Current climatology (no projection; inherits the id_period 0 baseline)",
    climate_framing = "current",
    gwl_trajectory = NA_character_,
    gwl_quantile = NA_character_
  ),
  trajectory_specs[
    ,
    .(gwl_quantile = gwl_quantiles$gwl_quantile, q_idx = gwl_quantiles$q_idx),
    by = .(gwl_trajectory, traj_idx, traj_description)
  ][, .(
    id_run = 1000L + 100L * traj_idx + q_idx,
    parent_id_run = 0L,
    description = paste0(
      "CH2025 GWL trajectory '", gwl_trajectory, "' (", traj_description, "), ", gwl_quantile
    ),
    climate_framing = "gwl",
    gwl_trajectory,
    gwl_quantile
  )]
)

# SSP leaves: the full cross of SSP x climate run. Leaves are cheap -- only the pairings
# actually allocated cost anything downstream.
ssp_runs <-
  climate_runs[
    ,
    .(ssp_idx = ssp_specs$ssp_idx),
    by = .(climate_id_run = id_run)
  ][
    climate_runs,
    on = c(climate_id_run = "id_run"),
    nomatch = NULL
  ][
    ssp_specs,
    on = "ssp_idx",
    nomatch = NULL
  ][, .(
    id_run = climate_id_run * 10L + ssp_idx,
    parent_id_run = climate_id_run,
    description = paste0(ssp, " (", ssp_description, ") under ", description),
    ssp,
    climate_framing,
    gwl_trajectory,
    gwl_quantile
  )]

db$runs_t <- as_runs_t(rbind(
  db$runs_t, # the base run 0 created on init
  climate_runs,
  ssp_runs,
  fill = TRUE
))

stopifnot(
  "duplicate id_run" = !anyDuplicated(db$runs_t$id_run),
  "unresolvable parent_id_run" = all(
    is.na(db$runs_t$parent_id_run) | db$runs_t$parent_id_run %in% db$runs_t$id_run
  )
)

db$runs_t
```
