---
title: "00 — Setup DB"
---

Sets up the SSP-CH experiment database: an `evoland_db` (DuckDB) backing store aimed at
Arealstatistik land-use plus the Swiss predictor stack. This step defines the two axes
that every later step keys against — the **coordinate grid** (`id_coord`) and the
**periods** (`id_period`) — so it must run first.

```{r}
#| label: setup-db-create
library(evoland)

db <- evoland_db$new(
  path = "ssp-ch.evolanddb",
  report_name = "ssp-ch",
  report_name_pretty = "Replication of SSP-CH in new evoland"
)
```

## Coordinate grid

A square 100 m grid in **CH1903+ / LV95 (EPSG:2056)** over the full Swiss bounding box.
The commented-out inner extent is a small central-Switzerland window kept for quick local
test runs; the active extent is the whole country.

```{r}
#| label: setup-db-coords
db$coords_t <- create_coords_t_square(
  epsg = 2056,
  extent = terra::ext(c(
    # somewhere in the middle of the country
    # xmin = 2697000,
    # xmax = 2698000,
    # ymin = 1252000,
    # ymax = 1253000

    # full extent
    xmin = 2480000,
    xmax = 2840000,
    ymin = 1070000,
    ymax = 1300000
  )),
  resolution = 100
)
```

## Periods

Decadal steps (`P10Y`): observed 1985–2020 (the Arealstatistik survey years) followed by
extrapolated steps out to 2060.

```{r}
#| label: setup-db-periods
db$periods_t <- create_periods_t(
  period_length_str = "P10Y",
  start_observed = "1985-01-01",
  end_observed = "2020-01-01",
  end_extrapolated = "2060-01-01"
)
```
