---
title: "02 — Predictors: DEM (DHM25)"
---

Ingest a digital elevation model and derive 100 m terrain predictors (elevation, slope,
aspect) for the model grid.

**Source:** swisstopo **DHM25** (`ch.swisstopo.digitales-hoehenmodell_25`), ~25 m, LV03
(EPSG:21781), vertical datum LN02, survey vintage ~1980s.
<https://www.swisstopo.admin.ch/en/height-model-dhm25>

DHM25 is **not** exposed through the swisstopo STAC API, so we download the whole-Swiss
ASCII-grid ZIP directly from the geo.admin.ch OGD topology server.

> **Optional expansion.** For hectare-scale terrain roughness, maximum slope, or other
> fine-scale metrics, swissALTI3D (STAC `ch.swisstopo.swissalti3d`, 2 m COGs, ~44k tiles,
> ~44 GB) would be the appropriate source — deferred here due to data volume.

```{r}
#| label: ingest-dem-download
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal
extent_wide <- db$extent |> terra::extend(1000)

sources_dhm25 <-
  list(
    list(
      url = "https://cms.geo.admin.ch/ogd/topography/DHM25_MM_ASCII_GRID.zip",
      md5sum = "511ffe2512f1445e2f179ec74136ebbd"
    )
  ) |>
  data.table::rbindlist() |>
  download_and_verify()

zip_path <- sources_dhm25[["local_path"]][[1L]]
```

## Mosaic, reproject and resample

The ZIP holds several ASCII grid tiles. They are in LV03 and may lack embedded CRS
metadata in the `.asc` files, so we assign EPSG:21781 explicitly after reading. DHM25
covers only Switzerland (~70 M cells at 25 m, ~280 MB float32), so the full in-memory
mosaic is feasible before reprojecting. Elevation is continuous, so **bilinear**
interpolation is used to reproject to EPSG:2056 at 100 m (matching every other predictor).

```{r}
#| label: ingest-dem-reproject
asc_file <-
  unzip(zip_path, list = TRUE)[["Name"]] |>
  purrr::keep(stringr::str_detect, "\\.asc$")

dem_25m <- terra::rast(paste0("/vsizip/", zip_path, "/", asc_file))
terra::crs(dem_25m) <- "EPSG:21781"

dem_100m <-
  dem_25m |>
  terra::project(y = "EPSG:2056", method = "bilinear", res = 100) |>
  terra::crop(extent_wide)
```

## Terrain derivatives

Slope and aspect are computed on the 100 m DEM.

> **Open — hillshade / insolation.** The original predictor set included a hillshade
> layer, probably intended as an insolation proxy. That signal is arguably already
> carried by slope + aspect; a proper treatment would be ray-traced insolation rather
> than a single-azimuth hillshade. Hillshade is therefore **omitted** here. Tracked in
> `TODO.md` ("DEM hillshade semantics").

```{r}
#| label: ingest-dem-terrain
slope_100m <- terra::terrain(dem_100m, v = "slope", unit = "deg")
aspect_100m <- terra::terrain(dem_100m, v = "aspect", unit = "deg")
```

## Register predictors

Each raster is sampled at the model coordinates and written to `pred_data_t` at
`id_period = 0` (static terrain). Full provenance is carried in each predictor's
`description`.

```{r}
#| label: ingest-dem-register
pred_specs <- list(
  elevation_mean_100m = list(
    rast = dem_100m,
    unit = "masl",
    pretty_name = "Elevation (metres above sea level)",
    description = paste0(
      "Elevation at 100m resolution, bilinear resampled from DHM25 (~25m, LV03/LN02, ",
      "survey vintage ~1980s). swisstopo ch.swisstopo.digitales-hoehenmodell_25. ",
      "https://www.swisstopo.admin.ch/en/height-model-dhm25"
    )
  ),
  slope_mean_100m = list(
    rast = slope_100m,
    unit = "degrees",
    pretty_name = "Slope (degrees)",
    description = paste0(
      "Slope in degrees computed with terra::terrain(v='slope', unit='degrees') on ",
      "the 100m DHM25-derived DEM (ch.swisstopo.digitales-hoehenmodell_25)."
    )
  ),
  aspect_mean_100m = list(
    rast = aspect_100m,
    unit = "degrees",
    pretty_name = "Aspect (degrees from North, clockwise)",
    description = paste0(
      "Aspect in degrees computed with terra::terrain(v='aspect', unit='degrees') on ",
      "the 100m DHM25-derived DEM. 0 = North, 90 = East."
    )
  )
)

for (pred_name in names(pred_specs)) {
  spec <- pred_specs[[pred_name]]
  r <- spec[["rast"]]
  terra::set.names(r, pred_name)

  pred_data <- extract_using_coords_t(r, coords_minimal)

  db$add_predictor(
    pred_data_raw = pred_data[, .(id_coord, id_period = 0L, value)],
    name = pred_name,
    fill_value = NA,
    unit = spec[["unit"]],
    pretty_name = spec[["pretty_name"]],
    orig_format = "DHM25 ASCII grid tiles (~25m, LV03/EPSG:21781), bilinear resampled to EPSG:2056 100m",
    description = spec[["description"]],
    sources = sources_dhm25[, .(url, md5sum)]
  )
}
```
