---
title: "02 — CH2025 climate: ETL"
---

Read the `-obs` CH2025 netCDFs cached by `02-ingest-preds-ch2025-1-download.qmd` and ingest
each as a predictor. All dimensions are recovered from the cached **file names** (see
`nc_filename()` in the download step), so nothing is passed between the two steps in memory.
Each file is read, transformed and written **one at a time** to keep memory flat — a single
long table over ~4 M coords × 67 files would not fit comfortably.

## Data model (`pred_data_t` dimensions)

- **`name` = `{indicator}_{time_of_year}`** (e.g. `tas_yearly`, `pr_DJF`). A predictor is
  one climate variable and must stay consistent across historic (observed) and future
  (extrapolated) periods, so the **reference period is *not* part of the name**.
- **`id_period`** encodes the temporal progression: the observed reference period here, and
  (deferred) the progression through global warming levels for `-gwl`.
- **`id_run`** encodes quantile × SSP scenario (relevant for the deferred `-gwl` runs).

**Scope now:** `-obs` files only, at `id_period = 0` (baseline) under the active
`id_run = 0`. Because the name drops the reference period, only *one* reference period can
occupy `id_period 0`; we use the current WMO normal **1991-2020**, which covers all 21
indicators (1961-1990 covers only 14). 1961-1990 stays cached for a later historic
`id_period` if wanted.

**Deferred — `-gwl` projections.** This step uses `db$add_predictor` (safe upsert of both
`pred_meta_t` and `pred_data_t`), right for the initial commit. When appending the `-gwl`
projections later, write directly to `pred_data_t` keyed by the `id_pred` that already
exists in `pred_meta_t` (one per `{indicator}_{time_of_year}`), at the mapped `id_period`
and the scenario/quantile `id_run`.

```{r}
#| label: ch2025-etl-params
library(data.table)
library(evoland)

base_url <- "https://service.meteoswiss.ch/pbbackend/api/v1/products"
cache_dir <- file.path(getOption("evoland.cachedir"), "ch2025")

# Reference period to populate the id_period = 0 baseline (see above).
obs_reference_period <- "1991-2020"
```

## Per-indicator metadata

One spec per indicator (pattern mirrors the swissTLM3D step), tied to files via the
indicator token parsed from each file name. `unit`/`pretty_name` track the netCDF
`long_name` & `units`; `description` adds the CH2025 definition + provenance. Commenting a
spec out skips that indicator at ingest even if its netCDF is already on disk — the
**heating/cooling degree-day** indicators (`HDD`, `HED`, `CoDD`, `COD`) are currently
skipped this way.

> **TODO:** this set is missing bioclimatic indicators — see
> <https://www.chelsa-climate.org/datasets/chelsa_bioclim> and `2-ingest-preds-ch2025-todo.md`.

```{r}
#| label: ch2025-etl-specs
prov <- "MeteoSwiss CH2025 observed climatology (1 km, EPSG:2056)."
indicator_specs <- list(
  ID = list(
    unit = "days",
    pretty_name = "Ice days",
    description = paste("Annual number of ice days (daily maximum temperature below 0 C).", prov)
  ),
  FD = list(
    unit = "days",
    pretty_name = "Frost days",
    description = paste("Annual number of frost days (daily minimum temperature below 0 C).", prov)
  ),
  # HDD = list(
  #   unit = "Kd",
  #   pretty_name = "Heating degree days",
  #   description = paste(
  #     "Heating degree days: annual sum of the daily shortfall of mean temperature below the heating target, i.e. difference on days below 12°C mean and 20°C heating target.",
  #     prov
  #   )
  # ),
  # HED = list(
  #   unit = "days",
  #   pretty_name = "Heating days",
  #   description = paste("Annual number of heating days, i.e. mean temp. ≤ 12°C", prov)
  # ),
  HD = list(
    unit = "days",
    pretty_name = "Heat days",
    description = paste("Annual number of heat days, i.e. max. temp. >= 30°C", prov)
  ),
  HW2 = list(
    unit = "warning days",
    pretty_name = "Heat warning days (Level 2)",
    description = paste(
      "Annual number of heat-warning days at MeteoSwiss warning level 2, mean temp >= 25",
      prov
    )
  ),
  HW3 = list(
    unit = "warning days",
    pretty_name = "Heat warning days (Level 3)",
    description = paste(
      "Annual number of heat-warning days at MeteoSwiss warning level 3, mean temp >= 25° for at least 3 days",
      prov
    )
  ),
  HW4 = list(
    unit = "warning days",
    pretty_name = "Heat warning days (Level 4)",
    description = paste(
      "Annual number of heat-warning days at MeteoSwiss warning level 4, mean temp >= 27°C for at least 3 days",
      prov
    )
  ),
  # CoDD = list(
  #   unit = "Kd",
  #   pretty_name = "Cooling degree days",
  #   description = paste(
  #     "Cooling degree days: annual sum of the daily excess of mean temperature above the cooling target (18.3°C).",
  #     prov
  #   )
  # ),
  # COD = list(
  #   unit = "days",
  #   pretty_name = "Cooling days",
  #   description = paste("Annual number of cooling days (daily mean >= 18.3).", prov)
  # ),
  VHD = list(
    unit = "days",
    pretty_name = "Very hot days",
    description = paste(
      "Annual number of very hot days (very high daily maximum temperature, >= 35).",
      prov
    )
  ),
  SD = list(
    unit = "days",
    pretty_name = "Summer days",
    description = paste(
      "Annual number of summer days (daily maximum temperature at or above the summer-day threshold >= 25).",
      prov
    )
  ),
  TN = list(
    unit = "days",
    pretty_name = "Tropical nights",
    description = paste(
      "Annual number of tropical nights (high daily minimum temperature >=20mm).",
      prov
    )
  ),
  pr = list(
    unit = "mm day-1",
    pretty_name = "Precipitation",
    description = paste("Mean daily precipitation amount.", prov)
  ),
  PR20 = list(
    unit = "days",
    pretty_name = "Moderate precipitation days",
    description = paste("Annual number of days with moderate precipitation. >= 20mm", prov)
  ),
  PR40 = list(
    unit = "days",
    pretty_name = "Heavy precipitation days",
    description = paste("Annual number of days with heavy precipitation. >= 40mm", prov)
  ),
  PR60 = list(
    unit = "days",
    pretty_name = "Very heavy precipitation days",
    description = paste("Annual number of days with very heavy precipitation. >= 60mm", prov)
  ),
  SNFD = list(
    unit = "days",
    pretty_name = "Snowfall days",
    description = paste("Annual number of days with snowfall, <2deg mean & >1mm/d precip", prov)
  ),
  tas = list(
    unit = "degrees_C",
    pretty_name = "Near-surface air temperature",
    description = paste("Mean near-surface air temperature.", prov)
  ),
  tasmax = list(
    unit = "degrees_C",
    pretty_name = "Daily maximum near-surface air temperature",
    description = paste("Daily maximum near-surface air temperature.", prov)
  ),
  tasmin = list(
    unit = "degrees_C",
    pretty_name = "Daily minimum near-surface air temperature",
    description = paste("Daily minimum near-surface air temperature.", prov)
  ),
  CDD = list(
    unit = "days",
    pretty_name = "Consecutive dry days",
    description = paste("Maximum run of consecutive dry days, <1mm/d", prov)
  )
)
```

## Source-URL reconstruction

Rebuild the source API URL from the parsed dimensions, mirroring the query in the download
step. There is no published md5sum (the endpoint is generated on demand).

```{r}
#| label: ch2025-etl-url
obs_url <- function(product, indicator, reference_period, time_of_year) {
  req <- httr2::request(paste0(base_url, "/", product, "/realizations")) |>
    httr2::req_url_query(
      language = "de",
      referencePeriod = reference_period,
      timeOfYear = time_of_year,
      productType = "data",
      productName = product,
      scenarioIndicator = indicator,
      mediaType = "application/netcdf"
    )
  req$url
}
```

## Recover the file inventory from cache

Obs file names are `{category}-obs_{indicator}_{reference_period}_{time_of_year}.nc`.
Indicator codes carry no underscore and the reference period uses a hyphen, so the stem
splits cleanly into exactly four underscore-separated tokens. We keep only the baseline
reference period.

```{r}
#| label: ch2025-etl-inventory
obs_files <- list.files(cache_dir, pattern = "-obs_.*\\.nc$")

inventory <- data.table(file = obs_files)
inventory[, stem := sub("\\.nc$", "", file)]
inventory[,
  c(
    "product_stem",
    "indicator",
    "reference_period",
    "time_of_year"
  ) := data.table::tstrsplit(
    stem,
    "_",
    fixed = TRUE
  )
]
inventory[, product := paste0("climate-ch2025-maps-", product_stem)]
inventory[, pred_name := paste(indicator, time_of_year, sep = "_")]

if (anyNA(inventory[, .(indicator, reference_period, time_of_year)])) {
  stop(
    "Unparseable obs file name(s):\n  ",
    paste(
      inventory[
        is.na(indicator) | is.na(reference_period) | is.na(time_of_year),
        file
      ],
      collapse = "\n  "
    )
  )
}

# id_period 0 baseline takes a single reference period
inventory <- inventory[reference_period == obs_reference_period]

message(
  "Found ",
  nrow(inventory),
  " obs files (",
  obs_reference_period,
  ") across ",
  data.table::uniqueN(inventory, by = "indicator"),
  " indicators on disk."
)
```

## Read → transform → write (one file at a time)

Each netCDF is cropped to the analysis extent and point-extracted at the 100 m model
coordinates (nearest 1 km cell, no bilinear interpolation), then upserted via
`db$add_predictor`. Indicators without a metadata spec (e.g. the skipped degree-day ones)
are warned and skipped.

```{r}
#| label: ch2025-etl-load
db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal
extent_wide <- db$extent |> terra::extend(1000)

orig_format <- paste0(
  "MeteoSwiss CH2025 netCDF (EPSG:2056, 1km); baseline observed ",
  obs_reference_period,
  " period, point-extracted at 100m model coords (nearest cell, no bilinear interp.)"
)

for (i in seq_len(nrow(inventory))) {
  row <- inventory[i]
  spec <- indicator_specs[[row[["indicator"]]]]
  if (is.null(spec)) {
    warning("No metadata spec for indicator '", row[["indicator"]], "'; skipping ", row[["file"]])
    next
  }

  message(sprintf("[%d/%d] %s", i, nrow(inventory), row[["pred_name"]]))

  r <- terra::rast(file.path(cache_dir, row[["file"]])) |> terra::crop(extent_wide)
  terra::set.names(r, row[["pred_name"]])
  pred_data <- extract_using_coords_t(r, coords_minimal)

  db$add_predictor(
    pred_data_raw = pred_data[, .(id_coord, id_period = 0L, value)],
    name = row[["pred_name"]],
    fill_value = NA,
    unit = spec[["unit"]],
    pretty_name = sprintf("%s (%s)", spec[["pretty_name"]], row[["time_of_year"]]),
    orig_format = orig_format,
    description = spec[["description"]],
    sources = data.table(
      url = obs_url(
        row[["product"]],
        row[["indicator"]],
        row[["reference_period"]],
        row[["time_of_year"]]
      ),
      md5sum = NA_character_
    )
  )
}

message("Done. Ingested ", nrow(inventory), " obs predictors at id_period = 0.")
```
