---
title: "02 — Predictors: biogeographic regions"
---

Ingest the BAFU **biogeographic regions of Switzerland** as a *categorical* predictor, giving
the transition models a coarse regional identity to condition on. The original SSP-CH work
fitted models per bioregion (`lulcc.listbybioregion.R`); here the region enters as a factor
predictor instead, so one model per transition can still express regional differences without
splitting the training data.

**Source:** BAFU `ch.bafu.biogeographische_regionen`, "N2020 Revision", EPSG:2056 shapefile.
<https://www.bafu.admin.ch/de/publication?id=fn-ZXl-NXhXbl> ·
[STAC](https://data.geo.admin.ch/browser/index.html#/collections/ch.bafu.biogeographische_regionen)

The layer holds **12 polygons**: 6 regions (`R1`–`R6`) subdivided into 12 subregions
(`U11`–`U62`). Both are ingested — they are strictly nested, hence collinear by construction,
so `05-covariate-selection.qmd` should retain at most one of them per transition. The
covariance/correlation filter is the right place to make that call, not this step.

| Region | Subregions |
| --- | --- |
| R1 Jura | U11 Jura und Randen |
| R2 Mittelland | U21 Genferseegebiet · U22 Hochrheingebiet · U23 Westliches Mittelland · U24 Östliches Mittelland |
| R3 Alpennordflanke | U31 Voralpen · U32 Nordalpen |
| R4 Westliche Zentralalpen | U41 Westliche Zentralalpen |
| R5 Östliche Zentralalpen | U51 Östliche Zentralalpen · U52 Engadin |
| R6 Alpensüdflanke | U61 Südalpen · U62 Südliches Tessin |

> **Categorical encoding.** `pred_meta_t` supports `data_type = "factor"` with
> `factor_levels`; values are stored as **1-based indices into `factor_levels`** and
> reconstructed as R factors by `set_pred_coltypes()`. So the raw `RegionNumm` / `Unterregio`
> codes must be *recoded* to level indices — `Unterregio` in particular runs 11, 21, 22, …,
> 62, which is not a 1-based sequence. Storing the raw codes as `int` instead would make a
> learner treat "Engadin" as numerically greater than "Jura", which is meaningless.

## Download

The published checksum is a SHA-256 multihash (STAC `checksum:multihash`, `1220…` prefix),
whereas `download_and_verify()` takes an md5sum — so the md5 below was computed from the
archive after confirming its SHA-256 against the STAC entry.

```{r}
#| label: ingest-bioregions-download
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal

sources_bioreg <-
  data.frame(
    url = paste0(
      "https://data.geo.admin.ch/ch.bafu.biogeographische_regionen/",
      "biogeographische_regionen/biogeographische_regionen_2056.shp.zip"
    ),
    # sha256 f3c0567d3ed460295981ba73028b1e283b7d9ab3e4b6c070c9fde38aa2acaaf0
    # matches the STAC multihash 1220f3c0567d...; md5 computed from that archive
    md5sum = "f1cb4ae01c5c0af1ac263f4060b5081d"
  ) |>
  download_and_verify(target_dir = getOption("evoland.cachedir"))
```

## Read the layer

The shapefile is read straight out of the ZIP via GDAL's `/vsizip/` — the archive is ~4 MB
and holds a single 12-feature polygon layer, so the unpack-and-cache dance that
`02-ingest-preds-swisstlm3d.qmd` needs for swissTLM3D is unnecessary here. The `.dbf` is
UTF-8 encoded (per its `.CPG`), which matters for the German region names.

```{r}
#| label: ingest-bioregions-read
bioreg <- terra::vect(
  paste0("/vsizip/", sources_bioreg[["local_path"]], "/N2020_Revision_BiogeoRegion.shp"),
  opts = "ENCODING=UTF-8"
)

stopifnot(
  "expected 12 bioregion polygons" = nrow(bioreg) == 12L,
  "expected 6 distinct regions" = data.table::uniqueN(bioreg$RegionNumm) == 6L
)

# Level order is fixed by the numeric code so it is stable across re-runs.
region_levels <-
  bioreg |>
  data.table::as.data.table() |>
  unique(by = "RegionName") |>
  (\(dt) dt[order(RegionNumm)])() |>
  (\(dt) dt[, .(code = RegionName, label = DERegionNa)])()

subregion_levels <-
  bioreg |>
  data.table::as.data.table() |>
  unique(by = "Unterreg_1") |>
  (\(dt) dt[order(Unterregio)])() |>
  (\(dt) dt[, .(code = Unterreg_1, label = DEBioBedeu)])()

region_levels
subregion_levels
```

## Declare the predictors

```{r}
#| label: ingest-bioregions-meta
db$pred_meta_t <- create_pred_meta_t(list(
  bioregion = list(
    pretty_name = "Biogeographic region",
    description = paste(
      "BAFU biogeographic region of Switzerland (N2020 Revision), 6 classes:",
      paste0(region_levels$code, " ", region_levels$label, collapse = "; "),
      "- Categorical regional identity; the original SSP-CH implementation instead fitted",
      "separate models per bioregion."
    ),
    data_type = "factor",
    factor_levels = region_levels$code,
    orig_format = "EPSG:2056 polygon shapefile, point-in-polygon at 100m model coords",
    unit = NA_character_
  ),
  biosubregion = list(
    pretty_name = "Biogeographic subregion",
    description = paste(
      "BAFU biogeographic subregion (N2020 Revision), 12 classes:",
      paste0(subregion_levels$code, " ", subregion_levels$label, collapse = "; "),
      "- Strictly nested within `bioregion`, hence collinear with it; feature selection",
      "should retain at most one of the two per transition."
    ),
    data_type = "factor",
    factor_levels = subregion_levels$code,
    orig_format = "EPSG:2056 polygon shapefile, point-in-polygon at 100m model coords",
    unit = NA_character_
  )
))
```

## Extract at model coordinates

Point-in-polygon at each hectare-cell centre. `extract_using_coords_t()` melts to long form
(`attribute`, `value`) and drops non-matching coordinates by default, so cells outside every
polygon — the `coords_t` grid is a rectangle over the country, so its corners fall outside
Switzerland — simply do not appear. That is the right behaviour here: there is no defensible
region to assign them. The codes are recoded to 1-based factor indices, as noted above.

```{r}
#| label: ingest-bioregions-extract
template_rast <- db$lulc_data_as_rast(1)

bioreg_rasterized <- c(
  terra::rasterize(bioreg, template_rast, "RegionName", touches = TRUE),
  terra::rasterize(bioreg, template_rast, "Unterreg_1", touches = TRUE)
)

extracted <-
  extract_using_coords_t(bioreg_rasterized, coords_minimal) |>
  data.table::dcast(id_coord ~ layer, value.var = "value")

# we get some NAs; calculate distance to all polygons and take nearest one (cheaper than
# terra::distance)
reg_sample <-
  extracted[
    coords_minimal,
    on = "id_coord"
  ][
    is.na(RegionName) | is.na(Unterreg_1)
  ] |>
  terra::vect(crs = paste0("epsg:", attr(coords_minimal, "epsg"))) |>
  terra::distance(x = bioreg, y = _, names = c("Unterregio", "id_coord")) |>
  data.table::as.data.table(keep.rownames = TRUE) |>
  data.table::melt(id.vars = "rn", variable.name = "id_coord")

bioreg_map <- data.table::as.data.table(bioreg)[
  ,
  .(rn = Unterregio, Unterreg_1, RegionName)
]
fixed_NAs <-
  reg_sample[
    , .(
      rn = as.numeric(rn),
      id_coord = as.integer(as.character(id_coord)),
      value
    )
  ][,
    .SD[order(value)][1],
    by = id_coord
  ][
    bioreg_map,
    on = "rn", nomatch = NULL
  ][
    ,
    .(id_coord, RegionName, Unterreg_1)
  ]

extracted <- rbind(extracted, fixed_NAs)

# recode to the 1-based level index that pred_data_t expects for factor predictors
extracted[, `:=`(
  bioregion = match(RegionName, region_levels$code),
  biosubregion = match(Unterreg_1, subregion_levels$code)
)]

stopifnot(
  "some region codes did not match the declared factor levels" =
    extracted[is.na(bioregion) | is.na(biosubregion), .N] == 0L
)

message(
  "Assigned a region to ", nrow(extracted), " of ", nrow(coords_minimal),
  " coordinates (", nrow(coords_minimal) - nrow(extracted), " fall outside Switzerland)."
)
```

## Commit

Both predictors are static, so they go to `id_period = 0` under the base run — the
static-fallback slot every period reads through.

```{r}
#| label: ingest-bioregions-commit
pred_ids <- db$pred_meta_t[name %in% c("bioregion", "biosubregion"), .(id_pred, name)]

bioreg_long <-
  data.table::melt(
    extracted[, .(id_coord, bioregion, biosubregion)],
    id.vars = "id_coord",
    variable.name = "name",
    value.name = "value",
    variable.factor = FALSE
  )
bioreg_long <- merge(bioreg_long, pred_ids, by = "name")

db$pred_data_t <- as_pred_data_t(
  bioreg_long[, .(id_run = 0L, id_period = 0L, id_pred, id_coord, value)]
)

db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key = , value = ,
  "bioregion_url", sources_bioreg[["url"]],
  "bioregion_md5sum", sources_bioreg[["md5sum"]],
  "bioregion_provider", "BAFU ch.bafu.biogeographische_regionen (N2020 Revision)"
))
```
