---
title: "02 — CH2025 climate: projected (GWL) predictors"
---

Ingest the `-gwl` CH2025 projections cached by `02-ingest-preds-ch2025-1-download.qmd` as
period- and run-specific overrides of the observed climatology committed by
`02-ingest-preds-ch2025-2-etl.qmd`.

The observed climatology sits at `id_period 0` on run `0`, which evoland treats as the
**static fallback**: `pred_data_wide.sql` reads `id_period in (0, <target period>)` and, since
the fix in evoland-plus, prefers the period-specific row. So writing a GWL value at
`id_period 5..8` under a *climate* run overrides the baseline for that run and every SSP leaf
below it, while any predictor left unwritten keeps falling back to the baseline. Nothing needs
deleting.

Note the precedence is decided **per slice**, not per coordinate: if a predictor has any data
at the target period, that whole `(id_pred, id_period)` slice is used and its period-0 rows are
ignored. So a projection must be written for *every* coordinate it should apply to — a
partially-populated slice yields `NA` (then `pred_meta_t$fill_value`) at the missing cells
rather than quietly falling back to the baseline there.

> **Ordering — this step is not a plain stage-02 step.** It is numbered `02-…-3-` because it
> belongs to the CH2025 family, but it should be run **after `05-covariate-selection.qmd`**,
> and re-run whenever the selected predictor set changes. See "Why this cannot ingest
> everything" below; ingesting the full indicator set for every scenario arm is not
> tractable, so the step defaults to projecting only the climate predictors that the
> selected models actually use.

## Why this cannot ingest everything

Projected climate is the only **per-run, per-period** predictor family, so its storage cost is
the product of every axis it varies along:

| Axis | Size |
| --- | --- |
| coordinates with LULC data | ~4.1 × 10⁶ |
| extrapolated periods (5–8) | 4 |
| GWL trajectories × quantiles | 4 × 1 (q50 default) |
| yearly climate indicators ingested | 18 |

Because `runs_t` puts **climate above SSP** (see `00-setup-db.qmd`), climate is stored once per
*climate* run rather than once per SSP arm — 4 copies at the q50 default instead of 20, and
12 instead of 60 if all three quantiles are registered. The SSP leaves inherit it.

Even so, the full indicator set is ~1.2 × 10⁹ rows at the q50 default. Two further levers
reduce it, and this step exposes both:

1. **Project only selected predictors.** After `05`, most climate indicators are pruned for
   most transitions. `pred_names` defaults to the climate predictors still present in
   `trans_preds_t`.
2. **Project only the quantiles you need.** `q50` is the default; `q5`/`q95` are an
   uncertainty sensitivity, and each one multiplies the volume.

The `current` framing costs nothing at all: it stores no climate and its SSP children inherit
the `id_period 0` baseline, which is exactly what "only society changes" means.

## Only `yearly` projections exist

The download step probes `-gwl` products for `time_of_year = "yearly"` only — MeteoSwiss
publishes the GWL aggregates annually, while the `-obs` family additionally carries `DJF`,
`MAM`, `JJA`, `SON`. Consequently the **seasonal predictors (`*_DJF`, `*_MAM`, `*_JJA`,
`*_SON`) have no projected counterpart**. Under a GWL run they silently keep their observed
baseline, i.e. they are held constant into the future.

That is a modelling decision, not a detail: if a seasonal predictor survives `05` and enters
a transition model, that model's projections carry a frozen seasonal climate while its annual
terms move. Either restrict the climate predictors offered to `05` to `*_yearly`, or accept
and document the freeze. This step warns about any seasonal climate predictor it finds in the
selected set.

## GWL trajectories

The period → GWL schedule is a property of the **climate trajectory**, not of an SSP: the
trajectories are shared, and several SSPs are realised against each. `00-setup-db.qmd`
registers one climate run per (trajectory × quantile) and names the trajectories; the schedule
itself lives here, where it is consumed.

> **PROVISIONAL — the open question in `TODO.md` ("SSP→GWL / CO₂ mapping").** CH2025 publishes
> exactly three levels (GWL1.5 / GWL2.0 / GWL3.0) as 30-year climatologies. These four
> schedules are a defensible reading of IPCC AR6 WG1 SPM.8a, not a settled decision.
> `fast30` is **capped**: AR6 puts SSP5-8.5 above 3 °C well before the 2060s, but CH2025
> publishes nothing higher. That understates late-century SSP5 warming and must be stated in
> reporting. Which trajectory is a given SSP's "own" realisation is recorded as
> `default_for_ssp` in `00-setup-db.qmd` and is reporting metadata only.

```{r}
#| label: ch2025-gwl-trajectories
library(data.table)
library(evoland)

# id_period 5..8 == decades 2025-34, 2035-44, 2045-54, 2055-64
gwl_trajectories <- data.table::rowwiseDT(
  gwl_trajectory =, id_period =, gwl      =,
  "stab15",         5L,          "GWL1.5",
  "stab15",         6L,          "GWL1.5",
  "stab15",         7L,          "GWL1.5",
  "stab15",         8L,          "GWL1.5",

  "stab20",         5L,          "GWL1.5",
  "stab20",         6L,          "GWL1.5",
  "stab20",         7L,          "GWL2.0",
  "stab20",         8L,          "GWL2.0",

  "rise30",         5L,          "GWL1.5",
  "rise30",         6L,          "GWL2.0",
  "rise30",         7L,          "GWL2.0",
  "rise30",         8L,          "GWL3.0",

  "fast30",         5L,          "GWL2.0",
  "fast30",         6L,          "GWL2.0",
  "fast30",         7L,          "GWL3.0",
  "fast30",         8L,          "GWL3.0"
)
```

## Parameters

```{r}
#| label: ch2025-gwl-params
cache_dir <- file.path(getOption("evoland.cachedir"), "ch2025")
base_url <- "https://service.meteoswiss.ch/pbbackend/api/v1/products"

# Which uncertainty arms to materialise. Add "q5"/"q95" for the uncertainty sensitivity.
quantiles_to_ingest <- "q50"

# NULL = derive the predictor set from trans_preds_t (i.e. what 05 retained).
# Set explicitly to override, e.g. c("tas_yearly", "pr_yearly").
pred_names_override <- NULL

db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal
extent_wide <- db$extent |> terra::extend(1000)
```

## Resolve which predictors to project

```{r}
#| label: ch2025-gwl-select-preds
pred_meta <- db$pred_meta_t

# climate predictors are exactly those named {indicator}_{time_of_year} by the -2- ETL step
climate_pred_names <- pred_meta[
  grepl("_(yearly|DJF|MAM|JJA|SON)$", name),
  name
]

pred_names <-
  if (!is.null(pred_names_override)) {
    pred_names_override
  } else {
    selected <- db$trans_preds_t
    if (nrow(selected) == 0L) {
      stop(
        "trans_preds_t is empty - run 05-covariate-selection.qmd first, or set ",
        "pred_names_override explicitly. See 'Why this cannot ingest everything' above."
      )
    }
    intersect(pred_meta[id_pred %in% unique(selected$id_pred), name], climate_pred_names)
  }

seasonal <- grep("_(DJF|MAM|JJA|SON)$", pred_names, value = TRUE)
if (length(seasonal) > 0L) {
  warning(
    "No -gwl projections exist for seasonal predictors; these will stay at their observed ",
    "baseline under every GWL run: ",
    paste(seasonal, collapse = ", ")
  )
}

pred_names <- grep("_yearly$", pred_names, value = TRUE)
stopifnot("No yearly climate predictors to project" = length(pred_names) > 0L)

message("Projecting ", length(pred_names), " yearly climate predictors: ",
        paste(pred_names, collapse = ", "))
```

## Build the work plan

Each cached `-gwl` file is `{category}-gwl_{indicator}_{GWL}_{quantile}_yearly.nc`. A file is
read **once** per `(indicator, gwl, quantile)` and fanned out to every `(id_run, id_period)`
that the crosswalk maps onto that level — reading 4.1 M points is far more expensive than
replicating the resulting column.

```{r}
#| label: ch2025-gwl-inventory
gwl_files <- list.files(cache_dir, pattern = "-gwl_.*\\.nc$")

inventory <- data.table(file = gwl_files)
inventory[, stem := sub("\\.nc$", "", file)]
inventory[,
  c("product_stem", "indicator", "gwl", "uncertainty", "time_of_year") :=
    data.table::tstrsplit(stem, "_", fixed = TRUE)
]
inventory[, product := paste0("climate-ch2025-maps-", product_stem)]
inventory[, pred_name := paste(indicator, time_of_year, sep = "_")]

if (anyNA(inventory[, .(indicator, gwl, uncertainty, time_of_year)])) {
  stop(
    "Unparseable gwl file name(s):\n  ",
    paste(inventory[is.na(indicator) | is.na(gwl) | is.na(uncertainty), file], collapse = "\n  ")
  )
}

inventory <- inventory[pred_name %in% pred_names & uncertainty %in% quantiles_to_ingest]

# climate runs: the gwl-framing arms registered in 00-setup-db.qmd. These sit ABOVE the
# SSP leaves, so writing here serves every SSP realised against the trajectory.
gwl_runs <- db$runs_t[climate_framing == "gwl" & gwl_quantile %in% quantiles_to_ingest]
stopifnot("No gwl-framing runs in runs_t - run 00-setup-db.qmd" = nrow(gwl_runs) > 0L)

# (climate id_run, id_period) -> the (gwl, quantile) file that supplies it
targets <-
  gwl_runs[
    gwl_trajectories,
    on = "gwl_trajectory",
    allow.cartesian = TRUE,
    nomatch = NULL
  ][, .(id_run, id_period, gwl, uncertainty = gwl_quantile)]

work <- inventory[targets, on = .(gwl, uncertainty), allow.cartesian = TRUE, nomatch = NULL]

# every (run, period, predictor) slice must resolve to exactly one source raster
slice_counts <- work[, .N, by = .(id_run, id_period, pred_name)]

stopifnot(
  "Some (run, period, predictor) slices resolve to more than one source raster" =
    all(slice_counts$N == 1L),
  "Some (run, period, predictor) slices have no source raster - check the crosswalk against the cached GWL levels" =
    nrow(slice_counts) ==
      data.table::uniqueN(targets, by = c("id_run", "id_period")) * length(pred_names)
)

message(
  "Plan: ", data.table::uniqueN(work, by = c("indicator", "gwl", "uncertainty")),
  " rasters to read, fanned out to ", nrow(unique(work[, .(id_run, id_period, pred_name)])),
  " (run, period, predictor) slices."
)
```

## Read → extract → write

One raster read per `(indicator, gwl, quantile)`; the extracted column is then written once
per target `(id_run, id_period)`. We write `pred_data_t` directly rather than via
`db$add_predictor()` — the predictors already exist in `pred_meta_t` from the `-2-` ETL step,
and we must not re-declare their metadata.

```{r}
#| label: ch2025-gwl-load
gwl_url <- function(product, indicator, gwl, uncertainty) {
  httr2::request(paste0(base_url, "/", product, "/realizations")) |>
    httr2::req_url_query(
      language = "de",
      globalWarmingLevel = gwl,
      uncertainty = uncertainty,
      timeOfYear = "yearly",
      productType = "data",
      productName = product,
      scenarioIndicator = indicator,
      mediaType = "application/netcdf"
    ) |>
    (\(req) req$url)()
}

pred_id_of <- setNames(pred_meta$id_pred, pred_meta$name)

sources <- unique(work[, .(file, product, indicator, gwl, uncertainty, pred_name)])

for (i in seq_len(nrow(sources))) {
  src <- sources[i]
  message(sprintf(
    "[%d/%d] %s @ %s %s", i, nrow(sources), src$pred_name, src$gwl, src$uncertainty
  ))

  r <- terra::rast(file.path(cache_dir, src$file)) |> terra::crop(extent_wide)
  terra::set.names(r, src$pred_name)
  extracted <- extract_using_coords_t(r, coords_minimal)[, .(id_coord, value)]

  # every (run, period) this raster supplies; one commit per raster, not per slice
  slices <- unique(work[file == src$file, .(id_run, id_period)])
  id_pred_i <- unname(pred_id_of[[src$pred_name]])

  db$pred_data_t <- as_pred_data_t(
    slices[,
      extracted[, .(id_coord, value)],
      by = .(id_run, id_period)
    ][, .(id_run, id_period, id_pred = id_pred_i, id_coord, value)]
  )

  rm(extracted)
  gc(verbose = FALSE)
}

message(
  "Done. Projected climate written for ", nrow(gwl_runs),
  " climate runs; SSP leaves inherit it via run lineage."
)
```

## Provenance

The `-gwl` source URLs are reconstructed the same way as in the `-obs` ETL; the endpoint is
generated on demand and publishes no md5sum. They are recorded against the run rather than
re-declared on the predictor, since the predictor metadata already carries its observed
provenance.

```{r}
#| label: ch2025-gwl-provenance
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key =                        , value =,
  "ch2025_gwl_quantiles",        paste(quantiles_to_ingest, collapse = ","),
  "ch2025_gwl_predictors",       paste(pred_names, collapse = ","),
  "ch2025_gwl_trajectories",     paste(
    gwl_trajectories[, paste0(gwl_trajectory, ":p", id_period, "=", gwl)],
    collapse = " "
  ),
  "ch2025_gwl_example_url",      gwl_url(
    sources$product[1], sources$indicator[1], sources$gwl[1], sources$uncertainty[1]
  )
))
```
