---
title: "02 — Predictors: STATENT employment under the SSPs"
---

Project the hectare-level FTE-by-sector predictors ingested by `02-ingest-preds-statent.qmd`
forward under each SSP, using the cantonal sectoral employment trajectories elicited for the
original SSP-CH project.

> **Ordering.** Like `02-ingest-preds-ch2025-3-gwl.qmd`, this writes per-run, per-period
> predictor data and therefore belongs *after* the SSP leaf runs exist (`00-setup-db.qmd`)
> and after `02-ingest-preds-statent.qmd` has committed the observed levels.

## Provenance: the input is elicited, and we do not know how

**Read this before using the outputs.** The scenario signal comes entirely from
`NCCS_future_FTE.csv`, a table of **canton × sector × year** full-time-equivalent totals per
SSP. Three things about it are worth stating plainly:

- **It is not derived in any code we have.** The original workflow
  (`NCCS-SSP-scenarios/Scripts/Preparation/Simulation_predictor_prep.R`) *reads* this file and
  does nothing scenario-specific of its own: it joins the base-year value from history,
  linearly interpolates to the model time steps with `stats::approx()`, rasterises per canton
  and differences consecutive periods. So the answer to "what makes SSP3's employment different
  from SSP1's" is not in the model — it is in this file.
- **Its provenance is unknown to us.** The original predictor table labels it
  `Data_citation = "Project Internal"`. We have not found a document describing the method,
  assumptions or expert panel behind the numbers, nor a version or date. It was recovered from
  the project's Zenodo data deposit ([doi:10.5281/zenodo.8263509](https://doi.org/10.5281/zenodo.8263509)).
- **It is therefore a borrowed assumption, not a reproducible derivation.** Everything this
  pipeline projects about employment inherits that opacity. This is a real gap against the
  experiment's stated goal of reproducible sources (see the top-level README) and should be
  disclosed wherever the SSP employment predictors influence a published result.

A companion `NCCS_future_population.csv` (canton × SSP × year resident population, in millions)
exists with the same provenance status. It is **not** ingested here — the README records that
municipal population was discarded as a predictor — but it is the natural input if a population
predictor is ever reinstated.

Structure of the FTE file: 5 scenarios × 3 sectors × 26 cantons = 390 rows; year columns
`2022, 2035, 2060, 2085, 2100`; cantons as two-letter abbreviations; **values in thousands of
FTE**. Note the base year is **2022**, not 2020.

## Method: scale the observed pattern by the cantonal trajectory

The trajectories are **cantonal totals**; our predictors are **hectare-level levels**. There is
no information in the input about *where within a canton* employment moves, so we do not invent
any: we hold the observed within-canton spatial pattern fixed and scale it by the canton's
sectoral growth ratio.

For canton $k$, sector $s$, target period $p$:

$$\text{FTE}_{\text{cell}, p} = \text{FTE}_{\text{cell}, p_{\text{obs}}} \times
  \frac{\text{FTE}^{\text{cant}}_{k,s,p}}{\text{FTE}^{\text{cant}}_{k,s,\text{base}}}$$

This means: a canton whose secondary sector halves by 2060 has every one of its hectares'
`fte_sec2` halved. Employment does not relocate within a canton, only expand or contract in
place. That is a real limitation and it is what the source data supports — the original does
the same thing, only coarser (it rasterises the cantonal value directly, so its predictor is
*constant* within a canton, discarding the hectare pattern entirely).

> **Departure from the original.** The original's predictor is `chg_FTE_*`, an annual *change*
> in FTE per labour-market region, at `Original_resolution = "Canton"`. This pipeline uses
> hectare-level *levels* (see the README provenance table). Scaling preserves that choice; it
> is not a reimplementation of the original predictor.

## Storage cost, and what it means for the run hierarchy

This is the second per-run, per-period predictor family, so it interacts with the
`base → climate → SSP` ordering chosen in `00-setup-db.qmd`. Because SSP is the **leaf**,
projected employment is replicated once per (SSP × climate framing) pairing, whereas projected
climate is stored once per climate trajectory:

| | rows per arm | copies | total |
| --- | --- | --- | --- |
| climate (18 yearly indicators × 4 periods × 4.1 M) | 295 M | 4 trajectories | 1.18 B |
| employment (3 sectors × 4 periods × 4.1 M) | 49 M | 5 SSP × 4 traj | 0.98 B |

so the inverted hierarchy costs ~2.2 B rows against ~6.2 B for SSP-on-top — **a factor 2.8, not
the factor 5 that climate alone suggests.** Employment is ~6× lighter per arm than climate, so
the inversion is still clearly the right way round; but the SSP payload is *not* negligible, and
the honest saving is smaller than "SSP tables are tiny" implies.

The practical mitigation is the same as for climate: only write the (SSP × climate) pairings
that will actually be allocated. `runs_to_fill` below defaults to that.

## Read the trajectories

```{r}
#| label: statent-ssp-read
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal

# The CSV is not fetchable over HTTP from a reproducible URL we have been able to verify
# (see the provenance note above), so it is expected in the evoland cache rather than
# downloaded. Its md5 is pinned so a swapped file is caught.
fte_csv <- file.path(getOption("evoland.cachedir"), "nccs", "NCCS_future_FTE.csv")
fte_md5 <- "9c1950fface91f98f7b82b96ca2854a4"

if (!file.exists(fte_csv)) {
  stop(
    "Missing ", fte_csv, "\n",
    "Obtain NCCS_future_FTE.csv from the SSP-CH Zenodo data deposit ",
    "(doi:10.5281/zenodo.8263509) and place it there. See the provenance note in this step."
  )
}
stopifnot(
  "NCCS_future_FTE.csv does not match the pinned md5" =
    tools::md5sum(fte_csv) == fte_md5
)

# The file carries a UTF-8 BOM and CRLF line endings; without header = TRUE, fread reads
# the BOM-prefixed header as a data row and names the columns V1..V8.
fte_raw <- data.table::fread(fte_csv, header = TRUE, encoding = "UTF-8")
data.table::setnames(fte_raw, sub("^\xef\xbb\xbf", "", names(fte_raw)))

# values are thousands of FTE; sectors are Sector1/2/3 -> the fte_sec{n} predictor names
year_cols <- grep("^[0-9]{4}$", names(fte_raw), value = TRUE)
fte_long <-
  data.table::melt(
    fte_raw,
    id.vars = c("Scenario", "Sector", "Canton"),
    measure.vars = year_cols,
    variable.name = "year",
    value.name = "fte_thousands",
    variable.factor = FALSE
  )[, .(
    ssp = Scenario,
    pred_name = sub("^Sector", "fte_sec", Sector),
    canton_abbr = Canton,
    year = as.integer(year),
    fte = fte_thousands * 1000
  )]

stopifnot(
  "expected 5 scenarios" = data.table::uniqueN(fte_long$ssp) == 5L,
  "expected 26 cantons" = data.table::uniqueN(fte_long$canton_abbr) == 26L,
  "expected 3 sectors" = data.table::uniqueN(fte_long$pred_name) == 3L
)

fte_long[, .N, by = .(ssp, pred_name)][order(ssp, pred_name)]
```

## Assign each coordinate to a canton

swissBOUNDARIES3D supplies the cantonal geometry. The FTE table keys cantons by two-letter
abbreviation while the boundary layer carries full names, so we bridge them through
`kantonsnummer`, whose 1–26 ordering is the standard BFS sequence — more robust than matching
names across encodings.

```{r}
#| label: statent-ssp-cantons
sources_boundaries <-
  data.frame(
    url = paste0(
      "https://data.geo.admin.ch/ch.swisstopo.swissboundaries3d/",
      "swissboundaries3d_2026-01/swissboundaries3d_2026-01_2056_5728.gpkg.zip"
    ),
    # sha256 68e922353c76fa5db3cef06a32f9711c0198faa6fbd2b5bcde9edc88b0f8999f
    # matches the STAC multihash 122068E92235...; md5 computed from that archive
    md5sum = "a8f2e3b9bc1d5f5b9a90c768ad99a6ec"
  ) |>
  download_and_verify(target_dir = getOption("evoland.cachedir"))

cantons <- terra::vect(
  paste0(
    "/vsizip/", sources_boundaries[["local_path"]],
    "/swissBOUNDARIES3D_1_5_LV95_LN02.gpkg"
  ),
  layer = "tlm_kantonsgebiet"
)

# BFS canton numbering 1..26; the FTE table's abbreviations in that same order
canton_abbr_by_number <- data.table::data.table(
  kantonsnummer = 1:26,
  canton_abbr = c(
    "ZH", "BE", "LU", "UR", "SZ", "OW", "NW", "GL", "ZG", "FR", "SO", "BS", "BL",
    "SH", "AR", "AI", "SG", "GR", "AG", "TG", "TI", "VD", "VS", "NE", "GE", "JU"
  )
)

stopifnot(
  "unexpected canton count in swissBOUNDARIES3D" = nrow(cantons) == 26L,
  "canton abbreviations do not cover the FTE table" =
    setequal(canton_abbr_by_number$canton_abbr, unique(fte_long$canton_abbr))
)

coord_canton <-
  extract_using_coords_t(cantons[, "kantonsnummer"], coords_minimal)[,
    .(id_coord, kantonsnummer = as.integer(value))
  ][
    canton_abbr_by_number,
    on = "kantonsnummer",
    nomatch = NULL
  ]

message(
  "Assigned a canton to ", nrow(coord_canton), " of ", nrow(coords_minimal), " coordinates."
)
```

## Interpolate trajectories onto the decadal periods

The trajectories are given at 2022/2035/2060/2085/2100; the pipeline's extrapolated periods are
decades 5–8 (2025–34, 2035–44, 2045–54, 2055–64). We evaluate each trajectory at the **mid-year
of each period** by linear interpolation, matching the original's use of `stats::approx()`, and
take ratios against the base year.

```{r}
#| label: statent-ssp-interpolate
periods <- db$periods_t[is_extrapolated == TRUE]
period_midyear <- periods[, .(
  id_period,
  midyear = as.integer(
    (as.integer(format(start_date, "%Y")) + as.integer(format(end_date, "%Y"))) / 2
  )
)]

base_year <- min(fte_long$year) # 2022, the first column of the elicited table

fte_interp <-
  fte_long[,
    {
      approx_fun <- stats::approx(
        x = year, y = fte, xout = c(base_year, period_midyear$midyear), rule = 2
      )
      .(
        year = approx_fun$x,
        fte = approx_fun$y
      )
    },
    by = .(ssp, pred_name, canton_abbr)
  ]

fte_ratio <-
  fte_interp[
    fte_interp[year == base_year, .(ssp, pred_name, canton_abbr, fte_base = fte)],
    on = .(ssp, pred_name, canton_abbr)
  ][
    year != base_year
  ][
    period_midyear,
    on = c(year = "midyear"),
    nomatch = NULL
  ][, .(
    ssp, pred_name, canton_abbr, id_period,
    ratio = data.table::fifelse(fte_base > 0, fte / fte_base, 0)
  )]

# a sanity read: how much does each sector move nationally by the last period?
fte_ratio[
  id_period == max(id_period),
  .(median_ratio = round(median(ratio), 3)),
  by = .(ssp, pred_name)
][order(pred_name, ssp)]
```

## Project and commit

```{r}
#| label: statent-ssp-commit
# Observed levels to scale: the last observed period.
id_period_obs <- max(db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period])

pred_ids <- db$pred_meta_t[name %in% c("fte_sec1", "fte_sec2", "fte_sec3"), .(id_pred, name)]
stopifnot("run 02-ingest-preds-statent.qmd first" = nrow(pred_ids) == 3L)

observed <-
  db$pred_data_t[id_run == 0L & id_period == id_period_obs & id_pred %in% pred_ids$id_pred][
    pred_ids,
    on = "id_pred"
  ][
    coord_canton,
    on = "id_coord",
    nomatch = NULL
  ][, .(id_coord, canton_abbr, pred_name = name, id_pred, value)]

# Only the (SSP x climate) pairings that will actually be allocated; see the storage note.
# Default: every SSP leaf. Narrow this to cut volume.
runs_to_fill <- db$runs_t[!is.na(ssp), .(id_run, ssp)]

for (i in seq_len(nrow(runs_to_fill))) {
  r <- runs_to_fill[i]
  message(sprintf("[%d/%d] run %d (%s)", i, nrow(runs_to_fill), r$id_run, r$ssp))

  projected <-
    observed[
      fte_ratio[ssp == r$ssp],
      on = .(canton_abbr, pred_name),
      allow.cartesian = TRUE,
      nomatch = NULL
    ][, .(
      id_run = r$id_run,
      id_period,
      id_pred,
      id_coord,
      value = value * ratio
    )]

  db$pred_data_t <- as_pred_data_t(projected)
  rm(projected)
  gc(verbose = FALSE)
}

db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key =                          , value =,
  "statent_ssp_source",            "NCCS_future_FTE.csv (SSP-CH project internal)",
  "statent_ssp_source_md5sum",     fte_md5,
  "statent_ssp_source_provenance", "UNKNOWN - elicited, no method documentation found",
  "statent_ssp_method",            paste(
    "cantonal sectoral FTE trajectory interpolated to period mid-years, applied as a",
    "multiplicative ratio to the observed hectare-level pattern at id_period",
    id_period_obs
  ),
  "boundaries_url",                sources_boundaries[["url"]],
  "boundaries_md5sum",             sources_boundaries[["md5sum"]]
))
```
