---
title: "02 — Predictors: Swiss Soil Property Map"
---

Ingest **sand**, **clay** and **organic carbon** from the Swiss Soil Property Map (SSPM) at
four depths as soil predictors, aggregated from the native 30 m grid to the 100 m hectare
grid.

**Source:** Gupta, Hasler & Alewell (2024), "Mining soil data of Switzerland" — Swiss Soil
Property Map, [Zenodo record 7821650](https://zenodo.org/records/7821650)
([doi:10.1016/j.geodrs.2023.e00747](https://doi.org/10.1016/j.geodrs.2023.e00747)). 30 m
GeoTIFFs, EPSG:4326, NoData `-3.4e38`, one file per `{property}_{depth}_{stat}`.

We fetch the **mean** layers only — the `error` (uncertainty) layers are left for a later
soil-uncertainty `id_run`. The same 12 files back
`2026-07-ssp-rsofun/2-forcing-soil-1-download.r`, so the md5sums here are shared with that
step; downloading in either place populates the cache for both.

> **~6 GB of downloads** (12 files × ~510 MB). Budget for it.

## This is only a partial replacement of the EIV soil layers

`TODO.md` framed this as replacing the six EIV soil predictors from
`02-ingest-preds-envidat-eiv.qmd`. It is not a like-for-like swap, and the EIV layers are
**not** removed by this step:

| EIV predictor | SSPM counterpart | Verdict |
| --- | --- | --- |
| `soil_humus` | `soil_oc_*` (organic carbon) | **Superseded** — OC is the direct measurement the EIV only indicates. |
| `soil_moisture`, `soil_moisture_variability`, `soil_aeration` | `soil_sand_*` / `soil_clay_*` | **Partly superseded** — texture drives all three, but only through a pedotransfer function. `2026-07-ssp-rsofun/2-forcing-soil-2-whc.r` derives water-holding capacity this way; until that WHC is ingested here, texture is a *rawer* predictor than the EIVs, not a better one. |
| `soil_ph` | — | **No counterpart.** SSPM as fetched here carries no pH layer. |
| `soil_nutrients` | — | **No counterpart.** The record has N and P layers that the rsofun step explicitly does not fetch; whether they can stand in for the EIV nutrient indicator is untested. |

So retiring the EIV soil layers wholesale would *lose* pH and nutrients. The defensible
sequence is: ingest SSPM alongside the EIVs, let `05-covariate-selection.qmd` score them
against each other per transition, and retire only what loses. That decision is left open in
`TODO.md` rather than taken here.

## Aggregation from 30 m to 100 m

The pipeline's usual idiom, `extract_using_coords_t()`, samples the value at each hectare-cell
*centre*. For a 30 m source that **subsamples** — it keeps roughly one ninth of the data and
inherits the sampled cell's noise. Soil properties are spatially noisy at 30 m, so we
aggregate instead.

`terra::project(x, template, method = "average")` does the CRS change (EPSG:4326 → 2056) and
the area-weighted averaging in one pass onto the exact model grid, which avoids both an
intermediate resampling step and any misalignment between an aggregated grid and `coords_t`.

**Mean is the right aggregation for all three properties.** Sand and clay are compositional
mass fractions and OC is a concentration; all are intensive quantities whose cell average is
the physically meaningful hectare value, and averaging is linear so
`sand + clay + silt ≈ 100 %` survives it. (Contrast a categorical layer, where the mode is
needed — see `02-ingest-preds-bioregions.qmd` — or an extensive quantity like employment
counts, where the sum is.)

Two further aggregations are worth considering, and one is implemented as an option:

- **Within-cell standard deviation** (`derive_heterogeneity`, off by default). Computed as
  `sqrt(E[x²] − E[x]²)` from two `average` projections. A hectare of uniform loam and a
  hectare that is half sand, half clay have the same mean texture but very different
  agricultural value, and that distinction is invisible to the mean. This is speculative as a
  land-use predictor — hence off by default — but it is nearly free once the raster is read.
- **Depth-weighted profile mean.** The four layers are *profile slices*, not spatial
  neighbours, so they are ingested separately rather than averaged: topsoil (0 cm) governs
  cultivation while the deeper layers govern water storage, and collapsing them would destroy
  that. The cost is 12 strongly collinear predictors where the EIVs offered 6. If that proves
  unwieldy in `05`, the alternative is a trapezoidal 0–100 cm mean per property (3 predictors)
  plus the 0 cm topsoil value — noted here, not implemented.

## Download

```{r}
#| label: ingest-soil-download
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal

# Set TRUE to additionally derive within-cell sd predictors (see above).
derive_heterogeneity <- FALSE

zenodo_record <- "7821650"

# {property, depth_cm} -> md5sum for the mean layers, from the Zenodo record listing.
# Shared with 2026-07-ssp-rsofun/2-forcing-soil-1-download.r.
sspm_mean_md5 <- data.table::fread(
  text = "
property depth_cm md5sum
sand 0   f1f714315b2180d062144bb0bed664fe
sand 30  2c7e0c6208c6cdcab312c9dc99e434b2
sand 60  9cb7031f5eae28dab9d31ca88932134b
sand 100 59a3e79208af5420300434aafdd68e47
clay 0   9d2739c0fd5347fa927098f9927cb952
clay 30  34fa92871bb7708f4ff0899930ebb580
clay 60  cfbc5cd5ec4787efe75d8cc4311807de
clay 100 b395942a5fdd0196fb371b2058197f65
OC   0   2984b33c24a695e95e133447a9014d41
OC   30  f5bde0ba345558da5068d004922f2997
OC   60  d80401aea525190d6d2e8cf095dbbc07
OC   100 40792da49a13a221fcb5f70f53cc7a30
",
  colClasses = list(character = "property", integer = "depth_cm", character = "md5sum")
)

sspm_mean_md5[, filename := sprintf("%s_%dcm_mean_30m.tif", property, depth_cm)]
sspm_mean_md5[,
  url := sprintf("https://zenodo.org/records/%s/files/%s?download=1", zenodo_record, filename)
]
sspm_mean_md5[, pred_name := sprintf("soil_%s_%dcm", tolower(property), depth_cm)]

sources_soil <-
  sspm_mean_md5[, .(url, md5sum)] |>
  download_and_verify()

sspm <- merge(
  sspm_mean_md5,
  data.table::as.data.table(sources_soil)[, .(url, md5sum, local_path)],
  by = c("url", "md5sum")
)
```

## Per-property metadata

```{r}
#| label: ingest-soil-specs
prov <- paste(
  "Swiss Soil Property Map (Gupta, Hasler & Alewell 2024), 30 m native resolution,",
  "area-weighted mean aggregated to the 100 m model grid."
)

property_specs <- list(
  sand = list(
    unit = "%",
    pretty_name = "Sand content",
    description = paste("Sand mass fraction of the fine earth.", prov)
  ),
  clay = list(
    unit = "%",
    pretty_name = "Clay content",
    description = paste("Clay mass fraction of the fine earth.", prov)
  ),
  oc = list(
    unit = "g kg-1",
    pretty_name = "Soil organic carbon",
    description = paste(
      "Soil organic carbon concentration. Direct measurement of what the EIV",
      "`soil_humus` indicator only proxies.",
      prov
    )
  )
)
```

## Read → aggregate → write

One file at a time: the sources are ~510 MB each and terra holds the projection result in
memory, so a loop keeps the footprint flat.

```{r}
#| label: ingest-soil-load
# exact model grid: 100 m, EPSG:2056, aligned with coords_t
template <- db$lulc_data_as_rast(id_period = 1)
  
orig_format <- paste(
  "SSPM 30 m GeoTIFF (EPSG:4326), reprojected to EPSG:2056 and area-weighted averaged",
  "onto the 100 m model grid via terra::project(method = 'average')"
)

for (i in seq_len(nrow(sspm))) {
  row <- sspm[i]
  prop_key <- tolower(row[["property"]])
  spec <- property_specs[[prop_key]]
  stopifnot("no spec for property" = !is.null(spec))

  message(sprintf("[%d/%d] %s", i, nrow(sspm), row[["pred_name"]]))

  r <- terra::rast(row[["local_path"]])
  terra::NAflag(r) <- -3.4e38

  # CRS change + area-weighted aggregation in one pass
  mean_2056 <- terra::project(r, template, method = "average")
  terra::set.names(mean_2056, row[["pred_name"]])

  db$add_predictor(
    pred_data_raw = extract_using_coords_t(mean_2056, coords_minimal)[,
      .(id_coord, id_period = 0L, value)
    ],
    name = row[["pred_name"]],
    fill_value = NA,
    unit = spec[["unit"]],
    pretty_name = sprintf("%s (%d cm)", spec[["pretty_name"]], row[["depth_cm"]]),
    orig_format = orig_format,
    description = spec[["description"]],
    sources = data.table(url = row[["url"]], md5sum = row[["md5sum"]])
  )

  if (derive_heterogeneity) {
    # sd = sqrt(E[x^2] - E[x]^2), both terms area-weighted over the same cells
    sq_2056 <- terra::project(r^2, template, method = "average")
    sd_2056 <- terra::math(sq_2056 - mean_2056^2, "sqrt")
    terra::set.names(sd_2056, paste0(row[["pred_name"]], "_sd"))

    db$add_predictor(
      pred_data_raw = extract_using_coords_t(sd_2056, coords_minimal)[,
        .(id_coord, id_period = 0L, value)
      ],
      name = paste0(row[["pred_name"]], "_sd"),
      fill_value = NA,
      unit = spec[["unit"]],
      pretty_name = sprintf("%s (%d cm), within-cell sd", spec[["pretty_name"]], row[["depth_cm"]]),
      orig_format = orig_format,
      description = paste(
        "Within-hectare standard deviation of",
        spec[["pretty_name"]],
        "across the 30 m source cells; a soil-heterogeneity indicator.",
        prov
      ),
      sources = data.table(url = row[["url"]], md5sum = row[["md5sum"]])
    )
    rm(sq_2056, sd_2056)
  }

  rm(r, mean_2056)
  gc(verbose = FALSE)
}

message("Done. Ingested ", nrow(sspm), " SSPM soil predictors at id_period = 0.")
```

## Sanity check

Sand and clay are fractions of the same fine earth, so their sum must stay below 100 %; a
violation would mean the aggregation or the NoData handling went wrong.

```{r}
#| label: ingest-soil-check
soil_wide <-
  db$pred_meta_t[grepl("^soil_(sand|clay)_", name), .(id_pred, name)][
    db$pred_data_t[id_period == 0L],
    on = "id_pred",
    nomatch = NULL
  ] |>
  data.table::dcast(id_coord ~ name, value.var = "value")

for (depth in c(0L, 30L, 60L, 100L)) {
  cols <- sprintf(c("soil_sand_%dcm", "soil_clay_%dcm"), depth)
  if (!all(cols %in% names(soil_wide))) next
  frac_sum <- rowSums(soil_wide[, ..cols], na.rm = TRUE)
  message(sprintf(
    "%3d cm: sand+clay range [%.1f, %.1f] %%, n over 100%%: %d",
    depth, min(frac_sum, na.rm = TRUE), max(frac_sum, na.rm = TRUE), sum(frac_sum > 100.5)
  ))
}
```
