---
title: "02 — Predictors: EnviDat EIV (SPEEDMIND)"
---

Ingest spatial **ecological indicator values (EIVs)** as biophysical predictors.

**Source:** Descombes et al. (2020), *Spatial modelling of ecological indicator values
improves predictions of plant distributions in complex landscapes*, Ecography 43(10),
1448–1463, [doi:10.1111/ecog.05117](https://doi.org/10.1111/ecog.05117). Data:
[doi:10.16904/ENVIDAT.153](https://doi.org/10.16904/ENVIDAT.153). The EIV operationalisation
follows Landolt et al. (2010), *Flora indicativa* (ISBN 9783258074610).

**What the layers actually are.** All `SPEEDMIND_Soil*` layers are **continuous
community-weighted means (CWMs)** of the underlying Landolt EIV classes across plant
communities — *not* discrete class indices. `SoilR` is the exception: it is stored as
modelled **soil pH** (≈4.7–7.1 for Switzerland) rather than the Landolt R class (1–5).

The EIV codes (Landolt et al. 2010): **R** = soil pH · **N** = nutrients · **F** = moisture
· **W** = moisture variability · **D** = aeration · **H** = humus · **L** = light ·
**K** = continentality.

> **To be replaced.** The soil layers (`soil_*`) are slated for replacement by the Swiss
> Soil Property Map ingested in `2026-07-ssp-rsofun/2-forcing-soil-1-download.r` — see
> `TODO.md`.

```{r}
#| label: ingest-envidat-eiv-download
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal
extent_wide <- db$extent |> terra::extend(1000)

sources_eiv <-
  list(
    list(
      url = "https://www.envidat.ch/dataset/4ab13d14-6f96-41fd-96b0-b3ea45278b3d/resource/81c046c3-8d1d-45bc-a833-7d8240cebd12/download/predictors_description.xlsx",
      md5sum = "9a49a27141863f37a5c39c87509f20c7"
    ),
    list(
      url = "https://www.envidat.ch/dataset/4ab13d14-6f96-41fd-96b0-b3ea45278b3d/resource/e0faab13-0d1b-492a-8539-5370d48b9e35/download/predictors.zip",
      md5sum = "a8e3bd3a7e929e48a73e7df293ea735d"
    )
  ) |>
  data.table::rbindlist() |>
  download_and_verify()

zip_path <- sources_eiv[["local_path"]][[2L]]
```

## Predictor specs

One spec per EIV layer; `zip_file` names are confirmed from `unzip(zip_path, list = TRUE)`.
**Continentality (`SoilK`) is intentionally omitted** (kept commented): it is very likely
redundant with — or weaker than — the CH2025 climatological predictors, and the
to-be-added bioregions (see `TODO.md`) are likely collinear with it as well.

```{r}
#| label: ingest-envidat-eiv-specs
eiv_specs <- list(
  soil_ph = list(
    zip_file = "Predictors/SPEEDMIND_SoilR.tif",
    unit = "pH",
    pretty_name = "Soil pH (EIV-R derived)",
    description = paste0(
      "Modelled soil pH (approx. 4.7-7.1 for Switzerland); derived from Landolt EIV-R ",
      "but stored as pH units, not the 1-5 class index. ",
      "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
    )
  ),
  soil_nutrients = list(
    zip_file = "Predictors/SPEEDMIND_SoilN.tif",
    unit = "Landolt N CWM",
    pretty_name = "Soil nutrients (EIV-N)",
    description = paste0(
      "Community-weighted mean of Landolt EIV-N across plant communities; ",
      "gradient from nutrient-poor (1) to nutrient-rich (5), mainly nitrogen. ",
      "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
    )
  ),
  soil_moisture = list(
    zip_file = "Predictors/SPEEDMIND_SoilF.tif",
    unit = "Landolt F CWM",
    pretty_name = "Soil moisture (EIV-F)",
    description = paste0(
      "Community-weighted mean of Landolt EIV-F across plant communities; ",
      "gradient from very dry soils (1) to plants growing in water (5). ",
      "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
    )
  ),
  soil_moisture_variability = list(
    zip_file = "Predictors/SPEEDMIND_SoilW.tif",
    unit = "Landolt W CWM",
    pretty_name = "Soil moisture variability (EIV-W)",
    description = paste0(
      "Community-weighted mean of Landolt EIV-W across plant communities; ",
      "gradient from low intraannual variability in soil moisture (1) to high (3). ",
      "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
    )
  ),
  soil_aeration = list(
    zip_file = "Predictors/SPEEDMIND_SoilD.tif",
    unit = "Landolt D CWM",
    pretty_name = "Soil aeration (EIV-D)",
    description = paste0(
      "Community-weighted mean of Landolt EIV-D across plant communities; ",
      "gradient from waterlogged/low-aerated soils (1) to soils rich in rocks or sand (5). ",
      "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
    )
  ),
  soil_humus = list(
    zip_file = "Predictors/SPEEDMIND_SoilH.tif",
    unit = "Landolt H CWM",
    pretty_name = "Soil humus (EIV-H)",
    description = paste0(
      "Community-weighted mean of Landolt EIV-H across plant communities; ",
      "gradient from humus-poor soils (1) to humus-rich soils (5). ",
      "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
    )
  ),
  light_100m = list(
    zip_file = "Predictors/SPEEDMIND_SoilL.tif",
    unit = "Landolt L CWM",
    pretty_name = "Light (EIV-L)",
    description = paste0(
      "Community-weighted mean of Landolt EIV-L across plant communities; ",
      "gradient from shaded areas (1) to sunny areas (5). ",
      "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
    )
  )
  # continentality_100m = list(
  #   zip_file = "Predictors/SPEEDMIND_SoilK.tif",
  #   unit = "Landolt K CWM",
  #   pretty_name = "Continentality (EIV-K)",
  #   description = paste0(
  #     "Community-weighted mean of Landolt EIV-K across plant communities; ",
  #     "gradient from atlantic climate (1; high humidity, mild winters) to ",
  #     "continental climate (5; low humidity, cold winters). ",
  #     "Mapped by Descombes et al. 2020, doi:10.1111/ecog.05117"
  #   )
  # )
)
```

## Reproject, extract, register

The source GeoTIFFs are 93 m in World Mercator (EPSG:3395); each is bilinear-resampled to
the EPSG:2056 100 m grid, sampled at the model coordinates, and written at `id_period = 0`.

```{r}
#| label: ingest-envidat-eiv-register
for (pred_name in names(eiv_specs)) {
  spec <- eiv_specs[[pred_name]]
  tif_path <- paste0("/vsizip/", zip_path, "/", spec[["zip_file"]])

  r <- terra::rast(tif_path) |>
    terra::project("EPSG:2056", method = "bilinear", res = 100) |>
    terra::crop(extent_wide)
  terra::set.names(r, pred_name)

  pred_data <- extract_using_coords_t(r, coords_minimal)

  db$add_predictor(
    pred_data_raw = pred_data[, .(id_coord, id_period = 0L, value)],
    name = pred_name,
    fill_value = NA,
    unit = spec[["unit"]],
    pretty_name = spec[["pretty_name"]],
    orig_format = "93m EPSG:3395 World Mercator GeoTIFF, bilinear resampled to EPSG:2056 100m",
    description = spec[["description"]],
    sources = sources_eiv[, .(url, md5sum)]
  )
}
```
