---
title: "02 — CH2025 climate: probe & download"
---

Discover and download the CH2025 climate-indicator netCDFs from the MeteoSwiss
[CH2025 web atlas](https://www.meteoschweiz.admin.ch/service-und-publikationen/applikationen/ext/climate-ch2025-maps.html).
This is stage 1 (**probe**) + stage 2 (**download**); the ETL that reads the cached files
into the DB is `02-ingest-preds-ch2025-2-etl.qmd`. That ETL recovers all dimensions from
the cached **file names**, so nothing is passed between the two steps in memory.

**Two product families** share the same indicator set:

- `-gwl` — projections, indexed by global warming level × uncertainty quantile (annual
  aggregate only).
- `-obs` — observed climatologies, indexed by reference period × time of year (four seasons
  + annual; no uncertainty quantile).

Not every indicator × dimension combination is published, so candidate URLs are first
**probed** with a HEAD request and only the live ones are downloaded. Background on the GWL
structure is in `2-ingest-preds-ch2025-todo.md`; the reverse-engineered URL patterns are in
`2-ingest-preds-ch2025-urls.md`.

> **Out of scope (TODO).** SSP5-8.5 late-century (~2071–2100) implies ~5–6 °C of warming,
> exceeding the highest GWL probed here (GWL3.0); those steps cannot be mapped to an
> existing CH2025 aggregate.

## Parameters

The dimension levels and the category → `scenarioIndicator` mapping, as observed from the
atlas URL patterns. The same indicator sets back both the `-gwl` and `-obs` families.

```{r}
#| label: ch2025-download-params
library(data.table)
library(evoland)

base_url <- "https://service.meteoswiss.ch/pbbackend/api/v1/products"

gwl_levels <- c("GWL1.5", "GWL2.0", "GWL3.0")
uncertainty_levels <- c("q5", "q50", "q95")
reference_periods <- c("1961-1990", "1991-2020")
times_of_year <- c("DJF", "MAM", "JJA", "SON", "yearly")

category_indicators <-
  tibble::tribble(
    ~category, ~indicator,
    "cold", c("ID", "FD", "HDD", "HED"),
    "heat", c("HD", "HW2", "HW3", "HW4", "CoDD", "COD", "VHD", "SD", "TN"),
    "precip", c("pr", "PR20", "PR40", "PR60"),
    "snow", c("SNFD"),
    "temperature", c("tas", "tasmax", "tasmin"),
    "drought", c("CDD")
  ) |>
  tidyr::unnest_longer(indicator)
```

## Build the candidate URL matrix

Projections are indicators × (GWL × uncertainty), annual only; observations are indicators
× (reference period × time of year), no uncertainty. `bind_rows` leaves each family's
inapplicable dimensions as `NA` (reference period for `-gwl`; GWL/uncertainty for `-obs`).

```{r}
#| label: ch2025-download-candidates
gwl_candidates <-
  category_indicators |>
  dplyr::cross_join(tidyr::expand_grid(
    gwl = gwl_levels,
    uncertainty = uncertainty_levels
  )) |>
  dplyr::mutate(
    product = paste0("climate-ch2025-maps-", category, "-gwl"),
    time_of_year = "yearly"
  )

obs_candidates <-
  category_indicators |>
  dplyr::cross_join(tidyr::expand_grid(
    reference_period = reference_periods,
    time_of_year = times_of_year
  )) |>
  dplyr::mutate(
    product = paste0("climate-ch2025-maps-", category, "-obs")
  )

candidates <- dplyr::bind_rows(gwl_candidates, obs_candidates)
```

## Rate limiting

Politeness cap on how hard we hit the MeteoSwiss host: capacity tokens refill over
`fill_time_s`, sustaining ~2 req/s with a small burst. `req_throttle` buckets by host
(default realm), so the probe and download stages share one budget and
`req_perform_parallel` honours it across concurrent requests.

```{r}
#| label: ch2025-download-throttle
throttle <- function(req) {
  httr2::req_throttle(req, capacity = 10, fill_time_s = 5)
}
```

## Stage 1 — probe

Build one HEAD request per candidate and probe in parallel. `req_error(is_error = FALSE)`
stops httr2 throwing on 4xx/5xx, so `req_perform_parallel` returns a response object for
every row. `req_url_query()` handles percent-encoding (including the slash in
`application/netcdf` → `application%2Fnetcdf`), and `NA`-valued query parameters are dropped
so one builder serves both families.

```{r}
#| label: ch2025-download-probe
make_request <- function(product, indicator, gwl, uncertainty, reference_period, time_of_year) {
  drop_na <- function(x) if (is.na(x)) NULL else x
  httr2::request(paste0(base_url, "/", product, "/realizations")) |>
    httr2::req_url_query(
      language = "de",
      globalWarmingLevel = drop_na(gwl),
      referencePeriod = drop_na(reference_period),
      timeOfYear = time_of_year,
      uncertainty = drop_na(uncertainty),
      productType = "data",
      productName = product,
      scenarioIndicator = indicator,
      mediaType = "application/netcdf"
    )
}

probe_reqs <- purrr::pmap(
  candidates,
  function(product, indicator, gwl, uncertainty, reference_period, time_of_year, ...) {
    make_request(product, indicator, gwl, uncertainty, reference_period, time_of_year) |>
      httr2::req_method("HEAD") |>
      httr2::req_error(is_error = \(r) FALSE) |>
      throttle()
  }
)

# Store the fully-built URLs extracted from the request objects
candidates$url <- purrr::map_chr(probe_reqs, \(r) r$url)

message("Probing ", length(probe_reqs), " URLs...")
probe_resps <- httr2::req_perform_parallel(probe_reqs, on_error = "return")

is_live <- function(r) {
  !inherits(r, "error") && httr2::resp_status(r) < 400L
}
candidates$live <- vapply(probe_resps, is_live, logical(1))

message(sum(candidates$live), " / ", nrow(candidates), " URLs live")
print(dplyr::count(candidates, product, live), n = Inf)
```

## Stage 2 — download

Download only the live candidates. The file name encodes the `-gwl`/`-obs` suffix plus the
dimensions that apply to that family (the other family's `NA`s are dropped); the encoding
must stay **reversible** because the ETL parses these tokens back out. Already-cached files
are skipped.

```{r}
#| label: ch2025-download-fetch
cache_dir <- file.path(getOption("evoland.cachedir"), "ch2025")
evoland:::ensure_dir(cache_dir)

nc_filename <- function(product, indicator, gwl, uncertainty, reference_period, time_of_year) {
  stem <- sub("^climate-ch2025-maps-", "", product)
  tokens <- c(indicator, gwl, uncertainty, reference_period, time_of_year)
  paste0(stem, "_", paste(tokens[!is.na(tokens)], collapse = "_"), ".nc")
}

download_nc <- function(url, product, indicator, gwl, uncertainty, reference_period, time_of_year) {
  fname <- nc_filename(product, indicator, gwl, uncertainty, reference_period, time_of_year)
  dest <- file.path(cache_dir, fname)
  if (file.exists(dest)) {
    message("Cached:      ", fname)
    return(dest)
  }
  message("Downloading: ", fname)
  httr2::request(url) |> throttle() |> httr2::req_perform(path = dest)
  dest
}

live_candidates <- dplyr::filter(candidates, live)

live_candidates$local_path <- purrr::pmap_chr(
  live_candidates,
  function(url, product, indicator, gwl, uncertainty, reference_period, time_of_year, ...) {
    download_nc(url, product, indicator, gwl, uncertainty, reference_period, time_of_year)
  }
)

message("Done. ", nrow(live_candidates), " files written to ", cache_dir)
```
