---
title: "02 — Predictors: population"
---

Ingest a **mean municipal population** predictor per period. This is more involved than the
other predictors because municipal boundaries and identities change over time: the raw
population series is keyed by BFS municipality number, but municipalities are renamed and
merged across the 1985–2020 window, so the series must be **reconciled to a single set of
BFS numbers** before it can be joined to boundary polygons and rasterised.

**Sources:** BFS *Demographic balance by institutional units* (PX) + swissBOUNDARIES3D
municipal boundaries. (The PX file could instead be pulled from the pxweb API —
`px-x-0102020000_201` — via the `pxweb` package.)

```{r}
#| label: ingest-pop-download
library(evoland)
library(terra)

sources_pop <-
  list(
    list(
      # Demographic balance by institutional units.
      # Alternative: pxweb API
      # "https://www.pxweb.bfs.admin.ch/api/v1/en/px-x-0102020000_201/px-x-0102020000_201.px"
      url = "https://dam-api.bfs.admin.ch/hub/api/dam/assets/23164063/master",
      md5sum = "528209137ea5c74a30e477a1ddc3da84"
    ),
    list(
      # swissboundaries3d
      url = "https://data.geo.admin.ch/ch.swisstopo.swissboundaries3d/swissboundaries3d_2021-07/swissboundaries3d_2021-07_2056_5728.shp.zip",
      md5sum = "6d217867b7bf6dfb2927b12f12481a8b"
    )
  ) |>
  data.table::rbindlist() |>
  download_and_verify()
```

## Read and filter the PX table

Keep the total-population series: end-of-year stock ("Bestand am 31. Dezember"), all
nationalities (Swiss + foreign), both sexes.

```{r}
#| label: ingest-pop-read
px_data <- as.data.frame(pxR::read.px(
  sources_pop$local_path[1]
))

# Total population, end-of-year stock, nationality total, sex total
raw_mun_popdata <- px_data[
  px_data$Demografische.Komponente == "Bestand am 31. Dezember" &
    px_data$Staatsangehörigkeit..Kategorie. == "Staatsangehörigkeit (Kategorie) - Total" &
    px_data$Geschlecht == "Geschlecht - Total",
]
```

## Clean municipality identifiers

Municipality records carry the BFS number inside the name string. Extract it into its own
column, strip it (and stray periods) from the name, pivot years to columns, and keep only
municipalities that still existed in 2021 (population > 0).

```{r}
#| label: ingest-pop-clean
# Identify municipality records by the numeric contained in their name
raw_mun_popdata <- raw_mun_popdata[
  grepl(
    ".*?([0-9]+).*",
    raw_mun_popdata$Kanton.......Bezirk........Gemeinde.........
  ),
  c(4:6)
]
names(raw_mun_popdata) <- c("Name_Municipality", "Year", "Population")
raw_mun_popdata <- raw_mun_popdata |>
  tidyr::pivot_wider(
    names_from = "Year",
    values_from = "Population"
  )
# Remove the periods in the name column
raw_mun_popdata$Name_Municipality <- gsub(
  "[......]",
  "",
  as.character(raw_mun_popdata$Name_Municipality)
)

# Separate BFS number from name
raw_mun_popdata$BFS_NUM <- as.numeric(gsub(
  ".*?([0-9]+).*",
  "\\1",
  raw_mun_popdata$Name_Municipality
))

# Remove BFS number from name
raw_mun_popdata$Name_Municipality <- gsub(
  "[[:digit:]]",
  "",
  raw_mun_popdata$Name_Municipality
)

# subset to only municipalities existing in 2021
raw_mun_popdata <- raw_mun_popdata[raw_mun_popdata$`2021` > 0, ]
```

## Municipal boundaries

Read the swissBOUNDARIES3D *Hoheitsgebiet* (territory) layer and keep Swiss municipal
polygons only.

> **To verify (TODO):** the shapefile is `swissBOUNDARIES3D_1_5_…`; confirm the `1_3` vs
> `1_5` product versions are interchangeable for our purposes.

```{r}
#| label: ingest-pop-boundaries
hoheitsgebiet_path <- paste0(
  "/vsizip/",
  sources_pop$local_path[2],
  "/swissBOUNDARIES3D_1_5_TLM_HOHEITSGEBIET.shp"
)

muni_shp <- terra::vect(hoheitsgebiet_path)

# filter out non-swiss municipalities
muni_shp <- muni_shp[
  muni_shp$ICC == "CH" &
    muni_shp$OBJEKTART == "Gemeindegebiet",
]
```

## Reconcile municipality mutations

Municipal identities drift over 1981–2022 through **name changes** and **mergers**. We
scrape the BFS mutation register (name-change mutations only — mergers are captured by the
boundary shapefile), remap each population row's BFS number to its current one, and where a
remap collapses several old municipalities onto one current BFS number, **sum** their
populations (a merger).

```{r}
#| label: ingest-pop-mutations
# Municipality mutations 01/01/1981–01/05/2022, name-change mutations only
content <- rvest::read_html(paste0(
  "https://www.agvchapp.bfs.admin.ch/de/mutated-communes/results?",
  "EntriesFrom=01.01.1981&EntriesTo=01.05.2022&NameChange=True"
))
muni_mutations <- rvest::html_table(content, fill = TRUE)[[1]]

# remove 1st row because it contains additional column names
muni_mutations <- muni_mutations[-1, ]

colnames(muni_mutations) <- c(
  "Mutation_Number",
  "Pre_canton_ID",
  "Pre_District_num",
  "Pre_BFS_num",
  "Pre_muni_name",
  "Post_canton_ID",
  "Post_district_num",
  "Post_BFS_num",
  "Post_muni_name",
  "Change_date"
)

# identify which municipalities have mutations associated with them
mutation_index <- match(raw_mun_popdata$BFS_NUM, muni_mutations$Pre_BFS_num)

# where a row has a mutation, adopt the post-mutation BFS number; else leave it
for (i in seq_len(nrow(raw_mun_popdata))) {
  if (!is.na(mutation_index[i])) {
    raw_mun_popdata$BFS_NUM[[i]] <- muni_mutations[[
      mutation_index[[i]],
      "Post_BFS_num"
    ]]
  }
}

# if remapping produced duplicate BFS numbers, those are mergers -> sum populations
if (length(unique(raw_mun_popdata$BFS_NUM)) != nrow(raw_mun_popdata)) {
  # get the indices of columns that represent the years
  time_points <- na.omit(as.numeric(gsub(
    ".*?([0-9]+).*",
    "\\1",
    colnames(raw_mun_popdata)
  )))

  # create a empty df for results
  muni_pop_final <- as.data.frame(matrix(
    ncol = length(time_points),
    nrow = length(unique(raw_mun_popdata$BFS_NUM))
  ))
  colnames(muni_pop_final) <- time_points

  # Add column for BFS number
  muni_pop_final$BFS_NUM <- sort(unique(raw_mun_popdata$BFS_NUM))

  # loop over date cols and rows summing values where BFS number is non-unique
  for (j in time_points) {
    for (i in seq_along(unique(raw_mun_popdata$BFS_NUM))) {
      muni_pop_final[i, paste(j)] <- sum(
        raw_mun_popdata[
          raw_mun_popdata$BFS_NUM == muni_pop_final[i, "BFS_NUM"],
          paste(j)
        ]
      )
    }
  }
  # replace old data with revised data
  raw_mun_popdata <- muni_pop_final
}
```

## Aggregate to periods, rasterize and register

Average each municipality's annual population within each decadal period, then join the
per-period values onto the boundary polygons and extract at the model coordinates.

> The per-period **mean** assumes a roughly regular time series; an uneven set of survey
> years would skew it slightly. Judged not worth correcting for now.

```{r}
#| label: ingest-pop-register
db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal
periods <- db$periods_t

agg_mun_popdata <-
  raw_mun_popdata |>
  tibble::as_tibble() |>
  dplyr::select(bfs_num = BFS_NUM, tidyselect::matches("[0-9]{4}")) |>
  tidyr::pivot_longer(
    cols = -bfs_num,
    names_to = "year",
    names_transform = as.integer,
    values_to = "no_inhabitants"
  ) |>
  dplyr::mutate(year = lubridate::make_date(year = year)) |>
  dplyr::inner_join(
    periods,
    by = dplyr::join_by(
      year >= start_date,
      year < end_date
    )
  ) |>
  dplyr::group_by(id_period, bfs_num) |>
  dplyr::summarise(
    mean_inhabitants = mean(no_inhabitants),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = id_period,
    values_from = mean_inhabitants
  )


munis_pops <- terra::merge(
  x = muni_shp[, c("BFS_NUMMER")],
  by.x = "BFS_NUMMER",
  y = agg_mun_popdata,
  by.y = "bfs_num"
)

predvals <- extract_using_coords_t(
  munis_pops[, -1], # drop BFS_NUMMER attribute
  coords_minimal
)

db$add_predictor(
  pred_data_raw = predvals[,
    .(id_coord, id_period = as.integer(attribute), value)
  ],
  name = "pop",
  fill_value = 0,
  unit = "mean_pop",
  pretty_name = "Mean Population over a given period",
  orig_format = "number per commune per year",
  description = "The FSO statistic for each commune's inhabitants is averaged over each period",
  sources = sources_pop[, .(url, md5sum)]
)
```
