---
title: "02 — Predictors: swissTLM3D distances"
---

Derive distance-to-nearest-feature predictors (lakes, rivers, roads) from swissTLM3D.

**Source:** swisstopo **swissTLM3D** (`ch.swisstopo.swisstlm3d`), the successor to
VECTOR25 and GWN07.
<https://www.swisstopo.admin.ch/en/topographic-landscape-model-swisstlm3d> ·
[STAC](https://data.geo.admin.ch/browser/index.html#/collections/ch.swisstopo.swisstlm3d)

**Lineage note.** The original `valpar-local.r` derived lake and river distances from
GWN07 ([geocat record](https://www.geocat.ch/geonetwork/srv/api/records/0351bc2e-3cdc-4e8a-b422-0142e494e7b4)),
which is no longer directly downloadable (WMS/WMTS only); swissTLM3D is its current
successor. The road distance was (probably) originally from swissTLM3D via
valpar.ch / Speedmind.

```{r}
#| label: ingest-swisstlm3d-download
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")
coords_minimal <- db$coords_minimal
extent_wide <- db$extent |> terra::extend(1000)

sources_tlm3d <-
  list(
    list(
      url = "https://data.geo.admin.ch/ch.swisstopo.swisstlm3d/swisstlm3d_2026-02-24/swisstlm3d_2026-02-24_2056_5728.gpkg.zip",
      md5sum = "b02f23b984cd848d63743005e7e75ce3" # the md5 of the zip, not the gpkg
    )
  ) |>
  data.table::rbindlist() |>
  download_and_verify()
```

## Unpack the GeoPackage

Reading vector layers directly from a zipped GPKG over `/vsizip/` is too slow, so the
archive is unpacked once and the ZIP deleted (this raises the cache footprint from ~4 to
~10 GiB). The step is skipped if the `.gpkg` is already unpacked in the cache.

```{r}
#| label: ingest-swisstlm3d-unpack
if (grepl("zip$", sources_tlm3d[["local_path"]])) {
  # skipped if we have the gpkg cached already
  zip_path <- sources_tlm3d[["local_path"]]
  zip_dir <- dirname(zip_path)
  unzip(zip_path, exdir = zip_dir)
  unlink(zip_path)

  sources_tlm3d[["local_path"]] <- list.files(zip_dir, pattern = "\\.gpkg$", full.names = TRUE)
}

gpkg_path <- sources_tlm3d[["local_path"]] # inspect layers with terra::vector_layers(gpkg_path)
```

## Read layers

Lakes are standing-water polygons; rivers are flowing watercourse lines. `tlm_strassen_strasse`
holds **all** road classes (Autobahn through Fahrweg/track); no `OBJEKTART` filter is
applied, matching the "distance to nearest road of any class" intent of the original
predictor. Filter by `OBJEKTART` if a class-specific distance is ever wanted.

```{r}
#| label: ingest-swisstlm3d-layers
lakes <- terra::vect(gpkg_path, layer = "tlm_gewaesser_stehendes_gewaesser")
rivers <- terra::vect(gpkg_path, layer = "tlm_gewaesser_fliessgewaesser")
roads <- terra::vect(gpkg_path, layer = "tlm_strassen_strasse")
```

## Rasterize and compute distances

Each layer is rasterized to a binary presence raster on a 100 m template aligned to the
analysis extent, then the 1-argument `terra::distance()` computes, for every empty cell,
the distance to the nearest present cell. That is markedly faster than
`terra::distance(template, vector)` for dense networks like the road and river layers.

```{r}
#| label: ingest-swisstlm3d-distance
template_100m <- terra::rast(extent_wide, resolution = 100, crs = "EPSG:2056")

lakes_rast <- terra::rasterize(lakes, template_100m, field = 1)
rivers_rast <- terra::rasterize(rivers, template_100m, field = 1)
roads_rast <- terra::rasterize(roads, template_100m, field = 1)

dist_lakes <- terra::distance(lakes_rast)
dist_rivers <- terra::distance(rivers_rast)
dist_roads <- terra::distance(roads_rast)
```

## Register predictors

```{r}
#| label: ingest-swisstlm3d-register
pred_specs <- list(
  distance_to_lakes_mean_100m = list(
    rast = dist_lakes,
    unit = "m",
    pretty_name = "Distance to nearest lake (m)",
    description = paste0(
      "Euclidean distance (m) to the nearest standing water body polygon ",
      "(tlm_gewaesser_stehendes_gewaesser) from swissTLM3D 2026-02-24 at 100m. ",
      "Successor to GWN07 (ch.swisstopo.vec25-gewaessernetz_referenz)."
    )
  ),
  distance_to_rivers_mean_100m = list(
    rast = dist_rivers,
    unit = "m",
    pretty_name = "Distance to nearest river or stream (m)",
    description = paste0(
      "Euclidean distance (m) to the nearest flowing watercourse line ",
      "(tlm_gewaesser_fliessgewaesser) from swissTLM3D 2026-02-24 at 100m. ",
      "Successor to GWN07 (ch.swisstopo.vec25-gewaessernetz_referenz)."
    )
  ),
  distance_to_roads_mean_100m = list(
    rast = dist_roads,
    unit = "m",
    pretty_name = "Distance to nearest road of any class (m)",
    description = paste0(
      "Euclidean distance (m) to the nearest road line of any class ",
      "(tlm_strassen_strasse, all OBJEKTART values) from swissTLM3D 2026-02-24 at 100m."
    )
  )
)

for (pred_name in names(pred_specs)) {
  spec <- pred_specs[[pred_name]]
  r <- spec[["rast"]]
  terra::set.names(r, pred_name)

  pred_data <- extract_using_coords_t(r, coords_minimal)

  db$add_predictor(
    pred_data_raw = pred_data[, .(id_coord, id_period = 0L, value)],
    name = pred_name,
    fill_value = NA,
    unit = spec[["unit"]],
    pretty_name = spec[["pretty_name"]],
    orig_format = "swissTLM3D GeoPackage (EPSG:2056), rasterized to 100m, Euclidean distance",
    description = spec[["description"]],
    sources = sources_tlm3d[, .(url, md5sum)]
  )
}
```
