---
title: "03 — Neighbourhood predictors"
---

Generate neighbourhood predictors: for each cell, summaries of the surrounding cells
within distance bands. These capture local spatial context (e.g. how much of a class sits
nearby) that point predictors miss.

Currently only **land-use categories** are used as the neighbour signal; extending this to
other predictors is open (see `TODO.md`).

## Neighbour relations

`set_neighbors` builds, for every coordinate, the set of neighbours out to `max_distance`,
bucketed into the given distance bands. This is the expensive part (a few minutes over the
full grid); it is stored in the DB so downstream steps reuse it.

```{r}
#| label: neighbors-relations
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")

# attn this may take a couple minutes
db$set_neighbors(
  max_distance = 1000,
  distance_breaks = c(0, 100, 500, 1000)
)
```

## Derive predictors

Given the stored relations, deriving the neighbour predictors themselves is fast.

```{r}
#| label: neighbors-generate
db$generate_neighbor_predictors()
```
