---
title: "04 — Viable transition identification"
---

Commit the set of **viable** transitions to model: `trans_meta_t` with `is_viable` set by a
minimum observed-transition count (`min_cardinality_abs`). The threshold is chosen from the
observed-transition diagnostic in `04d-viable-transition-identification.qmd` — see that report to
justify the value.

## `static` is no longer excluded as an anterior class

Earlier revisions passed `exclude_anterior = lulc_meta[name == "static", id_lulc]`, making
`static` a pure sink: transitions could enter it, none could leave.

**NCCS-SSP did not treat it as a sink either, because the elicitation had gone wrong; "static" was not taken to mean "static".**
`NCCS-SSP-scenarios/Scripts/Preparation/Transition_identification.R` deliberately retains
`Static → Shrubland` and `Static → Static`, and — notably — re-adds them *after* the
inclusion-threshold subset, so they bypass the frequency filter entirely:

```r
# remove transtions from static where the final classes are not shrubland or static
static_trans <- x[x$Initial_class == "Static" &
                    x$Final_class %in% c("Shrubland", "Static"), ]
...
x <- x[x$Rate * 100 >= Inclusion_thres, ]      # threshold applied
x <- x[!(x$Initial_class == "Static"), ]       # all other static-initial dropped
x <- rbind(x, urban_trans, static_trans, CF_trans)   # "SSP relevant" ones added back
```

The same block hand-picks `Urban → Int_AG`, `Urban → Shrubland` and
`Closed_Forest → {Shrubland, Static, Grassland}` for the same treatment. A related decision is
visible in `lulcc.listbylulc.R:18`, where an exclusion of `Static` as an initial class is present
but **commented out**.

SSP3's elicited 2060 target asks it to *shrink* (757,680 → 734,012 cells), and SSP0/SSP4 ask it to
grow by less than its inflow delivers. The reachability precheck in `07` returns a ratio of `Inf`
for those, and the LP misses target by 23,000–41,000 cells with another class forced to absorb the
discrepancy under mass balance.

So the exclusion is dropped and `min_cardinality_abs` decides, exactly as it does for every other
class. That differs from the original in one respect worth stating: the original **hand-curated**
which static-initial transitions survive and exempted them from its own threshold, whereas here
the same rule is applied uniformly. If a static-initial transition is too rare to model, it should
fail the threshold on its merits rather than be admitted by name.

> **This is a problem with the elicitation, not with the model.**
>
> `static` is a catch-all for infrastructure, water bodies, rock and scree — see the class
> definition in `01-ingest-lulc-data.qmd`. Some members genuinely convert (rock and scree revegetate
> to shrubland; that is the `Static → Shrubland` edge the original kept), but most do not: lakes
> and motorways do not become farmland at a rate anyone elicited.
>
> The area targets in `NCCS_simulation_LULC_areas.xlsx` nonetheless assign `static` a 2060 value
> per scenario, as though it were a land-use class whose extent is a policy outcome. It is not —
> it is an aggregation artefact, and a target for it is really an unstated bundle of assumptions
> about deglaciation, reservoir construction and sealing. That the original's own model could not
> deliver SSP3's static target is a symptom: **the elicitation asked for something the class
> structure cannot express.**
>
> Retaining it here is a deliberate choice to stay close to the original for the replication,
> **not** an endorsement. Two things would fix it properly, both out of scope,
> tracked in `TODO.md`: disaggregating `static` into its convertible and non-convertible members
> (which the deglaciation TODO already requires), and re-eliciting demand against classes that can
> actually carry a target. Until then, treat any reported `static` trajectory as an artefact and do
> not interpret it.

```{r}
#| label: viable-trans-identification
library(evoland)
db <- evoland_db$new(path = "ssp-ch.evolanddb")

lulc_meta <- db$lulc_meta_t

# commit with is_viable flag set according to criterion derived from the 04d graph
db$trans_meta_t <-
  create_trans_meta_t(
    db$trans_v,
    # min_frequency_rel = 0.005, # this is what ben used: the change rate relative
    # to the other change rates
    min_cardinality_abs = 10000, # TODO decrease until clean trans rate solution is possible
    exclude_anterior = lulc_meta[name == "glacier", id_lulc]
  )
```

Which static-initial transitions actually cleared the threshold? Worth looking at directly:
`static → shrubland` is not the only survivor; doesn't match the original's curated set by the 1k minimum threshold.

```{r}
#| label: viable-trans-static-check
static_id <- lulc_meta[name == "static", id_lulc]

db$trans_meta_t[id_lulc_anterior == static_id][
  lulc_meta[, .(id_lulc, posterior = name)],
  on = .(id_lulc_posterior = id_lulc),
  nomatch = NULL
][order(-cardinality)][, .(posterior, cardinality, is_viable)]
```
