---
title: "05 — Covariate selection"
---

Prune the predictor set per viable transition using random forest
importance, and commit the survivors to `trans_preds_t`. That table is what `06` trains on, and
what `02-ingest-preds-ch2025-3-gwl.qmd` reads to decide which climate predictors are worth
projecting.

## Rationale

Using Gini based importance ordering for now; we should really have a look at each transition in detail for a better heuristic on what variables to choose. Might be good to first do an ordering using GRRF or random forest, then FilterFindCorrelation for the continuous variables (cannot handle categorical ones like that).

GRRF should handle redundancy: the first (unregularised) forest splits importance across correlated
predictors, and the regularisation penalty then suppresses the weaker ones, see if that actually
carries a stronger signal.

## Score

```{r}
#| label: covariate-selection-params
library(data.table)
library(evoland)
library(ggplot2)

db <- evoland_db$new(path = "ssp-ch.evolanddb")

db$set_full_trans_preds()

grrf_gamma <- 0.5
grrf_num_trees <- 100L
grrf_max_depth <- 20L

grrf_learner <- LearnerClassifGrrf$new()
grrf_learner$param_set$values <- list(
  gamma = grrf_gamma,
  num.trees = grrf_num_trees,
  max.depth = grrf_max_depth
)
#
# ranger_lrn <- mlr3::lrn(
#   "classif.ranger",
#   num.trees = 100,
#   max.depth = 20L,
#   importance = "impurity"
# )

cluster <- mirai::make_cluster(n = 21) # roughly ceiling(n_viable/2)

imp_scored <- db$get_pred_filter_score(
  filter = mlr3filters::FilterImportance$new(learner = grrf_learner),
  cluster = cluster,
  ordered_pred_data = TRUE # deterministic across re-runs,
)

mirai::stop_cluster(cluster)

imp_scored[, importance_rel := importance / max(importance, na.rm = TRUE), by = id_trans]
```

## Visualize

Check out the distribution of importance scores. Does not look like 

```{r}
#| label: covsel-diagnostic-plot
#| fig-width: 10
#| fig-height: 7
trans_labels <-
  db$trans_meta_t[is_viable == TRUE][
    db$lulc_meta_t[, .(id_lulc, ant = name)],
    on = .(id_lulc_anterior = id_lulc),
    nomatch = NULL
  ][
    db$lulc_meta_t[, .(id_lulc, post = name)],
    on = .(id_lulc_posterior = id_lulc),
    nomatch = NULL
  ][, .(id_trans, trans_label = paste0(ant, "→", post))]
pred_labels <- db$pred_meta_t[, .(id_pred, pred_name = name)]

plot_dt <- imp_scored[!is.na(importance_rel)][trans_labels, on = "id_trans", nomatch = NULL]

data.table::fwrite(
  plot_dt[pred_labels, on = "id_pred"],
  "2026-05-ssp-ch/05-covariate-importances-grrf.csv"
)

p <-
  ggplot(plot_dt, aes(x = reorder(trans_label, importance_rel, FUN = median), y = importance_rel)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "firebrick") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "GRRF importance per predictor, relative to the best predictor of each transition",
    subtitle = paste(
      "Dashed line: the provisional importance_rel_cut = 0.5 in 05.",
      "Look for an empty band; put the cut in it."
    ),
    x = NULL,
    y = "Relative importance"
  ) +
  coord_flip()
ggsave("2026-05-ssp-ch/importance_scores_grrf.png", p)

# TODO this shows that we might want to approach the lower transitions differently
p2 <-
  ggplot(plot_dt, aes(x = reorder(trans_label, importance, FUN = median), y = importance)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 1) +
  labs(
    title = "GRRF importance per predictor, absolute",
    x = NULL,
    y = "absolute importance"
  ) +
  coord_flip()
ggsave("2026-05-ssp-ch/importance_scores_abs_grrf.png", p2)
```

## Select

Importance is normalised **within each transition**: raw impurity importance is not comparable
across transitions with different class balance or sample size, so an absolute threshold would
silently prune hard-to-model transitions more aggressively than easy ones.

```{r}
#| label: covariate-selection-select
max_preds_per_trans <- 7
trans_preds_selected <-
  imp_scored[is.na(importance) | importance_rel >= 0.2][
    order(-importance_rel),
    head(.SD, max_preds_per_trans),
    by = id_trans
  ]

message(sprintf(
  "Kept %d of %d (transition, predictor) pairs (%.1f%%); median %.0f predictors per transition.",
  nrow(trans_preds_selected), nrow(imp_scored),
  100 * nrow(trans_preds_selected) / nrow(imp_scored),
  median(trans_preds_selected[, .N, by = id_trans]$N)
))
```

## Check and commit

Two things must hold before overwriting `trans_preds_t`: every viable transition still has at
least one predictor, and selection introduced nothing that is not viable. A transition that
survives `04` but ends up with an empty predictor set cannot be modelled in `06` —
`fit_full_models()` would leave its `learner_full` NULL, which aborts allocation later with
*"No fitted model for viable transition(s)"*.

```{r}
#| label: covariate-selection-commit
viable_trans <- db$trans_meta_t[is_viable == TRUE, id_trans]
covered_trans <- unique(trans_preds_selected$id_trans)

missing_trans <- setdiff(viable_trans, covered_trans)
if (length(missing_trans) > 0L) {
  stop(
    "Viable transition(s) left with no predictors after selection: ",
    paste(missing_trans, collapse = ", "),
    ". Lower importance_rel_cut, or demote them in 04.",
    call. = FALSE
  )
}

stopifnot(
  "selection introduced transitions that are not viable" =
    length(setdiff(covered_trans, viable_trans)) == 0L
)

db$commit(
  as_trans_preds_t(trans_preds_selected[, .(id_run, id_pred, id_trans)]),
  "trans_preds_t",
  method = "overwrite"
)

# transitions down to one or two predictors are worth a second look before 06
trans_preds_selected[, .(n_preds = .N), by = id_trans][order(n_preds)][1:10]
```

```{r}
#| label: covariate-selection-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key = , value = ,
  "covsel_importance_rel_cut", as.character(0.2),
  "covsel_max_preds_per_trans", as.character(max_preds_per_trans),
  "covsel_grrf_gamma", as.character(grrf_gamma),
  "covsel_grrf_num_trees", as.character(grrf_num_trees),
  "covsel_grrf_max_depth", as.character(grrf_max_depth)
))
```


## How many predictors survive at each candidate cut?

The threshold is only useful if it leaves every viable transition modellable. A transition
reduced to zero predictors cannot be fitted in `06`, and `fit_full_models()` would leave its
`learner_full` NULL — which aborts allocation later.

```{r}
#| label: covsel-diagnostic-table
candidate_cuts <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7)

survival <- data.table::rbindlist(lapply(candidate_cuts, function(cut) {
  kept <- plot_dt[importance_rel >= cut, .N, by = id_trans]
  data.table(
    cut = cut,
    pairs_kept = sum(kept$N),
    median_preds = as.numeric(median(kept$N)),
    min_preds = if (nrow(kept)) min(kept$N) else 0L,
    trans_with_none = data.table::uniqueN(plot_dt$id_trans) - nrow(kept)
  )
}))

survival
```

Pick the largest cut for which `trans_with_none` is 0 and `min_preds` is still comfortably above
1, then set `importance_rel_cut` in `05-covariate-selection.qmd` to it and record the reasoning
there.

```{r}
#| label: covsel-diagnostic-bimodality
# A crude bimodality read per transition: the width of the largest empty band in the
# relative-importance values. A wide gap means a threshold is well-founded for that transition.
gaps <- plot_dt[
  order(id_trans, importance_rel),
  {
    v <- sort(unique(importance_rel))
    if (length(v) < 3L) {
      .(largest_gap = NA_real_, gap_lower = NA_real_)
    } else {
      d <- diff(v)
      i <- which.max(d)
      .(largest_gap = round(d[i], 3), gap_lower = round(v[i], 3))
    }
  },
  by = id_trans
][trans_labels, on = "id_trans", nomatch = NULL]

gaps[order(-largest_gap)]
```
