---
title: "05 — Covariate selection"
---

Prune the predictor set per viable transition using guided regularised random forest (GRRF)
importance, and commit the survivors to `trans_preds_t`. That table is what `06` trains on, and
what `02-ingest-preds-ch2025-3-gwl.qmd` reads to decide which climate predictors are worth
projecting.

> **This step refuses to run until `importance_rel_cut` is set.** It ships as `NA` on purpose.
> Choose it by reading the distribution off `05d-covariate-selection.qmd` against the real
> predictor set, then set it here with a one-line note on what you saw — exactly as `04`'s
> `min_cardinality_abs` is chosen from `04d`. See "why there is no parameter-free cut" below.

## What changed

**Rebuilt on the live API.** The previous revision called
`db$get_pred_filter_score(filter_fun = grrf_filter, num.trees =, max.depth =, gamma =, cores =)`.
None of `grrf_filter`, `covariance_filter` or `get_pruned_trans_preds_t` exist in evoland-plus,
at either the pinned commit or `main`. The live API is
`get_pred_filter_score(filter = <mlr3filters::Filter>, cluster =)`, and it **scores but does not
prune** — it returns `trans_preds_t` with one extra column named after `filter$id`, and
subsetting is the caller's job. The old first chunk assigned the scored table straight back to
`db$trans_preds_t`, so it committed the full cross product unpruned.

**The correlation pre-filter is gone.** An earlier draft ran `FilterFindCorrelation` first to
drop collinear predictors. That is hard to defend: the threshold is arbitrary, which member of a
correlated pair survives is decided by feature order rather than by usefulness, and the filter is
blind to the target — it would happily drop the predictor that actually carries the signal.
GRRF already handles redundancy in a target-aware way: the first (unregularised) forest splits
importance across correlated predictors, and the regularisation penalty then suppresses the
weaker ones. One defensible mechanism beats two stacked heuristics.

*(A practical corollary, kept as a note because it will bite again: `FilterFindCorrelation`
accepts only `integer`/`numeric` features. `bioregion` / `biosubregion` are `factor`, and a task
containing them makes that filter raise "unsupported feature types: factor" — which
`pred_filter_worker()` catches, warns about, and turns into an all-`NA` score for the whole
transition. It fails as a silent hole rather than a crash. GRRF's learner supports
`factor`/`ordered` natively, so this step has no such gap.)*

## Parameters

```{r}
#| label: covariate-selection-params
library(data.table)
library(evoland)

# Keep predictors scoring at least this fraction of the best predictor FOR THAT TRANSITION.
#
# Deliberately NA: there is no defensible default (see below). Run 05d, look at the
# distribution, pick a value inside the empty band, and record here what you saw.
importance_rel_cut <- NA_real_

# GRRF guidance strength. 0 = plain regularised forest (equal penalty everywhere), 1 = penalty
# scales fully with the unregularised importance ranking. 0.9 keeps a strong guiding effect
# while leaving room for correlated-but-useful predictors to enter.
grrf_gamma <- 0.9

# Trees per forest. GRRF trains two forests per transition over ~10^5-10^6 rows each, for every
# viable transition, so this is the main cost knob. 200 is past the point where the importance
# *ranking* stabilises for a few dozen predictors, without paying ranger's 500 default twice.
grrf_num_trees <- 200L

# Depth cap, mostly a runtime guard. 0 would mean unlimited.
grrf_max_depth <- 20L

# Safety valve: never keep more than this many predictors for one transition, whatever the
# threshold says. Guards 06 against a transition that scores everything highly.
max_preds_per_trans <- 20L

# Parallel workers. regularization.factor disables ranger's internal threading (it prints
# "Parallelization deactivated"), so this cluster is the ONLY parallelism in this step.
# Each worker holds its own DB connection and a forest; memory is the binding constraint.
n_workers <- 4L

if (is.na(importance_rel_cut)) {
  stop(
    "importance_rel_cut is not set.\n",
    "  Render 05d-covariate-selection.qmd, read the per-transition importance distribution,\n",
    "  and choose a cut that sits in the empty band between signal and noise while leaving\n",
    "  every viable transition with predictors. Then set it in this chunk.\n",
    "  There is deliberately no default -- see 'why there is no parameter-free cut'.",
    call. = FALSE
  )
}

db <- evoland_db$new(path = "ssp-ch.evolanddb")
```

> **Still open (see `TODO.md`):** `grrf_gamma`, `grrf_num_trees` and `grrf_max_depth` are
> reasoned but not yet justified by a sensitivity check on this data. Re-running across a small
> grid and reporting how much the retained set moves is the outstanding work.

## Why there is no parameter-free cut

GRRF is a two-stage procedure: an unregularised forest ranks the predictors, those ranks become
per-feature split penalties, and a second regularised forest is trained under them.

It is tempting to assume the regularisation drives uninformative predictors to exactly zero
importance, giving a free `importance > 0` cut. **It does not.** Reproducing evoland's
`LearnerClassifGrrf` training path on a synthetic task with 3 informative and 13 uninformative
predictors, at `gamma` 0.5 / 0.9 / 1.0, gives only **3 exact zeros out of 16** in every case — so
`importance > 0` would retain 13 predictors, 10 of them pure noise. The regularisation suppresses
splits on weak predictors; it does not eliminate them.

What it does produce is a strongly **bimodal** distribution. On that task:

| | relative importance |
| --- | --- |
| the 3 informative predictors | 0.85 – 1.00 |
| the 10 surviving noise predictors | 0.16 – 0.20 |
| the 3 fully suppressed | 0.00 |

The gap is wide and the cut belongs in it — but its position is a property of the data, not a
constant, and a conventional "keep above 1 % of the maximum" rule keeps all 13. Hence `05d`, and
hence the hard stop above.

## Score

```{r}
#| label: covariate-selection-score
db$set_full_trans_preds(overwrite = TRUE)

grrf_learner <- LearnerClassifGrrf$new()
grrf_learner$param_set$values <- list(
  gamma = grrf_gamma,
  num.trees = grrf_num_trees,
  max.depth = grrf_max_depth
)

cluster <- if (is.null(n_workers)) NULL else mirai::make_cluster(n = n_workers)

imp_scored <- db$get_pred_filter_score(
  filter = mlr3filters::FilterImportance$new(learner = grrf_learner),
  cluster = cluster,
  ordered_pred_data = TRUE # deterministic across re-runs
)

if (!is.null(cluster)) {
  mirai::stop_cluster(cluster)
}
```

## Select

Importance is normalised **within each transition**: raw impurity importance is not comparable
across transitions with different class balance or sample size, so an absolute threshold would
silently prune hard-to-model transitions more aggressively than easy ones.

```{r}
#| label: covariate-selection-select
imp_scored[, importance_rel := importance / max(importance, na.rm = TRUE), by = id_trans]

trans_preds_selected <-
  imp_scored[is.na(importance) | importance_rel >= importance_rel_cut][
    order(-importance_rel),
    head(.SD, max_preds_per_trans),
    by = id_trans
  ]

message(sprintf(
  "Kept %d of %d (transition, predictor) pairs (%.1f%%); median %.0f predictors per transition.",
  nrow(trans_preds_selected), nrow(imp_scored),
  100 * nrow(trans_preds_selected) / nrow(imp_scored),
  median(trans_preds_selected[, .N, by = id_trans]$N)
))
```

## Check and commit

Two things must hold before overwriting `trans_preds_t`: every viable transition still has at
least one predictor, and selection introduced nothing that is not viable. A transition that
survives `04` but ends up with an empty predictor set cannot be modelled in `06` —
`fit_full_models()` would leave its `learner_full` NULL, which aborts allocation later with
*"No fitted model for viable transition(s)"*.

```{r}
#| label: covariate-selection-commit
viable_trans <- db$trans_meta_t[is_viable == TRUE, id_trans]
covered_trans <- unique(trans_preds_selected$id_trans)

missing_trans <- setdiff(viable_trans, covered_trans)
if (length(missing_trans) > 0L) {
  stop(
    "Viable transition(s) left with no predictors after selection: ",
    paste(missing_trans, collapse = ", "),
    ". Lower importance_rel_cut, or demote them in 04.",
    call. = FALSE
  )
}

stopifnot(
  "selection introduced transitions that are not viable" =
    length(setdiff(covered_trans, viable_trans)) == 0L
)

db$commit(
  as_trans_preds_t(trans_preds_selected[, .(id_run, id_pred, id_trans)]),
  "trans_preds_t",
  method = "overwrite"
)

# transitions down to one or two predictors are worth a second look before 06
trans_preds_selected[, .(n_preds = .N), by = id_trans][order(n_preds)][1:10]
```

```{r}
#| label: covariate-selection-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key =                          , value =,
  "covsel_importance_rel_cut",     as.character(importance_rel_cut),
  "covsel_max_preds_per_trans",    as.character(max_preds_per_trans),
  "covsel_grrf_gamma",             as.character(grrf_gamma),
  "covsel_grrf_num_trees",         as.character(grrf_num_trees),
  "covsel_grrf_max_depth",         as.character(grrf_max_depth),
  "covsel_pairs_retained",         as.character(nrow(trans_preds_selected))
))
```
