---
title: "05 — Covariate selection"
---

Prune the predictor set per viable transition, in two stages: drop **redundant** predictors
(pairwise correlation), then drop **uninformative** ones (guided regularised random forest
importance). The survivors are committed to `trans_preds_t`, which is what `06` trains on and
what `02-ingest-preds-ch2025-3-gwl.qmd` reads to decide which climate predictors are worth
projecting.

## What changed, and why the previous version could not run

The earlier revision called `db$get_pred_filter_score(filter_fun = grrf_filter, num.trees =,
max.depth =, gamma =, cores =)`. None of `grrf_filter`, `covariance_filter` or
`get_pruned_trans_preds_t` exist in evoland-plus — they are absent from both the pinned commit
and `main`. The live API is `get_pred_filter_score(filter = <mlr3filters::Filter>, cluster =)`,
which **scores but does not prune**: it returns `trans_preds_t` with one extra column named
after `filter$id`, and subsetting it is the caller's job. The old first chunk assigned the
scored table straight back to `db$trans_preds_t`, so it committed the full cross product
unpruned and the correlation stage never happened at all.

> **Factor predictors force the two-stage split.** `mlr3filters::FilterFindCorrelation` accepts
> only `integer`/`numeric` features, and `bioregion` / `biosubregion` (from
> `02-ingest-preds-bioregions.qmd`) are `factor`. A task containing them makes the filter raise
> *"unsupported feature types: factor"* — which `pred_filter_worker()` catches, warns about, and
> converts into an **all-`NA` score for that whole transition**. The failure is silent in the
> sense that it looks like a scoring gap rather than a crash. So the correlation stage runs over
> numeric predictors only; categorical ones bypass it and go straight to the importance stage,
> which does support factors.

## Parameters

```{r}
#| label: covariate-selection-params
library(data.table)
library(evoland)

# Redundancy: drop a predictor correlated above |r| with one that is kept.
# 0.7 is the conventional collinearity cutoff in species-distribution / land-use modelling
# (Dormann et al. 2013, Ecography 36:27-46) and is the value the valparish prototype used.
# FilterFindCorrelation scores each feature as ~ (1 - max |r| against retained features),
# so the score threshold is 1 - corcut.
corcut <- 0.7

# GRRF guidance strength. 0 = plain regularised forest (equal penalty everywhere), 1 = penalty
# scales fully with the unregularised importance ranking. 0.9 keeps a strong guiding effect
# while leaving a little room for correlated-but-useful predictors to enter.
grrf_gamma <- 0.9

# Trees per forest. GRRF trains two forests per transition, over ~10^5-10^6 rows each, for
# every viable transition -- so this is the main cost knob. 200 is well past the point where
# the importance *ranking* stabilises for a few dozen predictors, without paying for ranger's
# 500 default twice over.
grrf_num_trees <- 200L

# Depth cap, mostly a runtime guard. 0 would mean unlimited.
grrf_max_depth <- 20L

# Importance cut, as a fraction of the most important predictor for that transition.
# PROVISIONAL -- read this off the 05d diagnostic, exactly as 04's min_cardinality_abs is read
# off 04d. See "why there is no parameter-free cut" below for why 0.5 rather than something
# small: GRRF leaves uninformative predictors with substantial non-zero importance.
importance_rel_cut <- 0.5

# Safety valve: never keep more than this many predictors for one transition, whatever the
# threshold says. Guards 06 against a transition that scores everything highly.
max_preds_per_trans <- 20L

# Parallel workers. Each holds its own DB connection and a forest; memory, not CPU, is the
# binding constraint. NULL runs serially.
n_workers <- 4L

db <- evoland_db$new(path = "ssp-ch.evolanddb")
```

> **Open (see `TODO.md`):** `corcut` has a literature-backed default; `grrf_gamma`,
> `grrf_num_trees` and `grrf_max_depth` are reasoned but **not yet justified by a sensitivity
> check** on this data, and `importance_rel_cut` must be read off `05d` once this has run on
> the real predictor set. Re-running `05` across a small grid and reporting how much the
> retained set moves is the outstanding work.

## Stage 0 — the full transition × predictor cross product

```{r}
#| label: covariate-selection-full
db$set_full_trans_preds(overwrite = TRUE)

pred_meta <- db$pred_meta_t
numeric_preds <- pred_meta[data_type %in% c("int", "float"), id_pred]
categorical_preds <- pred_meta[data_type %in% c("factor", "ordered", "bool"), id_pred]

message(
  "Scoring ", nrow(db$trans_meta_t[is_viable == TRUE]), " viable transitions against ",
  length(numeric_preds), " numeric and ", length(categorical_preds), " categorical predictors."
)
```

## Stage 1 — redundancy prune (numeric predictors only)

```{r}
#| label: covariate-selection-correlation
cluster <- if (is.null(n_workers)) NULL else mirai::make_cluster(n = n_workers)

trans_preds_numeric <- as_trans_preds_t(db$trans_preds_t[id_pred %in% numeric_preds])

cor_scored <- db$get_pred_filter_score(
  filter = mlr3filters::flt("find_correlation"),
  trans_preds = trans_preds_numeric,
  cluster = cluster,
  ordered_pred_data = TRUE # deterministic across re-runs
)

# Higher score = less redundant. A predictor scoring below (1 - corcut) is collinear with
# something already kept. NA means the filter could not score it; keep those rather than
# silently discarding a predictor because of a scoring failure.
cor_keep <- cor_scored[is.na(find_correlation) | find_correlation > (1 - corcut)]

message(sprintf(
  "Correlation stage: kept %d of %d numeric (transition, predictor) pairs (%.1f%%).",
  nrow(cor_keep), nrow(cor_scored), 100 * nrow(cor_keep) / nrow(cor_scored)
))

# which predictors are most often dropped as redundant? worth an eyeball
cor_scored[
  !is.na(find_correlation) & find_correlation <= (1 - corcut),
  .N,
  by = id_pred
][
  pred_meta[, .(id_pred, name)],
  on = "id_pred",
  nomatch = NULL
][order(-N)][1:15]
```

## Stage 2 — importance (survivors plus all categoricals)

GRRF is a two-stage procedure: an unregularised forest ranks the predictors, those ranks become
per-feature split penalties, and a second regularised forest is trained under them.

### Why there is no parameter-free cut

It is tempting to assume the regularisation drives uninformative predictors to exactly zero
importance, giving a free `importance > 0` cut. **It does not.** Reproducing evoland's
`LearnerClassifGrrf` training path on a synthetic task with 3 informative and 13 uninformative
predictors, at `gamma` 0.5 / 0.9 / 1.0, gives only **3 exact zeros out of 16** in every case —
so `importance > 0` would retain 13 predictors, 10 of them pure noise. The regularisation
suppresses splits on weak predictors; it does not eliminate them.

What it does produce is a **strongly bimodal** distribution. On that task:

| | relative importance |
| --- | --- |
| the 3 informative predictors | 0.85 – 1.00 |
| the 10 surviving noise predictors | 0.16 – 0.20 |
| the 3 fully suppressed | 0.00 |

The gap is wide and the cut belongs in it — but its position is a property of the data, not a
constant. A conventional "keep above 1 % of the maximum" threshold keeps all 13. So the cut is
chosen the same way `04`'s `min_cardinality_abs` is: **by reading the distribution off a
diagnostic**, here `05d-covariate-selection.qmd`.

> **Runtime note.** `ranger` prints *"Parallelization deactivated (regularization used)"* —
> setting `regularization.factor` disables its internal threading. The per-transition `mirai`
> cluster is therefore the only parallelism available in this step, which makes `n_workers`
> matter more than it would otherwise.

```{r}
#| label: covariate-selection-importance
grrf_learner <- LearnerClassifGrrf$new()
grrf_learner$param_set$values <- list(
  gamma = grrf_gamma,
  num.trees = grrf_num_trees,
  max.depth = grrf_max_depth
)

# categorical predictors skipped stage 1 (see the note above), so add them back here
trans_preds_stage2 <- as_trans_preds_t(rbind(
  cor_keep[, .(id_run, id_pred, id_trans)],
  db$trans_preds_t[id_pred %in% categorical_preds, .(id_run, id_pred, id_trans)]
))

imp_scored <- db$get_pred_filter_score(
  filter = mlr3filters::FilterImportance$new(learner = grrf_learner),
  trans_preds = trans_preds_stage2,
  cluster = cluster,
  ordered_pred_data = TRUE
)

if (!is.null(cluster)) {
  mirai::stop_cluster(cluster)
}

# Relative to the best predictor *for that transition*, since impurity importance is not
# comparable across transitions with different class balance or sample size.
imp_scored[, importance_rel := importance / max(importance, na.rm = TRUE), by = id_trans]

trans_preds_selected <-
  imp_scored[is.na(importance) | importance_rel >= importance_rel_cut][
    order(-importance_rel),
    head(.SD, max_preds_per_trans),
    by = id_trans
  ]

message(sprintf(
  "Importance stage: kept %d of %d (transition, predictor) pairs (%.1f%%).",
  nrow(trans_preds_selected), nrow(imp_scored),
  100 * nrow(trans_preds_selected) / nrow(imp_scored)
))
```

## Check and commit

Two things must hold before overwriting `trans_preds_t`: every viable transition still has at
least one predictor, and no viable transition was lost along the way. A transition that survives
`04` but ends up with an empty predictor set cannot be modelled in `06`, and `fit_full_models()`
would leave its `learner_full` NULL — which then aborts allocation.

```{r}
#| label: covariate-selection-commit
viable_trans <- db$trans_meta_t[is_viable == TRUE, id_trans]
covered_trans <- unique(trans_preds_selected$id_trans)

missing_trans <- setdiff(viable_trans, covered_trans)
if (length(missing_trans) > 0L) {
  stop(
    "Viable transition(s) left with no predictors after selection: ",
    paste(missing_trans, collapse = ", "),
    ". Loosen corcut or the importance cut, or demote them in 04."
  )
}

stopifnot(
  "selection introduced transitions that are not viable" =
    length(setdiff(covered_trans, viable_trans)) == 0L
)

db$commit(
  as_trans_preds_t(trans_preds_selected[, .(id_run, id_pred, id_trans)]),
  "trans_preds_t",
  method = "overwrite"
)

# how many predictors survived per transition? a transition down to 1-2 is worth a look
trans_preds_selected[, .(n_preds = .N), by = id_trans][order(n_preds)][1:10]
```

```{r}
#| label: covariate-selection-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key =                        , value =,
  "covsel_corcut",               as.character(corcut),
  "covsel_grrf_gamma",           as.character(grrf_gamma),
  "covsel_grrf_num_trees",       as.character(grrf_num_trees),
  "covsel_grrf_max_depth",       as.character(grrf_max_depth),
  "covsel_importance_rel_cut",   as.character(importance_rel_cut),
  "covsel_max_preds_per_trans",  as.character(max_preds_per_trans),
  "covsel_pairs_retained",       as.character(nrow(trans_preds_selected))
))
```
