---
title: "05d — Covariate selection: importance distribution diagnostic"
---

Diagnostic for step `05`. Plots the distribution of GRRF importance per transition so the
`importance_rel_cut` committed in `05-covariate-selection.qmd` can be **read off**, the same way
`04`'s `min_cardinality_abs` is read off `04d`. Read-only: scores predictors but writes nothing
back to the DB.

## Why this diagnostic exists

The obvious cut — keep predictors whose GRRF importance is above zero — does not work.
Reproducing evoland's `LearnerClassifGrrf` training path on a synthetic task with 3 informative
and 13 uninformative predictors produced only **3 exact zeros out of 16**, at every `gamma`
tried (0.5 / 0.9 / 1.0). Regularisation suppresses splits on weak predictors without eliminating
them, so `importance > 0` retained 10 pure-noise predictors.

What it does produce is a strongly **bimodal** distribution — informative predictors at 0.85–1.00
relative importance, surviving noise clustered tightly around 0.16–0.20. The cut belongs in that
gap, but the gap's position is a property of the data, so it has to be looked at rather than
assumed. If the real distribution turns out *not* to be bimodal, that is itself the finding: it
would mean the predictor set carries no clean signal/noise separation for that transition, and
a top-`k` rule would be more honest than a threshold.

```{r}
#| label: covsel-diagnostic-score
#| fig-width: 10
#| fig-height: 7
library(data.table)
library(evoland)
library(ggplot2)

db <- evoland_db$new(path = "ssp-ch.evolanddb")

# Mirror 05's parameters so the diagnostic describes the same model it is calibrating.
grrf_gamma <- 0.9
grrf_num_trees <- 200L
grrf_max_depth <- 20L
n_workers <- 4L

grrf_learner <- LearnerClassifGrrf$new()
grrf_learner$param_set$values <- list(
  gamma = grrf_gamma,
  num.trees = grrf_num_trees,
  max.depth = grrf_max_depth
)

cluster <- if (is.null(n_workers)) NULL else mirai::make_cluster(n = n_workers)

# Score the FULL cross product, unpruned -- the point is to see the whole distribution,
# including what stage 1 would have removed.
db$set_full_trans_preds(overwrite = TRUE)

imp_scored <- db$get_pred_filter_score(
  filter = mlr3filters::FilterImportance$new(learner = grrf_learner),
  cluster = cluster,
  ordered_pred_data = TRUE
)

if (!is.null(cluster)) {
  mirai::stop_cluster(cluster)
}

imp_scored[, importance_rel := importance / max(importance, na.rm = TRUE), by = id_trans]
```

## Where does the cut belong?

Each point is one (transition, predictor) pair, ranked within its transition. A bimodal column —
a dense cluster low down, a few points high up — means a threshold is defensible, and it belongs
in the empty band between them. A smooth column means it is not.

```{r}
#| label: covsel-diagnostic-plot
#| fig-width: 10
#| fig-height: 7
trans_labels <-
  db$trans_meta_t[is_viable == TRUE][
    db$lulc_meta_t[, .(id_lulc, ant = name)],
    on = .(id_lulc_anterior = id_lulc),
    nomatch = NULL
  ][
    db$lulc_meta_t[, .(id_lulc, post = name)],
    on = .(id_lulc_posterior = id_lulc),
    nomatch = NULL
  ][, .(id_trans, trans_label = paste0(ant, "→", post))]

plot_dt <- imp_scored[!is.na(importance_rel)][trans_labels, on = "id_trans", nomatch = NULL]

ggplot(plot_dt, aes(x = reorder(trans_label, importance_rel, FUN = median), y = importance_rel)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "firebrick") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "GRRF importance per predictor, relative to the best predictor of each transition",
    subtitle = paste(
      "Dashed line: the provisional importance_rel_cut = 0.5 in 05.",
      "Look for an empty band; put the cut in it."
    ),
    x = NULL,
    y = "Relative importance"
  ) +
  coord_flip()
```

## How many predictors survive at each candidate cut?

The threshold is only useful if it leaves every viable transition modellable. A transition
reduced to zero predictors cannot be fitted in `06`, and `fit_full_models()` would leave its
`learner_full` NULL — which aborts allocation later.

```{r}
#| label: covsel-diagnostic-table
candidate_cuts <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7)

survival <- data.table::rbindlist(lapply(candidate_cuts, function(cut) {
  kept <- plot_dt[importance_rel >= cut, .N, by = id_trans]
  data.table(
    cut = cut,
    pairs_kept = sum(kept$N),
    median_preds = as.numeric(median(kept$N)),
    min_preds = if (nrow(kept)) min(kept$N) else 0L,
    trans_with_none = data.table::uniqueN(plot_dt$id_trans) - nrow(kept)
  )
}))

survival
```

Pick the largest cut for which `trans_with_none` is 0 and `min_preds` is still comfortably above
1, then set `importance_rel_cut` in `05-covariate-selection.qmd` to it and record the reasoning
there.

```{r}
#| label: covsel-diagnostic-bimodality
# A crude bimodality read per transition: the width of the largest empty band in the
# relative-importance values. A wide gap means a threshold is well-founded for that transition.
gaps <- plot_dt[
  order(id_trans, importance_rel),
  {
    v <- sort(unique(importance_rel))
    if (length(v) < 3L) {
      .(largest_gap = NA_real_, gap_lower = NA_real_)
    } else {
      d <- diff(v)
      i <- which.max(d)
      .(largest_gap = round(d[i], 3), gap_lower = round(v[i], 3))
    }
  },
  by = id_trans
][trans_labels, on = "id_trans", nomatch = NULL]

gaps[order(-largest_gap)]
```
