---
title: "06 — Transition models"
---

Fit one transition-potential model per viable transition: a binary classifier answering "given a
cell in the anterior class, did it transition?", trained on the predictors `05` retained. The
fitted models go to `trans_models_t` and are what `predict_trans_pot()` calls at allocation time.

Two passes, following the evoland-plus workflow:

1. **Partial models** — trained on a stratified subsample, scored on the held-out remainder.
   These produce the goodness-of-fit numbers and the held-out predictions that `06d` plots.
2. **Full models** — the best partial model per transition is refitted on all available data.
   `predict_trans_pot()` uses these.

## Learner and selection criterion

A plain **random forest** (`classif.ranger`), selected on **AUC**. Deliberately unambitious for
a first pass:

- The valparish prototype used a GLM; a forest relaxes the linearity and additivity assumptions
  that a GLM imposes on transition potentials, which is the more obviously wrong constraint
  here — suitability surfaces are not monotone in elevation or distance.
- It needs no tuning to be usable, and `05` has already done the feature selection a forest
  would otherwise have to absorb.
- It handles the `factor` predictors (`bioregion`) natively, without contrast coding.

AUC as the criterion is the natural fit for what allocation actually consumes. `predict_trans_pot()`
stores per-cell probabilities, and `adjusted_trans_pot_v()` then column-scales them so the mean
over the source class matches the target rate — a monotone transformation. What matters is
therefore the *ranking* of cells by transition propensity, not the calibration of the absolute
probabilities, and AUC scores exactly that ranking. It is also insensitive to class imbalance,
which matters when a rare transition has a positive rate below 1 %.

We also fit a **featureless baseline** first. It costs nothing and makes the forest's AUC
interpretable: a transition where the forest barely beats 0.5 has no usable spatial signal, and
is better caught here than after allocation produces noise.

> **Open (see `TODO.md`):** `sample_frac = 0.7` is evoland's default and what both package
> vignettes use, so it is a convention rather than a justified choice. A learning-curve check —
> AUC against subsample size for a few representative transitions — would settle it, and would
> also say whether the rarer transitions are sample-starved.

```{r}
#| label: transition-modelling-params
library(data.table)
library(evoland)
library(mlr3learners)

sample_frac <- 0.7 # stratified train fraction; the remainder is the held-out scoring split
seed <- 42L
num_trees <- 100L
max_depth <- 20L
n_workers <- 21L

db <- evoland_db$new(path = "ssp-ch.evolanddb")

stopifnot(
  "trans_preds_t is empty - run 05-covariate-selection.qmd first" =
    nrow(db$trans_preds_t) > 0L
)
```

## Partial models

`fit_partial_models()` coerces the learner's `predict_type` to `"prob"` — AUC needs
probabilities, not hard class labels — and stores the held-out `PredictionClassif` per
transition, which is what makes `06d` possible.

```{r}
#| label: transition-modelling-partial
cluster <- if (is.null(n_workers)) NULL else mirai::make_cluster(n = n_workers)

# too naive
# db$trans_models_t <- db$fit_partial_models(
#   learner = mlr3::lrn("classif.rpart", predict_type = "prob"),
#   measures = c("classif.auc", "classif.acc"),
#   sample_frac = sample_frac,
#   seed = seed,
#   cluster = cluster
# )

db$trans_models_t <- db$fit_partial_models(
  learner = mlr3::lrn("classif.naive_bayes", predict_type = "prob"),
  measures = c("classif.auc", "classif.acc"),
  sample_frac = sample_frac,
  seed = seed,
  cluster = cluster
)

db$trans_models_t <- db$fit_partial_models(
  learner = mlr3::lrn("classif.log_reg", predict_type = "prob"),
  measures = c("classif.auc", "classif.acc"),
  sample_frac = sample_frac,
  seed = seed,
  cluster = cluster
)

db$trans_models_t <- db$fit_partial_models(
  learner = mlr3::lrn(
    "classif.ranger",
    predict_type = "prob",
    num.trees = num_trees,
    max.depth = max_depth
  ),
  measures = c("classif.auc", "classif.acc"),
  sample_frac = sample_frac,
  seed = seed,
  cluster = cluster
)

if (!is.null(cluster)) {
  mirai::stop_cluster(cluster)
}
```

## Does the forest beat the baseline?

A forest that cannot separate transitioning from non-transitioning cells will still produce a
potential surface — an uninformative one. Allocation would then place change essentially at
random within the anterior class, constrained only by the demand and the patch parameters. That
is worth catching here rather than inferring it from an odd-looking map in `09d`.

```{r}
#| label: transition-modelling-compare
# crossval_score is a MAP column: one named list per row, e.g. list(classif.auc = 0.83, ...),
# not a long-format (measure_id, score) pair.
partial <- db$trans_models_t
partial[, auc := vapply(
  crossval_score,
  function(s) if (is.null(s[["classif.auc"]])) NA_real_ else as.numeric(s[["classif.auc"]]),
  numeric(1L)
)]

# max() over an empty group returns -Inf with a warning, which would then be silently
# compared against the 0.6 threshold below; collapse missing combinations to NA instead.
max_or_na <- function(z) if (length(z)) max(z) else NA_real_

auc_by_learner <-
  partial[!is.na(auc) & learner_id != "error"] |>
  data.table::dcast(id_trans ~ learner_id, value.var = "auc", fun.aggregate = max_or_na)

auc_cols <- setdiff(names(auc_by_learner), "id_trans")
ranger_col <- grep("ranger", auc_cols, value = TRUE)
featureless_col <- grep("featureless", auc_cols, value = TRUE)

if (length(ranger_col) == 1L && length(featureless_col) == 1L) {
  auc_by_learner[, lift := get(ranger_col) - get(featureless_col)]

  # NA here means the forest produced no scoreable model at all, which is a harder failure
  # than a weak one -- separate the two so the message is actionable.
  failed <- auc_by_learner[is.na(get(ranger_col)), id_trans]
  weak <- auc_by_learner[!is.na(get(ranger_col)) & get(ranger_col) < 0.6, id_trans]

  if (length(failed) > 0L) {
    warning(
      "Transition(s) with no scoreable ranger model - they will be demoted below: ",
      paste(failed, collapse = ", ")
    )
  }
  if (length(weak) > 0L) {
    warning(
      "Transition(s) with ranger AUC < 0.6 - their potential surfaces carry little signal: ",
      paste(weak, collapse = ", ")
    )
  }
}

auc_by_learner[order(get(ranger_col))]
```

## Full models

`fit_full_models(select_score = "classif.auc", select_maximize = TRUE)` picks the best-scoring
partial model per transition and refits it on all available data.

```{r}
#| label: transition-modelling-full
cluster <- if (is.null(n_workers)) NULL else mirai::make_cluster(n = n_workers)
trans_models_full <- db$fit_full_models(
  select_score = "classif.auc",
  select_maximize = TRUE,
  cluster = cluster
)

if (!is.null(cluster)) {
  mirai::stop_cluster(cluster)
}

db$trans_models_t <- trans_models_full
```

## Reconcile viability with what actually trained

Fitting does not necessarily produce a usable model for every viable transition — a rare
transition can fail to train, leaving `learner_full` NULL. Such a transition would stay
`is_viable == TRUE` with no model, and allocation aborts with *"No fitted model for viable
transition(s)"*. We keep the two in lockstep by demoting any transition whose full model did not
train, reading the result returned by `fit_full_models()` directly rather than re-querying the
DB (per the evoland-plus stochastic-allocation vignette).

```{r}
#| label: transition-modelling-reconcile
modeled_trans <- unique(
  trans_models_full$id_trans[
    !vapply(trans_models_full$learner_full, is.null, logical(1L))
  ]
)

trans_meta_reconciled <- db$trans_meta_t
demoted <- trans_meta_reconciled[is_viable == TRUE & !id_trans %in% modeled_trans, id_trans]

if (length(demoted) > 0L) {
  warning(
    "Demoting viable transition(s) whose full model did not train: ",
    paste(demoted, collapse = ", ")
  )
}

trans_meta_reconciled[, is_viable := is_viable & id_trans %in% modeled_trans]
db$trans_meta_t <- trans_meta_reconciled

message(
  "Modelled ", length(modeled_trans), " transitions; ",
  nrow(db$trans_meta_t[is_viable == TRUE]), " remain viable."
)
```

```{r}
#| label: transition-modelling-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key = , value = ,
  "model_learner", "classif.ranger",
  "model_select_score", "classif.auc (maximise)",
  "model_sample_frac", as.character(sample_frac),
  "model_num_trees", as.character(num_trees),
  "model_max_depth", as.character(max_depth),
  "model_seed", as.character(seed)
))
```
