---
title: "06d — Transition models: ROC diagnostic"
---

Diagnostic for step `06`. Plots a **ROC curve per transition** from the held-out predictions
stored by `fit_partial_models()`, plus the AUC ranking that `fit_full_models()` selected on.
Read-only: reads `trans_models_t`, writes nothing back.

The curves come from the **held-out split**, not the training data, so they are an honest read of
each transition model's discrimination. What to look for:

- **A curve hugging the diagonal** — the model cannot separate transitioning from
  non-transitioning cells. Its potential surface is close to uninformative, and allocation will
  place that transition's change nearly at random within the anterior class, constrained only by
  the demand and the patch parameters. Either the predictor set is missing the relevant driver,
  or the transition is genuinely unpredictable at hectare resolution.
- **A curve that rises steeply then flattens** — the model finds a small set of cells it is
  confident about and is uninformative beyond them. Fine for allocation, which only needs the
  ranking at the top of the distribution, since only as many cells as the demand requires are
  ever converted.
- **A near-perfect curve on a rare transition** — treat with suspicion. Check for a predictor
  that encodes the outcome (a `static`-derived layer, or a neighbourhood predictor built from
  the posterior period).

> **Why not `db$get_crossval_plots()`?** It exists and deserialises the same predictions, but
> calls `mlr3viz::autoplot(prediction)` without a `type`, which yields the default
> class-distribution bar plot rather than a ROC. It takes no `type` argument to pass through, so
> this step deserialises directly.

```{r}
#| label: roc-diagnostic-load
library(data.table)
library(evoland)
library(ggplot2)

db <- evoland_db$new(path = "ssp-ch.evolanddb")

models <- db$fetch(
  "trans_models_t",
  cols = c("id_run", "id_trans", "learner_id", "crossval_score", "crossval_predictions")
)

stopifnot("no models with held-out predictions - run 06 first" = nrow(models) > 0L)

# label transitions by class names rather than ids, so the panels are readable
trans_labels <-
  db$trans_meta_t[
    db$lulc_meta_t[, .(id_lulc, ant = name)],
    on = .(id_lulc_anterior = id_lulc),
    nomatch = NULL
  ][
    db$lulc_meta_t[, .(id_lulc, post = name)],
    on = .(id_lulc_posterior = id_lulc),
    nomatch = NULL
  ][, .(id_trans, trans_label = paste0(ant, " => ", post))]

models[, auc := vapply(
  crossval_score,
  function(s) if (is.null(s[["classif.auc"]])) NA_real_ else as.numeric(s[["classif.auc"]]),
  numeric(1L)
)]
```

## ROC curves

One panel per transition, one line per learner, so the forest can be read against the
featureless baseline directly — a forest whose curve sits on top of the baseline's has found
nothing.

```{r}
#| label: roc-diagnostic-curves
#| fig-width: 11
#| fig-height: 9
# Deserialise each held-out PredictionClassif and reduce it to (fpr, tpr) points. Doing the
# arithmetic here rather than via mlr3viz::autoplot(type = "roc") keeps every transition in one
# faceted figure instead of a list of separate plots.
roc_points <- data.table::rbindlist(lapply(seq_len(nrow(models)), function(i) {
  blob <- models$crossval_predictions[[i]]
  if (is.null(blob)) {
    return(NULL)
  }
  pred <- qs2::qs_deserialize(blob)

  # evoland always builds the task with positive = "TRUE" and levels c("FALSE", "TRUE")
  # (see fit_partial_models / fit_full_models in trans_models_t.R).
  positive <- if ("TRUE" %in% colnames(pred$prob)) "TRUE" else colnames(pred$prob)[2L]
  prob <- pred$prob[, positive]
  truth <- pred$truth == positive

  ord <- order(prob, decreasing = TRUE)
  tp <- cumsum(truth[ord])
  fp <- cumsum(!truth[ord])
  n_pos <- sum(truth)
  n_neg <- sum(!truth)
  if (n_pos == 0L || n_neg == 0L) {
    return(NULL) # a split with one class only has no ROC
  }

  # thin to at most 500 points per curve; these predictions run to ~10^5-10^6 rows
  keep <- unique(round(seq(1, length(tp), length.out = min(500L, length(tp)))))

  data.table(
    id_trans = models$id_trans[i],
    learner_id = models$learner_id[i],
    fpr = c(0, fp[keep] / n_neg, 1),
    tpr = c(0, tp[keep] / n_pos, 1)
  )
}))

roc_points <- roc_points[trans_labels, on = "id_trans", nomatch = NULL]

p <-
  ggplot(
    roc_points[learner_id != "classif.featureless"],
    aes(x = fpr, y = tpr, colour = learner_id)
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey50") +
  geom_line(linewidth = 0.6) +
  facet_wrap(~trans_label) +
  # coord_equal() +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(
    title = "Held-out ROC per transition",
    subtitle = "Dotted diagonal = no discrimination capacity",
    x = "False positive rate",
    y = "True positive rate",
    colour = "Learner"
  ) +
  theme(legend.position = "bottom")
ggsave("2026-05-ssp-ch/06d-roc-plots.png", p, width = 11, height = 9)
```

## AUC ranking

The number `fit_full_models(select_score = "classif.auc")` actually selects on. Transitions are
ordered worst-first: the top of this plot is where the modelling is weakest and where a look at
the retained predictor set from `05` is most likely to pay off.

```{r}
#| label: roc-diagnostic-auc
#| fig-width: 9
#| fig-height: 7
auc_dt <- models[!is.na(auc) & learner_id != "error"][trans_labels, on = "id_trans", nomatch = NULL]

p2 <-
  ggplot(auc_dt, aes(x = reorder(trans_label, auc), y = auc, colour = learner_id)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "firebrick") +
  geom_point(size = 2.5) +
  coord_flip() +
  # scale_y_continuous(limits = c(0.4, 1)) +
  labs(
    title = "Held-out AUC per transition",
    subtitle = "Dashed line at 0.5: no discrimination capacity",
    x = NULL,
    y = "AUC",
    colour = "Learner"
  ) +
  theme(legend.position = "bottom")

ggsave("2026-05-ssp-ch/06d-auc-plot.png", p2, width = 9, height = 7)
```

```{r}
#| label: roc-diagnostic-table
# max() over an empty group returns -Inf with a warning; a missing (transition, learner)
# combination should read as NA.
max_or_na <- function(z) if (length(z)) max(z) else NA_real_

data.table::dcast(
  auc_dt,
  trans_label ~ learner_id,
  value.var = "auc",
  fun.aggregate = max_or_na
)[order(trans_label)]
```
