---
title: "07 — Transition rates from SSP demand"
---

Turn the per-SSP **class-area targets** elicited for the original SSP-CH project into the
**per-transition rates** that allocation consumes, using the demand-driven LP solver added to
evoland-plus for [#32](https://github.com/ethzplus/evoland-plus/issues/32).

This replaces `extrapolate_trans_rates()`, which fits one independent regression per transition
and has nowhere to state a scenario target at all. The gain is not the change of units — it is
that a *coupled* solver enforces mass balance and class-trajectory consistency across
transitions, which 21 independent univariate fits cannot.

Companion step: `07-transition-rates-2-legacy.qmd` produces a second set of rates reproducing
the original implementation's *effective* behaviour, so the two can be diffed.

## Requirements this step depends on

Verified against `evoland-plus@claude/linear-program-implementation-2wf1m7` by running it on
these exact demand numbers:

| Need | Provided by |
| --- | --- |
| bounds from observed history, irregular intervals | `trans_rate_bounds()` — annualises before min/max, re-inflates to the step length |
| "is this target even attainable?" | `trans_rate_reachability()` — pure LP, needs no targets |
| coupled solve to class-area targets | `solve_trans_rates()` |
| write `trans_rates_t` for several runs | `trans_rates_from_solution()` |
| recover the solved trajectory afterwards | `trans_rate_areas()` — replays a rate table forward |

The last one matters more than it looks. #32 asked whether the solver's predicted `area[l,t]`
should be persisted so realised-vs-predicted validation is possible after allocation. It does
not need to be: the trajectory is recoverable from `trans_rates_t` alone. Verified here — the
replay reproduces the solver's own `areas` matrix to **1.3 × 10⁻⁹ cells**.

> **Two things must be passed explicitly.** Both are defaults that do not survive contact with
> this data; see the two notes below.

## Grid portability: normalise to shares, rehydrate against our own total

The demand is stated on the original study's grid of **4,129,078 cells**. Ours is a
re-derivation from NOAS04 onto a 100 m grid restricted to coordinates carrying Arealstatistik
data, so the totals will not match, and an absolute target of "1,160,184 cells of closed forest"
is meaningless off the grid it was elicited on.

So we normalise on the source grid and rehydrate on ours. This is exact rather than approximate:
every scenario and both horizons sum to precisely 4,129,078, so the shares sum to 1 and the
rehydrated targets sum to our own observed total.

```{r}
#| label: trans-rates-setup
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")

# The anchor is the last observed period: the solver starts from what we actually observed,
# not from the spreadsheet's init_area.
anchor_period <- max(db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period])
step_periods <- db$periods_t[is_extrapolated == TRUE, id_period]

stopifnot(
  "no extrapolated periods" = length(step_periods) > 0L,
  "trans_meta_t is empty - run 04 (and 06, which reconciles viability)" =
    nrow(db$trans_meta_t[is_viable == TRUE]) > 0L
)
```

## Demand

The demand table and the share rehydration live in `R/ssp-demand.R`, sourced by this step and by
`07-transition-rates-2-legacy.qmd` so both solve against exactly the same targets. See that file
for the source, the `Int_AG` -> `arable` crosswalk, and the lowercase-shape trap.

```{r}
#| label: trans-rates-demand
source("2026-05-ssp-ch/R/ssp-demand.R")

lulc_meta <- db$lulc_meta_t
init_area <- ssp_demand_init_area(db, anchor_period)
our_total <- sum(init_area$area)

targets_by_ssp <- ssp_demand_targets(lulc_meta, our_total)
ssp_list <- sort(unique(targets_by_ssp$ssp))
```

Worth reporting before going further: where our re-derivation from NOAS04 disagrees with the
original aggregation. A large discrepancy here means the demand is being rehydrated onto a class
that does not mean quite the same thing.

```{r}
#| label: trans-rates-init-compare
comparison <- merge(
  init_area[, .(id_lulc, our_share = area / our_total)],
  ssp_demand_init_shares(lulc_meta),
  by = "id_lulc"
)[lulc_meta[, .(id_lulc, name)], on = "id_lulc", nomatch = NULL]

comparison[, pct_diff := round(100 * (our_share - their_share) / their_share, 2)]
comparison[order(-abs(pct_diff))][, .(name, our_share = round(our_share, 5),
                                      their_share = round(their_share, 5), pct_diff)]
```

## Historic rate bounds

> **`step_years` must be passed explicitly.** `trans_rate_bounds()` infers it from the
> extrapolated periods and `stopifnot`s that they are equal — but calendar decades are not
> exactly equal. On this pipeline's own `create_periods_t("P10Y", …)` output the four
> extrapolated intervals come out as 10.001369, 9.998631, 10.001369, 9.998631 years, differing
> only by leap-day placement, and the check fires with *"cannot infer step_years: extrapolated
> periods differ in length"*. Passing it explicitly is correct anyway — the nominal step is a
> modelling choice, not a calendar measurement — but the inference wants a tolerance upstream.

```{r}
#| label: trans-rates-bounds
step_years <- 10

bounds <- trans_rate_bounds(
  obs_rates = db$get_obs_trans_rates(),
  periods = db$periods_t,
  trans_meta = db$trans_meta_t,
  step_years = step_years
)

bounds[order(-max_rate)][1:12]
```

## Reachability — report, do not gate

Run *before* any solve, and treated as a result in its own right rather than a precondition.

> **`max_reachability_ratio` must be raised from its default of 10.** Exercised on these
> numbers, glacier alone asks for **11.2–12.4×** the change achievable under historic bounds, so
> the default aborts every single scenario. That is not a data error: the SSPs are normative
> visions, and "Switzerland deglaciates faster than 1985–2018 implies" is a claim about climate,
> not a violation of physics. The gate is kept only to catch the absurd.

```{r}
#| label: trans-rates-reachability
# High enough to let the glacier trajectory through, low enough to still catch a target that is
# outright impossible (e.g. asking a class with no outbound edges to shrink -> ratio Inf).
max_reachability_ratio <- 50

reachability <- trans_rate_reachability(
  init_area = init_area,
  bounds = bounds,
  n_steps = length(step_periods)
)

reachability[lulc_meta[, .(id_lulc, name)], on = "id_lulc", nomatch = NULL][
  order(-ratio)
]
```

```{r}
#| label: trans-rates-reachability-per-ssp
# How far outside the historic envelope does each scenario ask each class to go?
reach_by_ssp <- targets_by_ssp[
  reachability[, .(id_lulc, area_init, area_min, area_max)],
  on = "id_lulc",
  nomatch = NULL
][, `:=`(
  asked = area - area_init,
  achievable = data.table::fifelse(area > area_init, area_max - area_init, area_min - area_init)
)][, ratio := data.table::fifelse(
  abs(achievable) < 1e-9,
  data.table::fifelse(abs(asked) < 1e-9, 0, Inf),
  abs(asked / achievable)
)]

reach_by_ssp[lulc_meta[, .(id_lulc, name)], on = "id_lulc", nomatch = NULL][
  ratio > 1
][order(-ratio)][, .(ssp, name, asked = round(asked), achievable = round(achievable),
                     ratio = round(ratio, 2))][1:20]
```

## Solve, one LP per SSP

```{r}
#| label: trans-rates-solve
solutions <- vector("list", length(ssp_list))
names(solutions) <- ssp_list

for (s in ssp_list) {
  d <- targets_by_ssp[ssp == s]
  message("Solving ", s, " ...")
  solutions[[s]] <- solve_trans_rates(
    init_area = init_area,
    targets = d[, .(id_lulc, area)],
    shapes = d[, .(id_lulc, shape)],
    bounds = bounds,
    periods = db$periods_t,
    max_reachability_ratio = max_reachability_ratio
  )
}

solve_summary <- data.table::rbindlist(lapply(ssp_list, function(s) {
  sol <- solutions[[s]]
  fs <- sol$diagnostics$flow_summary
  data.table(
    ssp = s,
    status = sol$status,
    objective = round(sol$objective, 4),
    max_target_err = max(abs(sol$diagnostics$target_error$error), na.rm = TRUE),
    total_flow = round(fs$total_flow),
    pct_flow_above_max = round(100 * fs$flow_above_max_rate / fs$total_flow, 1),
    flow_non_viable = round(fs$flow_non_viable)
  )
}))

stopifnot("an LP did not solve to optimality" = all(solve_summary$status == 0))
solve_summary
```

### `static` is absorbing, and the demand does not know that

Expect non-zero `max_target_err` for SSP0, SSP3 and SSP4. Exercised on the real demand, those
three miss by 34,000–67,000 cells while SSP1 and SSP5 hit exactly, and the misses attribute
cleanly:

| SSP | class | asked | miss |
| --- | --- | --- | --- |
| SSP0 | static | +10,498 | **+41,319** |
| SSP0 | closed_forest | −41,938 | −41,319 |
| SSP3 | static | −23,148 | +23,148 |
| SSP3 | glacier | −41,247 | +41,247 |
| SSP3 | closed_forest | +128,946 | −66,677 |
| SSP4 | static | +7,060 | **+34,187** |

`04-viable-transition-identification.qmd` excludes `static` as an *anterior* class, so the edge
set has transitions into static but none out of it. Static can therefore only grow, and its
inflow cannot be shed — it overshoots, and under mass balance another class (here
`closed_forest`) has to undershoot by the same amount. Glacier is the same story in miniature:
its only outbound edge is bounded by a historic rate that predates the current melt.

This is a **modelling conflict, not a solver defect**, and it is worth deciding rather than
absorbing: either leave `static` as an anterior class for the transitions that genuinely occur
(sealing is not irreversible), or pre-adjust the demand so static never shrinks and redistribute
the difference. Tracked in `TODO.md`.

`pct_flow_above_max` is the number to watch. It is how much of the solved change sits outside
the historic envelope, and it propagates: `adjusted_trans_pot_v()` scales potentials by an
**unbounded** `rate / mean_value` multiplier, so when the LP demands several times the observed
rate the scaled probabilities exceed 1, the row-closure branch clips them, uPAM's quota cannot
be met and the pool empties early. The scenario then silently under-delivers its own demand.
`09d` should compare realised areas against `trans_rate_areas()` for exactly this reason.

## Commit to the SSP runs

Each solution is written to every leaf run of its SSP — the leaves differ by climate framing,
which does not change the demand.

```{r}
#| label: trans-rates-commit
runs <- db$runs_t

# 07-2 adds a `rate_solver` column and legacy sibling runs; exclude those here if present.
ssp_runs <- if ("rate_solver" %in% names(runs)) {
  runs[!is.na(ssp) & (is.na(rate_solver) | rate_solver == "solver")]
} else {
  runs[!is.na(ssp)]
}

stopifnot("no SSP leaf runs - run 00-setup-db.qmd" = nrow(ssp_runs) > 0L)

for (s in ssp_list) {
  run_ids <- ssp_runs[ssp == s, id_run]
  if (length(run_ids) == 0L) {
    warning("No runs registered for ", s, "; skipping")
    next
  }
  db$trans_rates_t <- trans_rates_from_solution(solutions[[s]], id_run = run_ids)
}

db$trans_rates_t[, .(n_rows = .N), by = id_run][order(id_run)][1:10]
```

## Verify the trajectory round-trips

The solved trajectory has to be recoverable from what we stored, or the realised-vs-predicted
comparison in `09d` is impossible.

```{r}
#| label: trans-rates-roundtrip
check_ssp <- ssp_list[1]
check_run <- ssp_runs[ssp == check_ssp, id_run][1]

replayed <- trans_rate_areas(
  init_area = init_area,
  rates = db$trans_rates_t[id_run == check_run],
  trans_meta = db$trans_meta_t
)

roundtrip <- merge(
  data.table::as.data.table(replayed),
  data.table::as.data.table(solutions[[check_ssp]]$areas),
  by = c("id_lulc", "id_period"),
  suffixes = c("_replay", "_solved")
)

max_drift <- max(abs(roundtrip$area_replay - roundtrip$area_solved))
message(sprintf("Round-trip drift for %s: %.3g cells", check_ssp, max_drift))
stopifnot("rate table does not reproduce the solved trajectory" = max_drift < 1)
```

```{r}
#| label: trans-rates-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key =                           , value =,
  "trans_rates_source",             "NCCS_simulation_LULC_areas.xlsx (class-area targets, 2060)",
  "trans_rates_method",             "solve_trans_rates() LP, shares normalised on the source grid",
  "trans_rates_step_years",         as.character(step_years),
  "trans_rates_max_reach_ratio",    as.character(max_reachability_ratio),
  "trans_rates_scenarios",          paste(ssp_list, collapse = ",")
))
```
