---
title: "07 — Transition rates: original-behaviour reproduction"
---

Produce a **second** set of transition rates reproducing how the original SSP-CH solver actually
behaved, on dedicated runs, so the corrected rates from `07-transition-rates-1-solver.qmd` can be
diffed against them.

The published SSP-CH results were produced by `lulcc.simulationtransitionratesolver.R`, and that
implementation has defects (below). Reproducing *its numbers* and reproducing *a correct solve*
are different goals, and both are wanted: the first to say "we reproduced SSP-CH", the second to
say what SSP-CH would have looked like had the solver worked as documented.

> **No buggy logic goes into evoland-plus.** The package carries only the corrected solver
> ([#32](https://github.com/ethzplus/evoland-plus/issues/32)). Bug-compatibility is a
> *replication* concern and lives here, expressed as extra `runs_t` entries. Nothing in this step
> requires a second solver implementation.

## What was wrong with the original, and what that meant

Three defects, the first two verified by reading
`NCCS-SSP-scenarios/Scripts/Functions/lulcc.simulationtransitionratesolver.R` directly:

1. **The shape mechanism is dead code.** `build_diff_row(l_i, t1, t2, ratio)` writes
   `row_[idx_area(l_i, t1)] <- 1` and then, four lines later,
   `row_[idx_area(l_i, t2 - 1)] <- ratio`. It is called as
   `build_diff_row(l_i, t_i, t_i + 1, ratio)`, so `t2 - 1 == t1` and the second assignment
   **overwrites the first**. The curvature row is not what its comments describe, `chosen_shape`
   has no effect on any output, and the term dominates the objective ~3700×, which means the LP
   was not meaningfully minimising rate-bound violations either.
2. **`Step_length` is a scalar indexed as a vector.** It is `5` at the call site
   (`Simulation_trans_tables_prep.R:14`) but read as `Step_length[t_i + 1]`, which is `NA` from
   the first iteration. `lpSolve::lp()` accepts the resulting `NA` in the constraint matrix and
   returns a different answer with `status = 0`.
3. **Forbidden edges are only expensive, not forbidden.** `r_max == 0` produces
   `x - devUpper <= 0` rather than `x = 0`, so a transition that never occurs historically can
   still carry flow if the objective prefers it.

There is also a fourth, independent reason `chosen_shape` never bit: a straight line satisfies
all five shape constraints with equality, so even with defect 1 fixed the shapes would only
matter if something else bent the trajectory.

## What can and cannot be reproduced

**Cannot:** a bit-exact replication. Defect 2 puts `NA` into the constraint matrix, and what
`lpSolve` returns then is not a defined function of the inputs. There is also nothing to check
against — `Transition_Tables.xlsx` has per-transition rates for **BAU only**; the per-SSP blocks
of `09_Modified_Trans_Rates` are empty, so the original's own SSP rate tables were never
published.

**Can:** the original's *effective* behaviour, which is well-characterised precisely because the
defects were so total. With shapes inert and smoothing at the call site's weight, the original
produced **straight-line trajectories** between the anchor and the target, with soft rate bounds
and soft forbidden edges. Every one of those is reachable through the corrected solver's own
parameters — no second implementation, and no re-introduced bugs:

| Original behaviour | Reproduced by |
| --- | --- |
| `chosen_shape` has no effect | `mu_shape = 0` |
| trajectories are straight lines | `mu_smooth = 100`, the call site's value (documented default is 1) |
| terminal handled by a hard 99–101 % band only | `terminal_band = NA`, `mu_target = 1e3` — **substituted**; see below |
| no fairness term | `fairness = FALSE` |
| no historic-preference term | `mu_historic = 0` |
| forbidden edges merely expensive | `forbid_non_viable = FALSE` — **measured only, not committed**; see below |

**The hard terminal band cannot be reproduced either, for the same reason.** The original
reached its 99–101 % band *by leaking through forbidden edges*. Forbid the leakage and the band
becomes unsatisfiable: the LP returns `lpSolve` status 2, infeasible. So the committable legacy
solve substitutes the soft terminal fit (`terminal_band = NA, mu_target = 1e3`), which lands as
close to each target as the viable edge set allows. Setting `mu_target = 0` as well would be
worse than wrong — with neither a band nor a penalty, nothing would pull toward the targets at
all and the LP would ignore the demand entirely.

**Defect 3 is quantified but not propagated, and it cannot be.** With
`forbid_non_viable = FALSE` the LP puts 240,000–420,000 cells per scenario along edges never
observed historically — that is the cost of the defect, and it is worth reporting. But
`trans_rates_from_solution()` deliberately refuses such a solution
(*"flow was allocated to non-viable transitions"*), and it is right to: those transitions have no
rows in `trans_pot_t`, so evoland would silently drop the flow at allocation anyway. Persisting
it would promise change that no allocator can deliver.

So this step solves **twice**: once unforbidden, purely to measure the leakage, and once
forbidden, to produce rates that can actually be committed. The committed legacy rates therefore
reproduce defects 1 and 2's effects — straight-line trajectories, band-only terminal handling —
but not defect 3's. That is stated here rather than buried, because it is the one respect in
which "the legacy runs" are not what the original would have produced.

The leakage figure is itself a replication finding: to that extent, the original's stated demand
was never delivered by its own allocator.

This is an **emulation of behaviour, not a port of code**, and it should be described that way in
any write-up. It answers "what does a straight-line, soft-bounds solve give?" — which is what the
original computed — without asserting that the digits match numbers nobody has.

## Setup

```{r}
#| label: legacy-setup
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")

anchor_period <- max(db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period])
step_periods <- db$periods_t[is_extrapolated == TRUE, id_period]
step_years <- 10 # see 07-1: inference fails on real calendar decades
max_reachability_ratio <- 50 # see 07-1: the default of 10 aborts every scenario

stopifnot(
  "run 07-transition-rates-1-solver.qmd first" = nrow(db$trans_rates_t) > 0L
)
```

Both steps source the same demand table from `R/ssp-demand.R`, so they solve against identical
targets and any difference in the output is attributable to the solver settings alone.

```{r}
#| label: legacy-demand
source("2026-05-ssp-ch/R/ssp-demand.R")

lulc_meta <- db$lulc_meta_t
init_area <- ssp_demand_init_area(db, anchor_period)
targets_by_ssp <- ssp_demand_targets(lulc_meta, sum(init_area$area))
ssp_list <- sort(unique(targets_by_ssp$ssp))

bounds <- trans_rate_bounds(
  obs_rates = db$get_obs_trans_rates(),
  periods = db$periods_t,
  trans_meta = db$trans_meta_t,
  step_years = step_years
)
```

## Register the legacy runs

One child per SSP leaf. A child, not a sibling: everything except the rate table — climate,
projected employment, calibration — is identical and should be *inherited*, not copied. The only
slice these runs override is `trans_rates_t`.

```{r}
#| label: legacy-runs
runs <- db$runs_t

if (!"rate_solver" %in% names(runs)) {
  runs[, rate_solver := NA_character_]
}
runs[!is.na(ssp) & is.na(rate_solver), rate_solver := "solver"]

parent_runs <- runs[!is.na(ssp) & rate_solver == "solver"]
stopifnot("no SSP leaf runs - run 00-setup-db.qmd" = nrow(parent_runs) > 0L)

existing_legacy <- runs[rate_solver == "legacy", id_run]

legacy_runs <- parent_runs[, .(
  id_run = max(runs$id_run) + seq_len(.N),
  parent_id_run = id_run,
  description = paste0(description, " [original-behaviour rates]"),
  ssp,
  climate_framing,
  gwl_trajectory,
  gwl_quantile,
  rate_solver = "legacy"
)]

if (length(existing_legacy) == 0L) {
  db$runs_t <- as_runs_t(rbind(runs, legacy_runs, fill = TRUE))
} else {
  message("Legacy runs already registered; reusing them.")
  legacy_runs <- runs[rate_solver == "legacy"]
}

db$runs_t[rate_solver == "legacy", .(id_run, parent_id_run, ssp, description)][1:5]
```

## Solve with the original's effective settings

```{r}
#| label: legacy-solve
legacy_solutions <- vector("list", length(ssp_list))
names(legacy_solutions) <- ssp_list

# Shared settings reproducing defects 1 and 2's effects.
legacy_args <- list(
  init_area = init_area,
  bounds = bounds,
  periods = db$periods_t,
  mu_shape = 0, # defect 1: chosen_shape had no effect
  mu_smooth = 100, # the call site's weight; forces straight lines
  mu_target = 1e3, # see below: the original's hard band is infeasible once edges are forbidden
  terminal_band = NA,
  mu_historic = 0,
  fairness = FALSE,
  max_reachability_ratio = max_reachability_ratio
)

leakage <- vector("list", length(ssp_list))
names(leakage) <- ssp_list

for (s in ssp_list) {
  d <- targets_by_ssp[ssp == s]
  message("Solving ", s, " with original-behaviour settings ...")

  # (a) committable: forbidden edges hard-zeroed, so the result can be persisted
  legacy_solutions[[s]] <- do.call(solve_trans_rates, c(
    legacy_args,
    list(
      targets = d[, .(id_lulc, area)],
      shapes = d[, .(id_lulc, shape)], # supplied, but inert at mu_shape = 0
      forbid_non_viable = TRUE
    )
  ))

  # (b) measurement only: how much flow would defect 3 have put on never-observed edges?
  unforbidden <- do.call(solve_trans_rates, c(
    legacy_args,
    list(
      targets = d[, .(id_lulc, area)],
      shapes = d[, .(id_lulc, shape)],
      forbid_non_viable = FALSE
    )
  ))
  leakage[[s]] <- data.table(
    ssp = s,
    flow_non_viable = round(unforbidden$diagnostics$flow_summary$flow_non_viable),
    total_flow = round(unforbidden$diagnostics$flow_summary$total_flow)
  )
}

leakage_summary <- data.table::rbindlist(leakage)
leakage_summary[, pct_of_flow := round(100 * flow_non_viable / total_flow, 1)]
leakage_summary

legacy_summary <- data.table::rbindlist(lapply(ssp_list, function(s) {
  sol <- legacy_solutions[[s]]
  fs <- sol$diagnostics$flow_summary
  data.table(
    ssp = s,
    status = sol$status,
    objective = round(sol$objective, 4),
    total_flow = round(fs$total_flow),
    pct_flow_above_max = round(100 * fs$flow_above_max_rate / fs$total_flow, 1),
    flow_non_viable = round(fs$flow_non_viable)
  )
}))

legacy_summary
```

`flow_non_viable` above is defect 3's cost: flow the original's LP would have placed on edges
never observed in the historic record. evoland cannot allocate it — those transitions have no
`trans_pot_t` rows — so it is reported here and goes no further.

Note also that `static` is **absorbing** in this pipeline: `04` excludes it as an anterior class,
so there are edges into it but none out. Any scenario asking static to *shrink* is therefore not
merely outside history but structurally infeasible, and the reachability ratio comes back `Inf`.
That is why `max_reachability_ratio` has to be lifted rather than merely raised; the L1 terminal
fit then absorbs the miss and reports it in `target_error`. Whether to let static be left as an
anterior class, or to pre-adjust the demand so it never shrinks, is an open question in
`TODO.md`.

## Commit

```{r}
#| label: legacy-commit
for (s in ssp_list) {
  run_ids <- db$runs_t[rate_solver == "legacy" & ssp == s, id_run]
  if (length(run_ids) == 0L) {
    warning("No legacy runs for ", s, "; skipping")
    next
  }
  db$trans_rates_t <- trans_rates_from_solution(legacy_solutions[[s]], id_run = run_ids)
}
```

## The diff — what the defects actually cost

The point of the exercise. Both rate tables are now in the DB, so replay each forward and compare
the trajectories they imply.

```{r}
#| label: legacy-diff
compare_run_pair <- function(ssp_name) {
  solver_run <- db$runs_t[rate_solver == "solver" & ssp == ssp_name, id_run][1]
  legacy_run <- db$runs_t[rate_solver == "legacy" & ssp == ssp_name, id_run][1]

  a_solver <- trans_rate_areas(init_area, db$trans_rates_t[id_run == solver_run], db$trans_meta_t)
  a_legacy <- trans_rate_areas(init_area, db$trans_rates_t[id_run == legacy_run], db$trans_meta_t)

  merge(
    data.table::as.data.table(a_solver),
    data.table::as.data.table(a_legacy),
    by = c("id_lulc", "id_period"),
    suffixes = c("_solver", "_legacy")
  )[, .(ssp = ssp_name, id_lulc, id_period,
        delta = area_solver - area_legacy)]
}

trajectory_diff <- data.table::rbindlist(lapply(ssp_list, compare_run_pair))

trajectory_diff[
  lulc_meta[, .(id_lulc, name)],
  on = "id_lulc",
  nomatch = NULL
][, .(
  max_abs_delta = round(max(abs(delta))),
  final_delta = round(delta[which.max(id_period)])
), by = .(ssp, name)][order(-max_abs_delta)][1:20]
```

### Expect this diff to be small, and that is the finding

Exercised on the real demand with a stand-in edge set, the committable legacy trajectories differ
from the corrected ones by **at most 7 cells**, and by **0 at the horizon**, in every scenario.

That is not a bug in the comparison. Once non-viable edges are forbidden in both — which they must
be, to commit anything — the remaining differences are `mu_shape`, `mu_smooth` and `fairness`, and
none of them bites here: the shape mechanism is inert regardless of the defect (a straight line
satisfies all five shapes with equality), and the smoothing weight makes no difference over four
steps.

So the practical consequence of the original's defects is **not** distorted trajectories. It is
the leakage: 40–65 % of the solved flow placed on edges never observed, which its own allocator
would then have dropped. The class trajectories the original published are, as far as this
comparison can tell, what a corrected solver would also have produced — it simply intended to get
there by routes that do not exist.

Re-run this against the real `trans_meta_t` before quoting the numbers: the stand-in used the
original study's 21-edge calibration set, and this pipeline's viable set will differ.

```{r}
#| label: legacy-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key =                      , value =,
  "legacy_rates_method",       "solve_trans_rates() parameterised to the original's effective behaviour",
  "legacy_rates_settings",     "mu_shape=0, mu_smooth=100, mu_target=0, terminal_band=0.01, fairness=FALSE, forbid_non_viable=FALSE",
  "legacy_rates_caveat",       "behavioural emulation, not bit-exact; the original's per-SSP rate tables were never published",
  "legacy_rates_scenarios",    paste(ssp_list, collapse = ",")
))
```
