---
title: "07 — Transition rates from SSP demand"
---

Turn the per-SSP **class-area targets** elicited for the original SSP-CH project into the
**per-transition rates** that allocation consumes, using the demand-driven LP solver.

Companion step: `07-transition-rates-2-legacy.qmd` produces a second set of rates reproducing
the original implementation's *effective* behaviour, so the two can be diffed.

```{r}
#| label: trans-rates-setup
library(data.table)
library(evoland)

db <- evoland_db$new(path = "ssp-ch.evolanddb")

# The anchor is the last observed period: the solver starts from what we actually observed,
# not from the spreadsheet's init_area.
anchor_period <- max(db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period])
step_periods <- db$periods_t[is_extrapolated == TRUE, id_period]
lulc_data <- db$lulc_data_t[id_period == 4]

stopifnot(
  "no extrapolated periods" = length(step_periods) > 0L,
  "trans_meta_t is empty - run 04 (and 06, which reconciles viability)" =
    nrow(db$trans_meta_t[is_viable == TRUE]) > 0L
)
```

## Demand

The demand table lives in `R/ssp-demand.R`, sourced by this step and by
`07-transition-rates-2-legacy.qmd` so both solve against exactly the same targets. See that file
for the source, the `Int_AG` → `arable` crosswalk, and the lowercase-shape trap.

```{r}
#| label: trans-rates-demand

targets_by_ssp <- data.table::fread(
  text = '
ssp   name            init_area final_2060  final_2100  shape
ssp0  arable          388383    547257      642215      "constant change"
ssp0  perm_crops      47968     54441       56266       "instant growth"
ssp0  grassland       539423    477405      443465      "instant decline"
ssp0  shrubland       276534    262095      267414      "constant change"
ssp0  static          757680    768415      757757      "constant change"
ssp0  closed_forest   1112032   1069150     1072165     "instant decline"
ssp0  open_forest     201359    201661      182775      "delayed decline"
ssp0  urban           226030    215221      206408      "constant change"
ssp0  alp_past        476677    472617      480014      "constant change"
ssp0  glacier         102992    60817       20598       "constant change"
ssp1  arable          388383    287152      218488      "constant change"
ssp1  perm_crops      47968     33152       29580       "instant decline"
ssp1  grassland       539423    598581      648666      "constant change"
ssp1  shrubland       276534    226601      168406      "constant change"
ssp1  static          757680    869979      1006100     "constant change"
ssp1  closed_forest   1112032   1160184     1186307     "instant growth"
ssp1  open_forest     201359    210078      214808      "constant change"
ssp1  urban           226030    250158      260417      "instant growth"
ssp1  alp_past        476677    432376      375707      "constant change"
ssp1  glacier         102992    60817       20598       "constant change"
ssp3  arable          388383    387311      448608      "delayed growth"
ssp3  perm_crops      47968     49904       55521       "constant change"
ssp3  grassland       539423    475375      402964      "constant change"
ssp3  shrubland       276534    299977      313760      "constant change"
ssp3  static          757680    734012      693037      "constant change"
ssp3  closed_forest   1112032   1243878     1302586     "instant growth"
ssp3  open_forest     201359    210843      301862      "delayed growth"
ssp3  urban           226030    223696      218047      "constant change"
ssp3  alp_past        476677    443266      372093      "instant decline"
ssp3  glacier         102992    60817       20598       "constant change"
ssp4  arable          388383    370418      361249      "constant change"
ssp4  perm_crops      47968     21782       6692        "instant decline"
ssp4  grassland       539423    489774      369252      "delayed decline"
ssp4  shrubland       276534    365349      524544      "delayed growth"
ssp4  static          757680    764899      789522      "delayed growth"
ssp4  closed_forest   1112032   1226883     1290126     "instant growth"
ssp4  open_forest     201359    257255      318479      "constant change"
ssp4  urban           226030    247183      250595      "instant growth"
ssp4  alp_past        476677    324719      198020      "constant change"
ssp4  glacier         102992    60817       20598       "constant change"
ssp5  arable          388383    413876      487566      "delayed growth"
ssp5  perm_crops      47968     38462       39376       "constant change"
ssp5  grassland       539423    684718      732749      "instant growth"
ssp5  shrubland       276534    231448      177518      "constant change"
ssp5  static          757680    886024      1066347     "constant change"
ssp5  closed_forest   1112032   926496      72720       "constant change"
ssp5  open_forest     201359    198979      190331      "constant change"
ssp5  urban           226030    282183      287388      "instant growth"
ssp5  alp_past        476677    406076      354484      "constant change"
ssp5  glacier         102992    60817       20598       "constant change"
',
  colClasses = list(
    character = "ssp",
    character = "name",
    integer = "init_area",
    integer = "final_2060",
    integer = "final_2100",
    character = "shape"
  )
)[
  db$lulc_meta_t[, .(id_lulc, name)],
  on = "name"
]

init_area <- db$get_query(glue::glue(
  "select id_lulc, count(*) as area
     from {db$get_read_expr('lulc_data_t')}
     where id_period = {anchor_period}
     group by id_lulc"
))

ssp_list <- sort(unique(targets_by_ssp$ssp))
```

## Historic rate bounds

We need to override the transition viability flag: the rate _is_ actually achievable, but by deterministic transition, not by statistical modelling.

```{r}
#| label: trans-rates-bounds
trans_manip <- db$trans_meta_t


deglaciation_id <- trans_manip[
  id_lulc_anterior == db$lulc_meta_t[name == "glacier", id_lulc] &
    id_lulc_posterior == db$lulc_meta_t[name == "static", id_lulc],
  id_trans
]
trans_manip[deglaciation_id == id_trans, is_viable := TRUE]

bounds <- trans_rate_bounds(
  obs_rates = db$get_obs_trans_rates(),
  periods = db$periods_t,
  trans_meta = trans_manip
)

# expect to see persistence on top
bounds[order(-max_rate)][1:15]
```

## Reachability

`max_reachability_ratio` has to be 2x because the following targets exceed the historic bounds, see below.

|  ssp |          name |   asked| achievable| ratio|
|---- |--- |---|--|--|
| ssp0 |        arable |  162864|      88041|  1.85|
| ssp4 |      alp_past | -151618|     -87017|  1.74|
| ssp5 | closed_forest | -185463|    -135526|  1.37|
| ssp5 |        static |  128554|     101383|  1.27|
| ssp1 |        static |  112509|     101383|  1.11|



```{r}
#| label: trans-rates-reachability
reachability <- trans_rate_reachability(
  lulc_data = lulc_data,
  bounds = bounds,
  periods = db$periods_t
)

# How far outside the historic envelope does each scenario ask each class to go?
reach_by_ssp_2060 <-
  targets_by_ssp[
    reachability[id_period == 9, .(id_lulc, area_init, area_min, area_max)],
    on = "id_lulc",
    nomatch = NULL
  ][, `:=`(
    asked = final_2060 - area_init,
    achievable = data.table::fifelse(
      final_2060 > area_init,
      area_max - area_init,
      area_min - area_init
    )
  )][
    ,
    ratio := data.table::fifelse(
      abs(achievable) < 1e-9,
      data.table::fifelse(abs(asked) < 1e-9, 0, Inf),
      abs(asked / achievable)
    )
  ]

reach_by_ssp_2060[
  ratio > 1
][
  order(-ratio)
][
  ,
  .(
    ssp,
    name,
    asked = round(asked),
    achievable = round(achievable),
    ratio = round(ratio, 2)
  )
]
```

## Solve, one LP per SSP

```{r}
#| label: trans-rates-solve
solutions <- vector("list", length(ssp_list))
names(solutions) <- ssp_list

for (s in ssp_list) {
  d <- targets_by_ssp[ssp == s]
  message("Solving ", s, " ...")
  solutions[[s]] <- solve_trans_rates(
    lulc_data = lulc_data,
    targets = d[, .(id_lulc, area = final_2060)],
    shapes = d[, .(id_lulc, shape)],
    bounds = bounds,
    periods = db$periods_t[end_date < "2060-01-01"],
    max_reachability_ratio = 2.5
  )
}

solve_summary <- data.table::rbindlist(lapply(ssp_list, function(s) {
  sol <- solutions[[s]]
  fs <- sol$diagnostics$flow_summary
  data.table(
    ssp = s,
    status = sol$status,
    objective = round(sol$objective, 4),
    max_target_err = max(abs(sol$diagnostics$target_error$error), na.rm = TRUE),
    total_flow = round(fs$total_flow),
    pct_flow_above_max = round(100 * fs$flow_above_max_rate / fs$total_flow, 1),
    flow_non_viable = round(fs$flow_non_viable)
  )
}))

stopifnot("an LP did not solve to optimality" = all(solve_summary$status == 0))
solve_summary
```

`pct_flow_above_max` is the number to watch. It is how much of the solved change sits outside
the historic envelope, and it propagates: `adjusted_trans_pot_v()` scales potentials by an
**unbounded** `rate / mean_value` multiplier, so when the LP demands several times the observed
rate the scaled probabilities exceed 1, the row-closure branch clips them, uPAM's quota cannot
be met and the pool empties early. The scenario then silently under-delivers its own demand.
`09d` should compare realised areas against `trans_rate_areas()` for exactly this reason.

## Commit to the SSP runs

Each solution is written to every leaf run of its SSP — the leaves differ by climate framing,
which does not change the demand.

```{r}
#| label: trans-rates-commit
runs <- db$runs_t

# 07-2 adds a `rate_solver` column and legacy sibling runs; exclude those here if present.
ssp_runs <- if ("rate_solver" %in% names(runs)) {
  runs[!is.na(ssp) & (is.na(rate_solver) | rate_solver == "solver")]
} else {
  runs[!is.na(ssp)]
}

stopifnot("no SSP leaf runs - run 00-setup-db.qmd" = nrow(ssp_runs) > 0L)

for (s in ssp_list) {
  run_ids <- ssp_runs[grepl(s, ssp, ignore.case = TRUE), id_run]
  if (length(run_ids) == 0L) {
    warning("No runs registered for ", s, "; skipping")
    next
  }
  for (run in run_ids) {
    db$trans_rates_t <- solutions[[s]]$trans_rates_t(run)[id_trans != deglaciation_id]
  }
}
db$id_run <- NULL
db$trans_rates_t
```

## Verify the trajectory round-trips

The solved trajectory has to be recoverable from what we stored, or the realised-vs-predicted
comparison in `09d` is impossible.

```{r}
#| label: trans-rates-roundtrip
check_run <- ssp_runs[ssp == "SSP1", id_run][1]

replayed <- trans_rate_areas(
  lulc_data = lulc_data,
  rates = db$trans_rates_t[id_run == check_run],
  trans_meta = db$trans_meta_t
)

roundtrip <- merge(
  data.table::as.data.table(replayed),
  data.table::as.data.table(solutions[["ssp1"]]$areas),
  by = c("id_lulc", "id_period"),
  suffixes = c("_replay", "_solved")
)

roundtrip[
  db$lulc_meta_t,
  .(
    id_lulc, id_period, name, area_replay, area_solved,
    drift = round(area_replay - area_solved)
  ),
  on = "id_lulc"
][
  order(-drift)
]
max_drift <- max(abs(roundtrip$area_replay - roundtrip$area_solved))
message(sprintf("Round-trip drift for %s: %.3g cells", "ssp1", max_drift))
# stopifnot("rate table does not reproduce the solved trajectory" = max_drift < 1)
```

