---
title: "08 — Backcasting validation"
---

Validate the calibrated model against the past: replay the **observed** transition rates
through the fitted transition-potential models and the estimated allocation parameters, then
compare the resulting maps against the Arealstatistik surveys we actually observed, substeps:

(a) **estimate patch / allocation parameters** from the observed periods
(b) **run the backcast** over a set of parameter choices, several stochastic replicates each;
(c) **validate** the backcast against the observed periods, and define acceptance criteria.

## What this does and does not test

The rates fed to the backcast are the **observed** ones (`get_obs_trans_rates()`), so the
*quantity* of change per transition and period is prescribed, not predicted. What is on trial
is therefore only the **allocation** — whether the model puts the right changes in the
right places — plus the allocator's ability to deliver the prescribed quantity at all.

Two consequences worth stating before any number is read:

- **Non-viable transitions bound the score from above.** Observed change belonging to a
  transition that `04`/`06` dropped can never be reproduced, because `predict_trans_pot()` only
  predicts for viable transitions and `adjusted_trans_pot_v()` inner-joins potentials to rates.
  Rows for non-viable transitions in `trans_rates_t` are inert. The `unmodellable` share below
  quantifies that ceiling.
- **Persistence is not skill.** The great majority of Swiss hectares keep their class across a
  survey interval — `040d`'s observed-transition plot has the exact figure — so any whole-map
  agreement statistic is dominated by cells nobody had to get right. Every metric here is
  therefore computed on **change**, not on the map. `agreement` is reported alongside the figure
  of merit only to show how uninformative it is.

## Parameters

`n_perturbations` × `n_replicates` chained backcasts, each covering three period steps, is the
cost driver: every allocated period re-runs `predict_trans_pot()` over the full 4.1 M-cell grid
for every viable transition, because `alloc_clumpy()` predicts inside the same call that
allocates. Start small and grow the grid once the runtime is known.

```{r}
#| label: backcast-params
library(data.table)
library(evoland)

n_perturbations <- 3L # perturbed parameter sets *in addition to* the best estimate
perturb_sd <- 0.05 # sd of the frac_expander perturbation (evoland default)
n_replicates <- 3L # stochastic allocations per parameter set
seed <- 42L
fuzzy_window <- 11L # odd; evoland's default, ~1 km at 100 m resolution
select_score <- "classif.auc" # must match 06
select_maximize <- TRUE
id_run_backcast_base <- 2000L # 2000-2999 is free: 001 uses 0/100/1xxx, 071 uses max+n

db <- evoland_db$new(path = "ssp-ch.evolanddb")
db$id_run <- 0L

# evoland's table classes carry print methods that assume their own columns; a summary derived
# from one inherits the class and prints nonsense ("Rate range: [Inf, -Inf]"). Strip it.
plain <- function(x) {
  data.table::setattr(data.table::copy(x), "class", c("data.table", "data.frame"))[]
}

obs_periods <- db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period]
id_period_initial <- min(obs_periods) # the backcast's starting state, never simulated
backcast_periods <- sort(setdiff(obs_periods, id_period_initial))
id_period_final <- max(backcast_periods)

stopifnot(
  "need at least two observed periods" = length(backcast_periods) >= 1L,
  "backcast periods must be contiguous" = all(diff(backcast_periods) == 1L),
  "no viable transitions - run 04 (and 06, which reconciles viability)" =
    nrow(db$trans_meta_t[is_viable == TRUE]) > 0L,
  "no fitted models - run 06-transition-modelling.qmd first" =
    db$row_count("trans_models_t") > 0L
)

message(
  "Backcasting periods ", paste(backcast_periods, collapse = " -> "),
  " from observed period ", id_period_initial
)
```

## Observed rates as the backcast demand

Written to run `0`: they are a property of the observed record, not of a scenario, and every
backcast run inherits them through the run lineage. The scenario rates `07` wrote sit on the SSP
leaf runs at the *extrapolated* periods, so the two never resolve to the same
`(id_period, id_trans)` slice.

The unmodellable cells are from deglaciation and relatively rare transitions, respectively. We will be deterministically overwriting the glacier transitions, because that's also how the extrapolation is done.

```{r}
#| label: backcast-obs-rates
obs_rates <- db$get_obs_trans_rates()[id_period %in% backcast_periods]
db$trans_rates_t <- obs_rates

# How much observed change is out of reach because its transition is not viable?
viable_trans <- db$trans_meta_t[is_viable == TRUE]
unmodellable <- plain(obs_rates)[
  db$trans_meta_t[
    db$lulc_meta_t[, .(id_lulc, name_anterior = name)],
    on = c(id_lulc_anterior = "id_lulc")
  ][
    db$lulc_meta_t[, .(id_lulc, name_posterior = name)],
    on = c(id_lulc_posterior = "id_lulc")
  ],
  on = "id_trans",
  nomatch = NULL
][
  ,
  .(
    name_anterior,
    name_posterior,
    cells = sum(count),
    cells_viable = sum(count[is_viable == TRUE]),
    is_viable
  ),
  by = .(id_period, id_trans)
][, `:=`(
  unmodellable_pct = round(100 * (1 - cells_viable / cells), 2)
)][]
unmodellable[is_viable == FALSE & cells > 1000]
```

## (a) Allocation parameters

`create_alloc_params_t()` measures, per viable transition and per observed period pair, the mean
and variance of new-patch area, patch elongation, and the split between cells that **expand** an
existing patch of the target class and cells that start a **new** patch; it then averages over
the period pairs and emits `n_perturbations` further sets that jitter `frac_expander`
(`frac_patcher` following as `1 - frac_expander`).

Only `frac_expander` is perturbed upstream, so this sweep explores the expander/patcher split
and nothing else. That is the parameter the SSP interventions also target, which makes it the
defensible one to sweep first, but it means a poor `mean_patch_size` estimate will not show up
as parameter sensitivity here.

```{r}
#| label: backcast-alloc-params
alloc_params <- db$create_alloc_params_t(
  n_perturbations = n_perturbations,
  sd = perturb_sd
)

# The returned id_run is a 1-based parameter-set index, not a real run.
param_sets <- sort(unique(alloc_params$id_run))
plain(alloc_params)[, .(
  n_trans = .N,
  mean_patch_size = round(mean(mean_patch_size, na.rm = TRUE), 2),
  frac_expander = round(mean(frac_expander, na.rm = TRUE), 3)
), by = id_run]
```

A transition that never produced a new patch (all conversions expanded existing ones) comes back
with `patch_elongation = NA`; `alloc_clumpy()` substitutes `0.5`, and `mean_patch_size = 0` is
handled as a mono-pixel draw in the C++ patcher. Both are safe, but a transition in that state
has no estimated patch geometry at all — worth knowing before reading its similarity.

```{r}
#| label: backcast-alloc-params-gaps
plain(alloc_params)[id_run == min(param_sets) & (is.na(patch_elongation) | mean_patch_size <= 1)][
  db$trans_meta_t,
  on = "id_trans",
  nomatch = NULL
][, .(id_trans, id_lulc_anterior, id_lulc_posterior, mean_patch_size, patch_elongation)]
```

## Run registration

One parent run per parameter set holding the `alloc_params_t` rows, and one child run per
stochastic replicate holding only the simulated `lulc_data_t`. The replicates inherit the
parameters through the lineage, so — as in the evoland stochastic-allocation vignette — the only
thing that differs between replicates of a set is the RNG state.

```{r}
#| label: backcast-register-runs
param_run_id <- function(p) id_run_backcast_base + 100L * as.integer(p)
member_run_id <- function(p, r) param_run_id(p) + as.integer(r)

param_runs <- data.table(
  id_run = param_run_id(param_sets),
  parent_id_run = 0L,
  description = paste0(
    "Backcasting allocation parameter set ", param_sets,
    fifelse(param_sets == min(param_sets), " (unperturbed estimate)", " (perturbed)")
  ),
  kind = "backcast parameters"
)

member_runs <- CJ(param_set = param_sets, replicate = seq_len(n_replicates))[, .(
  id_run = member_run_id(param_set, replicate),
  parent_id_run = param_run_id(param_set),
  description = paste0("Backcast replicate ", replicate, " of parameter set ", param_set),
  kind = "backcast replicate",
  param_set = param_set,
  replicate = replicate,
  seed = seed + 1000L * param_set + replicate
)]

# Re-running this step must replace its own runs, not collide with them; anything else in the
# 2000-block is somebody else's and is a hard error.
backcast_kinds <- c("backcast parameters", "backcast replicate")
existing_runs <- db$runs_t
mine <- if ("kind" %in% names(existing_runs)) {
  existing_runs[kind %in% backcast_kinds, id_run]
} else {
  integer(0)
}
clash <- setdiff(
  intersect(c(param_runs$id_run, member_runs$id_run), existing_runs$id_run),
  mine
)
stopifnot("backcast run ids collide with runs registered by 001/071" = length(clash) == 0L)

if (length(mine) > 0L) {
  message(
    "Replacing ", length(mine), " backcast runs from an earlier execution; ",
    "their lulc_data_t / trans_pot_t rows are overwritten in place."
  )
}

db$commit(
  as_runs_t(rbind(
    existing_runs[!id_run %in% c(param_runs$id_run, member_runs$id_run)],
    param_runs,
    member_runs,
    fill = TRUE
  )),
  "runs_t",
  "overwrite" # adding columns
)

for (p in param_sets) {
  db$alloc_params_t <- as_alloc_params_t(
    copy(alloc_params[id_run == p])[, id_run := param_run_id(p)]
  )
}

# --- interventions (not implemented here) ----------------------------------------------
# The SSP `Pre-allocation` interventions are perturbations of exactly this table. Each
# NCCS-SSP-scenarios/Tools/SSP*_interventions.yml opens with a `Spatial_zoning` block whose
# `Param_adjust_name` is one of Mean_Patch_Size / Patch_Size_Variance / Patch_Isometry /
# Perc_expander / Perc_patcher -- i.e. the alloc_params_t columns estimated above -- applied to
# the transitions targeting one class (`Transition_target_classes: Urban` in all five files).
# To realise them, write a modified copy of the winning parameter rows onto the SSP run instead
# of letting it inherit run 0's:
#
#   mod <- copy(alloc_params[id_run == best_param])[
#     id_trans %in% urban_trans, frac_patcher := 0.15][      # SSP1 Absolute 0.15
#     id_trans %in% urban_trans, frac_expander := 1 - frac_patcher]
#   db$alloc_params_t <- as_alloc_params_t(mod[, id_run := ssp_run_id])
#
# Two conversions are needed and are not free:
#   * `Absolute` replaces the estimated value (SSP0/3/5 set Perc_patcher to 0, SSP1 to 0.15);
#     `Relative` (only SSP4, 0.5) scales it, but the YAML header does not say what it scales.
#     Whatever it is, the numbers were tuned against the *original's* parameter estimates:
#     Spatial_intervention_updates.txt records SSP1's 0.20 target being cut to 0.15 "because
#     some of the transitions were already close". They do not transfer unexamined to ours.
#   * `Patch_Isometry` is a Dinamica parameter. alloc_clumpy consumes `patch_elongation`
#     (alloc_params_clumpy_v maps it straight through), so an isometry intervention has to be
#     inverted back through isometry_from_elongation(). That mapping is a 1-D interpolation of
#     Mazy fig. 3.10 and is not injective over its flat segments.
```

## (b) Run the backcast

Each replicate starts from the observed `id_period_initial` map — which the run lineage resolves
to run `0`, since the replicate writes nothing for that period — and then chains: the simulated
period 2 map is the anterior state for period 3, and so on. The error therefore **compounds**,
which is the point: a backcast across three decades of surveys tests the trajectory, not one
step.

> An alternative design, noted in the handover, is to branch *N* members per allocation step from
> the observed anterior state instead of from the previous simulation. That isolates one-step
> allocation skill and is strictly cheaper to interpret. It needs no new machinery here: register
> the members under the same parent and allocate a **single** period each
> (`id_periods = k`), writing only period `k`; the lineage then resolves `k - 1` to run `0`'s
> observed map. Both designs are worth having; the chained one is the one a backcast is usually
> asked for, so it is the one implemented.

```{r}
#| label: backcast-allocate
#| output: false
backcast_replicate_runs <- db$runs_t[kind == "backcast replicate", .(id_run, seed)]
future::plan(future.mirai::mirai_multisession, workers = 10)
for (id in backcast_replicate_runs[2:12, id_run]) {
  db$id_run <- id

  # NB: db$alloc_clumpy() documents a `seed` argument, but alloc_clumpy() itself does not
  # accept one at the pinned commit -- create_method_binding() forwards it verbatim and the
  # call dies with "unused argument (seed = ...)". The vignette
  # (stochastic-allocation-sensitivity.qmd) passes it and would fail for the same reason.
  # All CLUMPY randomness goes through R's RNG (src/alloc_clumpy.cpp), so set.seed() is an
  # exact substitute. Fix upstream, then move the seed back into the call.
  set.seed(backcast_replicate_runs[id_run == id, seed])

  db$alloc_clumpy(
    id_periods = backcast_periods,
    select_score = select_score,
    select_maximize = select_maximize
  )
}
future::plan("sequential")
db$id_run <- 0L
```

Each of these runs writes its own `trans_pot_t` slice (one value per cell, viable transition and
period), which is by far the largest thing the backcast stores. The potentials for the *first*
backcast period are in fact identical across every replicate and every parameter set — same
observed anterior state, same models, same predictors — so recomputing them per run is pure
waste. `alloc_clumpy_one_period()` has a `use_parent_trans_pot` argument for this, but the
multi-period `alloc_clumpy()` wrapper does not expose it; exposing it upstream would remove
roughly a third of the prediction cost of this step.

## Observed and simulated end states

```{r}
#| label: backcast-maps
db$id_run <- 0L
observed <- merge(
  db$fetch("lulc_data_t", where = glue::glue(
    "id_run = 0 and id_period = {id_period_initial}"
  ))[, .(id_coord, id_lulc_initial = id_lulc)],
  db$fetch("lulc_data_t", where = glue::glue(
    "id_run = 0 and id_period = {id_period_final}"
  ))[, .(id_coord, id_lulc_observed = id_lulc)],
  by = "id_coord"
)

db$id_run <- NULL # fetch() is lineage-filtered; NULL reads across runs
simulated <- db$fetch("lulc_data_t", where = glue::glue(
  "id_run in ({toString(member_runs$id_run)}) and id_period = {id_period_final}"
))[, .(id_run, id_coord, id_lulc_simulated = id_lulc)]
db$id_run <- 0L

cells <- member_runs[, .(id_run, param_set, replicate)][
  simulated[observed, on = "id_coord", nomatch = NULL],
  on = "id_run"
]

stopifnot(
  "no simulated cells - did allocation write anything?" = nrow(cells) > 0L,
  "a replicate is missing cells" =
    uniqueN(cells$id_run) == nrow(member_runs)
)
cells[, .N, by = .(param_set, replicate)]
```

## Quantity check: did the allocator deliver the prescribed change?

The first thing to establish is that the allocator hit the demand it was given. uPAM draws
patches without replacement from the source pool; when the prescribed rate is large relative to
the pool, or `avoid_aggregation` rejects many candidate patches, the pool empties before the
quota is met and the run silently under-delivers. `07` flags the same failure mode for the
scenario rates, where it is worse because the LP pushes rates above the historic envelope.

```{r}
#| label: backcast-quantity
realised_counts <- rbindlist(lapply(member_runs$id_run, function(r) {
  db$id_run <- r
  db$get_obs_trans_rates()[
    id_period %in% backcast_periods,
    .(id_run = r, id_period, id_trans, count_realised = count)
  ]
}))
db$id_run <- 0L

quantity <- merge(
  CJ(id_run = member_runs$id_run, id_trans = viable_trans$id_trans, id_period = backcast_periods)[
    obs_rates[, .(id_period, id_trans, count_prescribed = count)],
    on = c("id_period", "id_trans"),
    nomatch = NULL
  ],
  realised_counts,
  by = c("id_run", "id_period", "id_trans"),
  all.x = TRUE
)
quantity[is.na(count_realised), count_realised := 0L]
quantity[, shortfall_pct := 100 * (count_prescribed - count_realised) / count_prescribed]

quantity[order(-shortfall_pct)][1:15][
  db$trans_meta_t,
  on = "id_trans",
  nomatch = NULL
][, .(
  id_run, id_period, id_lulc_anterior, id_lulc_posterior,
  count_prescribed, count_realised,
  shortfall_pct = round(shortfall_pct, 1)
)]
```

## (c1) Fuzzy similarity of differences

`calc_transition_similarity()` is evoland's port of Dinamica's "similarity of differences": for
one transition it builds the observed and simulated **change** maps, then compares them with a
distance-decayed moving window, so a patch placed one cell off still scores.

Two things about the returned number:

- It is computed over the **whole raster**, not over the changed cells: the change maps are
  binarised with `NA -> 0` and the background agrees almost everywhere, so `similarity` sits
  near 1 for every transition and every parameter set and discriminates poorly. We therefore
  also report `similarity_change`, the mean of the same similarity surface **masked to cells
  that changed in either map**, which is the quantity Dinamica's validation actually reads.
- `similarity` is the *minimum* of the two directional means; `similarity_change` is masked from
  `sim1` only (observed → simulated), because that is the only surface the function returns.

```{r}
#| label: backcast-fuzzy
db$id_run <- 0L
rast_initial <- db$lulc_data_as_rast(id_period = id_period_initial)
rast_obs_final <- db$lulc_data_as_rast(id_period = id_period_final)

fuzzy_one_run <- function(id_run_member) {
  db$id_run <- id_run_member
  rast_sim <- db$lulc_data_as_rast(id_period = id_period_final)

  res <- mapply(
    FUN = calc_transition_similarity,
    from_class = viable_trans$id_lulc_anterior,
    to_class = viable_trans$id_lulc_posterior,
    MoreArgs = list(
      initial_map = rast_initial,
      observed_map = rast_obs_final,
      simulated_map = rast_sim,
      window_size = fuzzy_window,
      use_exp_decay = TRUE,
      decay_divisor = 2.0
    ),
    SIMPLIFY = FALSE
  )

  masked_mean <- function(x) {
    if (is.null(x$similarity_map) || !is.finite(x$similarity)) {
      return(NA_real_)
    }
    changed <- !is.na(x$observed_change) | !is.na(x$simulated_change)
    terra::global(
      terra::mask(x$similarity_map, changed, maskvalues = c(0, NA)),
      "mean",
      na.rm = TRUE
    )[1, 1]
  }

  data.table(
    id_run = id_run_member,
    id_trans = viable_trans$id_trans,
    similarity = vapply(res, \(x) as.numeric(x$similarity), numeric(1L)),
    similarity_change = vapply(res, masked_mean, numeric(1L)),
    n_observed = vapply(res, \(x) as.numeric(x$n_observed), numeric(1L)),
    n_simulated = vapply(res, \(x) as.numeric(x$n_simulated), numeric(1L))
  )
}

fuzzy_dt <- member_runs[, .(id_run, param_set, replicate)][
  rbindlist(lapply(member_runs$id_run, fuzzy_one_run)),
  on = "id_run"
]
db$id_run <- 0L

fuzzy_by_param <- fuzzy_dt[, .(
  similarity = mean(similarity, na.rm = TRUE),
  similarity_change = mean(similarity_change, na.rm = TRUE),
  similarity_change_sd = sd(similarity_change, na.rm = TRUE),
  n_observed = mean(n_observed),
  n_simulated = mean(n_simulated)
), by = .(param_set, id_trans)]

fuzzy_by_param[db$trans_meta_t, on = "id_trans", nomatch = NULL][
  order(param_set, similarity_change)
][, .(
  param_set, id_lulc_anterior, id_lulc_posterior,
  similarity = round(similarity, 3),
  similarity_change = round(similarity_change, 3),
  n_observed, n_simulated = round(n_simulated)
)]
```

## (c2) Cross-tabulation and figure of merit

The cross-tabulation of observed against simulated end class, restricted to the cells that
*should* have changed, decomposes into the four quantities of Pontius' figure of merit:

| | simulated persistence | simulated a change |
| --- | --- | --- |
| **observed a change to _v_** | miss (A) | **hit (B)** if simulated to _v_, wrong class (C) otherwise |
| **observed persistence** | correct null | false alarm (D) |

$\mathrm{FoM} = B / (A + B + C + D)$ — the intersection of observed and simulated change over
their union. It ignores the correct-null cell, which is why it is the right statistic here.

```{r}
#| label: backcast-fom
fom_components <- function(dt) {
  dt[, .(
    hit = sum(id_lulc_observed != id_lulc_initial & id_lulc_simulated == id_lulc_observed),
    miss = sum(id_lulc_observed != id_lulc_initial & id_lulc_simulated == id_lulc_initial),
    wrong_class = sum(
      id_lulc_observed != id_lulc_initial &
        id_lulc_simulated != id_lulc_initial &
        id_lulc_simulated != id_lulc_observed
    ),
    false_alarm = sum(id_lulc_observed == id_lulc_initial & id_lulc_simulated != id_lulc_initial),
    n_cells = .N
  )]
}

fom_by_run <- cells[, fom_components(.SD), by = .(param_set, replicate, id_run)]
fom_by_run[, `:=`(
  fom = hit / (hit + miss + wrong_class + false_alarm),
  agreement = 1 - (miss + wrong_class + false_alarm) / n_cells
)]
fom_by_run[order(-fom)]
```

### The null: allocating change at random within the anterior class

The FoM of a model that gets the quantity right but the location purely by chance is available
in closed form, which is cheaper and less arbitrary than running a scrambled-potential control.
Within an anterior class $u$ of $N_u$ cells, of which the model converts $S_{uv}$ to $v$ and the
record converts $O_{uv}$, random placement gives $\mathbb{E}[B_{uv}] = S_{uv} O_{uv} / N_u$, and
the expected union of change over the class is $S_u + O_u - S_u O_u / N_u$.

```{r}
#| label: backcast-null
n_by_ant <- observed[, .(n_ant = as.numeric(.N)), by = .(id_lulc_ant = id_lulc_initial)]

# Observed flow is a property of the record; simulated flow is the replicate mean per set.
obs_flow <- observed[
  id_lulc_initial != id_lulc_observed,
  .(obs = as.numeric(.N)),
  by = .(id_lulc_ant = id_lulc_initial, id_lulc_post = id_lulc_observed)
]
sim_flow <- cells[
  id_lulc_initial != id_lulc_simulated,
  .(sim = .N / n_replicates),
  by = .(param_set, id_lulc_ant = id_lulc_initial, id_lulc_post = id_lulc_simulated)
]

flow <- merge(
  CJ(param_set = param_sets, unique = TRUE)[, k := 1L][
    obs_flow[, k := 1L],
    on = "k",
    allow.cartesian = TRUE
  ][, -"k"],
  sim_flow,
  by = c("param_set", "id_lulc_ant", "id_lulc_post"),
  all = TRUE
)
flow[is.na(obs), obs := 0][is.na(sim), sim := 0]
flow <- n_by_ant[flow, on = "id_lulc_ant"]

null_by_param <- flow[,
  {
    by_ant <- .SD[, .(
      obs_u = sum(obs),
      sim_u = sum(sim),
      hits = sum(obs * sim / n_ant),
      n_ant = n_ant[1]
    ), by = id_lulc_ant]
    by_ant[, union_u := obs_u + sim_u - obs_u * sim_u / n_ant]
    .(fom_null = sum(by_ant$hits) / sum(by_ant$union_u))
  },
  by = param_set
]

# Per-transition, so a transition allocated at chance can be named rather than averaged away.
null_by_trans <- flow[, .(
  param_set, id_lulc_ant, id_lulc_post,
  hits_null = obs * sim / n_ant,
  union_null = obs + sim - obs * sim / n_ant
)][, fom_null := hits_null / union_null]

null_by_param
```

Per-transition FoM uses the same decomposition, so it is directly comparable with the null
above: hits are cells the record and the model both moved from `u` to `v`, and the denominator
is the union of the two flows.

```{r}
#| label: backcast-fom-by-trans
sim_t <- cells[
  id_lulc_initial != id_lulc_simulated,
  .(sim_t = .N / n_replicates),
  by = .(param_set, id_lulc_ant = id_lulc_initial, id_lulc_post = id_lulc_simulated)
]
hit_t <- cells[
  id_lulc_initial != id_lulc_observed & id_lulc_simulated == id_lulc_observed,
  .(hit_t = .N / n_replicates),
  by = .(param_set, id_lulc_ant = id_lulc_initial, id_lulc_post = id_lulc_observed)
]

fom_by_trans <- Reduce(
  function(x, y) merge(x, y, by = c("param_set", "id_lulc_ant", "id_lulc_post"), all = TRUE),
  list(
    flow[, .(param_set, id_lulc_ant, id_lulc_post, obs_t = obs)],
    sim_t,
    hit_t
  )
)
for (j in c("obs_t", "sim_t", "hit_t")) set(fom_by_trans, which(is.na(fom_by_trans[[j]])), j, 0)

fom_by_trans[, fom := hit_t / (obs_t + sim_t - hit_t)]
fom_by_trans <- null_by_trans[
  , .(param_set, id_lulc_ant, id_lulc_post, fom_null)
][fom_by_trans, on = c("param_set", "id_lulc_ant", "id_lulc_post")]
fom_by_trans[, at_chance := is.finite(fom) & is.finite(fom_null) & fom <= fom_null]

fom_by_trans[order(param_set, fom)][, .(
  param_set, id_lulc_ant, id_lulc_post,
  obs_t = round(obs_t), sim_t = round(sim_t),
  fom = round(fom, 3), fom_null = round(fom_null, 3), at_chance
)]
```

## (c3) Probabilistic cross-tabulation over the replicate ensemble

A single realisation's cross-tabulation is one draw from the allocator. Averaging the per-cell
class **frequency** across the replicates of a parameter set turns the hard cross-tabulation
into a soft one whose entries are expected cell counts, and the FoM computed from it is the
ensemble expectation rather than one sample of it. This is the cross-tabulation "weighted by
probabilistic allocation" the handover asks for; the weights come from the ensemble, so
`n_replicates` is what controls their resolution.

```{r}
#| label: backcast-ensemble
ens <- cells[, .(p = .N / n_replicates), by = .(param_set, id_coord, id_lulc_simulated)]
ens <- observed[ens, on = "id_coord"]

soft_crosstab <- ens[, .(expected_cells = sum(p)), by = .(
  param_set, id_lulc_observed, id_lulc_simulated
)]

dcast(
  soft_crosstab[param_set == min(param_sets)],
  id_lulc_observed ~ id_lulc_simulated,
  value.var = "expected_cells",
  fill = 0
)
```

```{r}
#| label: backcast-ensemble-fom
soft_fom <- ens[, .(
  hit = sum(p[id_lulc_observed != id_lulc_initial & id_lulc_simulated == id_lulc_observed]),
  miss = sum(p[id_lulc_observed != id_lulc_initial & id_lulc_simulated == id_lulc_initial]),
  wrong_class = sum(p[
    id_lulc_observed != id_lulc_initial &
      id_lulc_simulated != id_lulc_initial &
      id_lulc_simulated != id_lulc_observed
  ]),
  false_alarm = sum(p[id_lulc_observed == id_lulc_initial & id_lulc_simulated != id_lulc_initial])
), by = param_set][, fom_expected := hit / (hit + miss + wrong_class + false_alarm)][]

soft_fom[order(-fom_expected)]
```

## Acceptance criteria

Three, in the order they should be read. The first is a hard gate on the machinery; the second
is the substantive one; the third governs how the results may be reported.

1. **Quantity fidelity.** For every viable transition with `count_prescribed > 1000`, the
   replicate-mean shortfall must be below 5 %. A larger shortfall means the allocator could not
   deliver a demand taken straight from the observed record, and nothing downstream —
   least of all `09`, where the LP pushes rates *above* that record — can be trusted until it is
   understood.
2. **Allocation skill.** The ensemble FoM must beat the random-within-class null by a factor of
   two. This is a deliberately modest bar: published land-change models mostly land between a
   few per cent and ~15 % absolute FoM (Pontius et al. 2008, *Ann. Reg. Sci.* 42, comparing 13
   applications), and absolute FoM scales with how much change there is to get right, so the
   only meaningful comparison is against a null on this same landscape and period.
3. **Per-transition honesty.** Transitions flagged `at_chance` above must be reported as
   spatially uninformative. Their quantity is still prescribed and still lands, but their
   *pattern* is noise, and `09d` must not present maps of them as model output.

```{r}
#| label: backcast-acceptance
quantity_gate <- quantity[
  count_prescribed > 1000,
  .(mean_shortfall_pct = mean(shortfall_pct)),
  by = .(id_trans, id_period)
][mean_shortfall_pct > 5]

skill_gate <- merge(
  soft_fom[, .(param_set, fom_expected)],
  null_by_param,
  by = "param_set"
)[, ratio := fom_expected / fom_null][]

if (nrow(quantity_gate) > 0L) {
  warning(
    "Quantity gate failed for ", nrow(quantity_gate),
    " (transition, period) pairs - see quantity_gate"
  )
}
if (all(skill_gate$ratio < 2)) {
  warning(
    "No parameter set reaches 2x the random-allocation null (best ratio ",
    signif(max(skill_gate$ratio), 3),
    "). The predictor set, not the allocation parameters, is the thing to revisit."
  )
}
list(quantity_gate = quantity_gate, skill_gate = skill_gate)
```

## Select the parameter set and promote it

The winner is chosen on the ensemble FoM, with `similarity_change` as the tie-break: FoM answers
"are the right cells changing", the fuzzy similarity answers "is the change the right shape",
and only the first is comparable against a null.

The chosen set is written to run `0` so that every SSP run in `09` inherits it through the
lineage without further plumbing.

```{r}
#| label: backcast-select
param_score <- Reduce(
  function(x, y) merge(x, y, by = "param_set"),
  list(
    soft_fom[, .(param_set, fom_expected)],
    null_by_param,
    fuzzy_by_param[, .(similarity_change = mean(similarity_change, na.rm = TRUE)), by = param_set]
  )
)[order(-fom_expected, -similarity_change)]

best_param <- param_score$param_set[1]
message("Selected parameter set ", best_param)
param_score
```

```{r}
#| label: backcast-commit-params
# Record each set's measured similarity alongside its parameters (the `similarity` column
# exists for exactly this and is what eval_alloc_params_t() would have filled).
for (p in param_sets) {
  db$alloc_params_t <- as_alloc_params_t(merge(
    copy(alloc_params[id_run == p])[, id_run := param_run_id(p)][, -"similarity"],
    fuzzy_by_param[param_set == p, .(id_trans, similarity = similarity_change)],
    by = "id_trans",
    all.x = TRUE
  ))
}

db$id_run <- 0L
db$alloc_params_t <- as_alloc_params_t(merge(
  copy(alloc_params[id_run == best_param])[, id_run := 0L][, -"similarity"],
  fuzzy_by_param[param_set == best_param, .(id_trans, similarity = similarity_change)],
  by = "id_trans",
  all.x = TRUE
))
db$alloc_params_t
```

> **Why not `eval_alloc_params_t()`?** The upstream helper does exactly this loop —
> allocate the observed periods per parameter set, then score with
> `calc_transition_similarity()` — but it calls `alloc_dinamica()`, so it needs a
> DinamicaConsole installation, it does not use the `runs_t` hierarchy for the parameter
> sets, it runs one realisation per set (no ensemble), and it reports only the unmasked
> similarity. Folding the CLUMPY backend and an `n_replicates` argument into it would let
> this step collapse back to one call.

```{r}
#| label: backcast-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key = , value = ,
  "backcast_periods", paste(backcast_periods, collapse = ","),
  "backcast_initial_period", as.character(id_period_initial),
  "backcast_n_param_sets", as.character(length(param_sets)),
  "backcast_n_replicates", as.character(n_replicates),
  "backcast_perturb_sd", as.character(perturb_sd),
  "backcast_fuzzy_window", as.character(fuzzy_window),
  "backcast_selected_param_set", as.character(best_param),
  "backcast_fom_expected", as.character(signif(param_score$fom_expected[1], 4)),
  "backcast_fom_null", as.character(signif(param_score$fom_null[1], 4)),
  "backcast_similarity_change", as.character(signif(param_score$similarity_change[1], 4)),
  "backcast_unmodellable_pct", as.character(signif(max(unmodellable$unmodellable_pct), 3))
))
```
