---
title: "080 — Backcasting validation"
---

Validate the calibrated model against the past: replay the **observed** transition rates
through the fitted transition-potential models and the estimated allocation parameters, then
compare the resulting maps against the Arealstatistik surveys we actually observed, substeps:

(a) **estimate patch / allocation parameters** from the observed periods
(b) **run the backcast** over a set of parameter choices, several stochastic replicates each;
(c) **validate** the backcast against the observed periods, and define acceptance criteria.

## What this does and does not test

The rates fed to the backcast are the **observed** ones (`get_obs_trans_rates()`), so the
*quantity* of change per transition and period is prescribed, not predicted. What is on trial
is therefore only the **allocation** — whether the model puts the right changes in the
right places — plus the allocator's ability to deliver the prescribed quantity at all.

Two consequences worth stating before any number is read:

- **Non-viable transitions bound the score from above.** Observed change belonging to a
  transition that `040`/`060` dropped can never be reproduced, because `predict_trans_pot()` only
  predicts for viable transitions and `adjusted_trans_pot_v()` inner-joins potentials to rates.
  Rows for non-viable transitions in `trans_rates_t` are inert. The `unmodellable` share below
  quantifies that ceiling.
- **Persistence is not skill.** The great majority of Swiss hectares keep their class across a
  survey interval — `040d`'s observed-transition plot has the exact figure — so any whole-map
  agreement statistic is dominated by cells nobody had to get right. Every metric here is
  therefore computed on **change**, not on the map.

## Parameters

`n_perturbations` × `n_replicates` chained backcasts, each covering three period steps, is the
cost driver: every allocated period re-runs `predict_trans_pot()` over the full 4.1 M-cell grid
for every viable transition, because `alloc_clumpy()` predicts inside the same call that
allocates. Start small and grow the grid once the runtime is known.

```{r}
#| label: backcast-params
library(data.table)
library(evoland)

n_perturbations <- 3L # perturbed parameter sets *in addition to* the best estimate
perturb_sd <- 0.05 # sd of the frac_expander perturbation (evoland default)
n_replicates <- 3L # stochastic allocations per parameter set
seed <- 42L
fuzzy_window <- 11L # odd; evoland's default, ~1 km at 100 m resolution
select_score <- "classif.auc" # must match 06
select_maximize <- TRUE
id_run_backcast_base <- 2000L # 2000-2999 is free: 001 uses 0/100/1xxx, 071 uses max+n

db <- evoland_db$new(path = "ssp-ch.evolanddb")
db$id_run <- 0L

# evoland's table classes carry print methods that assume their own columns; a summary derived
# from one inherits the class and prints nonsense ("Rate range: [Inf, -Inf]"). Strip it.
plain <- function(x) {
  data.table::setattr(data.table::copy(x), "class", c("data.table", "data.frame"))[]
}

obs_periods <- db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period]
id_period_initial <- min(obs_periods) # the backcast's starting state, never simulated
backcast_periods <- sort(setdiff(obs_periods, id_period_initial))
id_period_final <- max(backcast_periods)

stopifnot(
  "need at least two observed periods" = length(backcast_periods) >= 1L,
  "backcast periods must be contiguous" = all(diff(backcast_periods) == 1L),
  "no viable transitions - run 040 (and 060, which reconciles viability)" =
    nrow(db$trans_meta_t[is_viable == TRUE]) > 0L,
  "no fitted models - run 060-transition-modelling.qmd first" =
    db$row_count("trans_models_t") > 0L
)

message(
  "Backcasting periods ", paste(backcast_periods, collapse = " -> "),
  " from observed period ", id_period_initial
)
```

## Observed rates as the backcast demand

Written to run `0`: they are a property of the observed record, not of a scenario, and every
backcast run inherits them through the run lineage. The scenario rates `070` wrote sit on the SSP
leaf runs at the *extrapolated* periods, so the two never resolve to the same
`(id_period, id_trans)` slice.

The unmodellable cells are from deglaciation and relatively rare transitions, respectively. We will be deterministically overwriting the glacier transitions, because that's also how the extrapolation is done.

```{r}
#| label: backcast-obs-rates
obs_rates <- db$get_obs_trans_rates()[id_period %in% backcast_periods]
db$trans_rates_t <- obs_rates

# How much observed change is out of reach because its transition is not viable?
viable_trans <- db$trans_meta_t[is_viable == TRUE]
unmodellable <- plain(obs_rates)[
  db$trans_meta_t[
    db$lulc_meta_t[, .(id_lulc, name_anterior = name)],
    on = c(id_lulc_anterior = "id_lulc")
  ][
    db$lulc_meta_t[, .(id_lulc, name_posterior = name)],
    on = c(id_lulc_posterior = "id_lulc")
  ],
  on = "id_trans",
  nomatch = NULL
][
  ,
  .(
    name_anterior,
    name_posterior,
    cells = sum(count),
    cells_viable = sum(count[is_viable == TRUE]),
    is_viable
  ),
  by = .(id_period, id_trans)
][, `:=`(
  unmodellable_pct = round(100 * (1 - cells_viable / cells), 2)
)][]
unmodellable[is_viable == FALSE & cells > 1000]
```

## (a) Allocation parameters

`create_alloc_params_t()` measures, per viable transition and per observed period pair, the mean
and variance of new-patch area, patch elongation, and the split between cells that **expand** an
existing patch of the target class and cells that start a **new** patch; it then averages over
the period pairs and returns that best estimate. The `n_perturbations` further sets are built
here: each jitters `frac_expander` by a normal draw with sd `perturb_sd`, clamped to [0, 1], with
`frac_patcher` following as `1 - frac_expander`.

Only `frac_expander` is perturbed, so this sweep explores the expander/patcher split and
nothing else. That is the parameter the SSP interventions also target, which makes it the
defensible one to sweep first, but it means a poor `mean_patch_size` estimate will not show up
as parameter sensitivity here. Widening the sweep is a change to `perturb_alloc_params()` only.

```{r}
#| label: backcast-alloc-params
db$id_run <- 0L
best_estimate <- db$create_alloc_params_t()

perturb_alloc_params <- function(params, sd) {
  jittered <- pmin(1, pmax(0, params$frac_expander + rnorm(nrow(params), sd = sd)))
  copy(params)[, `:=`(frac_expander = jittered, frac_patcher = 1 - jittered)]
}

# id_run here is a 1-based parameter-set index, not a real run; set 1 is the best estimate.
set.seed(seed)
alloc_params <- rbindlist(c(
  list(copy(best_estimate)[, id_run := 1L]),
  lapply(seq_len(n_perturbations), function(i) {
    perturb_alloc_params(best_estimate, perturb_sd)[, id_run := 1L + i]
  })
))
param_sets <- sort(unique(alloc_params$id_run))
plain(alloc_params)[, .(
  n_trans = .N,
  mean_patch_size = round(mean(mean_patch_size, na.rm = TRUE), 2),
  frac_expander = round(mean(frac_expander, na.rm = TRUE), 3)
), by = id_run]
```

A transition that never produced a new patch (all conversions expanded existing ones) comes back
with `patch_elongation = NA`; `alloc_clumpy()` substitutes `0.5`, and `mean_patch_size = 0` is
handled as a mono-pixel draw in the C++ patcher. Both are safe, but a transition in that state
has no estimated patch geometry at all — worth knowing before reading its similarity.

```{r}
#| label: backcast-alloc-params-gaps
plain(alloc_params)[id_run == min(param_sets) & (is.na(patch_elongation) | mean_patch_size <= 1)][
  db$trans_meta_t,
  on = "id_trans",
  nomatch = NULL
][, .(id_trans, id_lulc_anterior, id_lulc_posterior, mean_patch_size, patch_elongation)]
```

## Run registration

One parent run per parameter set holding the `alloc_params_t` rows, and one child run per
stochastic replicate holding only the simulated `lulc_data_t`. The replicates inherit the
parameters through the lineage, so — as in the evoland stochastic-allocation vignette — the only
thing that differs between replicates of a set is the RNG state.

```{r}
#| label: backcast-register-runs
param_run_id <- function(p) id_run_backcast_base + 100L * as.integer(p)
member_run_id <- function(p, r) param_run_id(p) + as.integer(r)

param_runs <- data.table(
  id_run = param_run_id(param_sets),
  parent_id_run = 0L,
  description = paste0(
    "Backcasting allocation parameter set ", param_sets,
    fifelse(param_sets == min(param_sets), " (unperturbed estimate)", " (perturbed)")
  ),
  kind = "backcast parameters"
)

member_runs <- CJ(param_set = param_sets, replicate = seq_len(n_replicates))[, .(
  id_run = member_run_id(param_set, replicate),
  parent_id_run = param_run_id(param_set),
  description = paste0("Backcast replicate ", replicate, " of parameter set ", param_set),
  kind = "backcast replicate",
  param_set = param_set,
  replicate = replicate,
  seed = seed + 1000L * param_set + replicate
)]

# Re-running this step must replace its own runs, not collide with them; anything else in the
# 2000-block is somebody else's and is a hard error.
backcast_kinds <- c("backcast parameters", "backcast replicate")
existing_runs <- db$runs_t
mine <- if ("kind" %in% names(existing_runs)) {
  existing_runs[kind %in% backcast_kinds, id_run]
} else {
  integer(0)
}
clash <- setdiff(
  intersect(c(param_runs$id_run, member_runs$id_run), existing_runs$id_run),
  mine
)
stopifnot("backcast run ids collide with runs registered by 001/071" = length(clash) == 0L)

if (length(mine) > 0L) {
  message(
    "Replacing ", length(mine), " backcast runs from an earlier execution; ",
    "their lulc_data_t / trans_pot_t rows are overwritten in place."
  )
}

db$commit(
  as_runs_t(rbind(
    existing_runs[!id_run %in% c(param_runs$id_run, member_runs$id_run)],
    param_runs,
    member_runs,
    fill = TRUE
  )),
  "runs_t",
  "overwrite" # adding columns
)

for (p in param_sets) {
  db$alloc_params_t <- as_alloc_params_t(
    copy(alloc_params[id_run == p])[, id_run := param_run_id(p)]
  )
}

# --- interventions (not implemented here) ----------------------------------------------
# The SSP `Pre-allocation` interventions are perturbations of exactly this table. Each
# NCCS-SSP-scenarios/Tools/SSP*_interventions.yml opens with a `Spatial_zoning` block whose
# `Param_adjust_name` is one of Mean_Patch_Size / Patch_Size_Variance / Patch_Isometry /
# Perc_expander / Perc_patcher -- i.e. the alloc_params_t columns estimated above -- applied to
# the transitions targeting one class (`Transition_target_classes: Urban` in all five files).
# To realise them, write a modified copy of the winning parameter rows onto the SSP run instead
# of letting it inherit run 0's:
#
#   mod <- copy(alloc_params[id_run == best_param])[
#     id_trans %in% urban_trans, frac_patcher := 0.15][      # SSP1 Absolute 0.15
#     id_trans %in% urban_trans, frac_expander := 1 - frac_patcher]
#   db$alloc_params_t <- as_alloc_params_t(mod[, id_run := ssp_run_id])
#
# Two conversions are needed and are not free:
#   * `Absolute` replaces the estimated value (SSP0/3/5 set Perc_patcher to 0, SSP1 to 0.15);
#     `Relative` (only SSP4, 0.5) scales it, but the YAML header does not say what it scales.
#     Whatever it is, the numbers were tuned against the *original's* parameter estimates:
#     Spatial_intervention_updates.txt records SSP1's 0.20 target being cut to 0.15 "because
#     some of the transitions were already close". They do not transfer unexamined to ours.
#   * `Patch_Isometry` is a Dinamica parameter. alloc_clumpy consumes `patch_elongation`
#     (alloc_params_clumpy_v maps it straight through), so an isometry intervention has to be
#     inverted back through isometry_from_elongation(). That mapping is a 1-D interpolation of
#     Mazy fig. 3.10 and is not injective over its flat segments.
```

## (b) Run the backcast

Each replicate starts from the observed `id_period_initial` map — which the run lineage resolves
to run `0`, since the replicate writes nothing for that period — and then chains: the simulated
period 2 map is the anterior state for period 3, and so on. The error therefore **compounds**,
which is the point: a backcast across three decades of surveys tests the trajectory, not one
step.

> An alternative design, noted in the handover, is to branch *N* members per allocation step from
> the observed anterior state instead of from the previous simulation. That isolates one-step
> allocation skill and is strictly cheaper to interpret. It needs no new machinery here: register
> the members under the same parent and allocate a **single** period each
> (`id_periods = k`), writing only period `k`; the lineage then resolves `k - 1` to run `0`'s
> observed map. Both designs are worth having; the chained one is the one a backcast is usually
> asked for, so it is the one implemented.

### Deterministic glacier change

Glacier transitions are not modelled (`040` excludes glacier as an anterior class): their
extent follows the climate, not the land use drivers, so they are imposed on the simulated map
after each period is allocated, before the next period is allocated from it. In the backcast
the imposed change is the **observed** one: every cell that turned into or out of glacier
between two surveys takes its observed class. In the extrapolation (`090`) the same step will
take its assignments from GloGEM glacier model output instead; `apply_deterministic_change()`
only needs a table of `id_coord` and the class to impose.

```{r}
#| label: backcast-glacier
id_lulc_glacier <- db$lulc_meta_t[name == "glacier", id_lulc]

db$id_run <- 0L
glacier_change <- db$get_query(
  "select post.id_period, post.id_coord, post.id_lulc
  from {lulc_read_expr} ant
  inner join {lulc_read_expr} post
    on post.id_coord = ant.id_coord and post.id_period = ant.id_period + 1
  where post.id_period in ({backcast_periods*})
    and ant.id_lulc != post.id_lulc
    and (ant.id_lulc in ({id_lulc_glacier*}) or post.id_lulc in ({id_lulc_glacier*}))",
  lulc_read_expr = db$get_read_expr("lulc_data_t")
)

# impose `assignments` (id_coord, id_lulc) on an allocated lulc_data_t, by reference
apply_deterministic_change <- function(lulc, assignments) {
  lulc[assignments, on = "id_coord", id_lulc := i.id_lulc]
}

glacier_change[, .N, by = id_period]
```

### Allocation

Every replicate is allocated one period at a time with `alloc_clumpy_one_period()`, so the
glacier change can be imposed between periods; the multi-period `alloc_clumpy()` has no such
hook. Each replicate is seeded with its own registered seed: CLUMPY draws through R's RNG, so
`set.seed()` makes each replicate reproducible and the replicates independent of each other.

The transition potentials for the *first* backcast period are identical across every replicate
of a parameter set (same observed anterior state, same models, same predictors), so the first
period reads them from the parameter-set run with `use_parent_trans_pot = TRUE`: the first
replicate predicts them onto the parent and its siblings reuse them. Later periods start from
each replicate's own simulated map and are predicted per replicate.

```{r}
#| label: backcast-allocate
#| output: false
replicate_runs <- db$runs_t[kind == "backcast replicate", .(id_run, seed)]
options("evoland.use_prefetch_predict" = TRUE)
future::plan(future.mirai::mirai_multisession, workers = 20)
for (i in seq_len(nrow(replicate_runs))) {
  db$id_run <- replicate_runs$id_run[i]
  set.seed(replicate_runs$seed[i])

  for (id_period_post in backcast_periods) {
    lulc_simulated <- alloc_clumpy_one_period(
      db = db,
      id_period_post = id_period_post,
      select_score = select_score,
      select_maximize = select_maximize,
      use_parent_trans_pot = id_period_post == min(backcast_periods)
    )
    apply_deterministic_change(
      lulc_simulated,
      glacier_change[id_period == id_period_post, .(id_coord, id_lulc)]
    )
    db$commit(lulc_simulated, "lulc_data_t", method = "upsert")
    # the next period's prediction reads the neighbour predictors of this one
    if (id_period_post != max(backcast_periods)) {
      db$upsert_new_neighbors(id_period_post)
    }
  }
}
future::plan("sequential")
db$id_run <- 0L
```

Each replicate still writes its own `trans_pot_t` slice for every period after the first (one
value per cell, viable transition and period), which is by far the largest thing the backcast
stores.

## Simulated end states

The backcast writes each allocated run's maps to `lulc_data_t`; the validation below reads them
from there. Here we only check that every allocated run covers the observed grid.

```{r}
#| label: backcast-maps
allocated_runs <- db$runs_t[kind == "backcast replicate", .(id_run)]
run_param_set <- member_runs[, .(id_run, param_set)] # parameter set of each replicate

db$id_run <- NULL # the lineage-free read expression spans all runs
simulated_cells <- db$get_query(
  "select id_run, count(*)::integer as n_cells from {lulc_read_expr}
  where id_run in ({allocated_runs$id_run*}) and id_period = {id_period_final}
  group by id_run",
  lulc_read_expr = db$get_read_expr("lulc_data_t")
)
db$id_run <- 0L
n_observed_cells <- db$get_query(
  "select count(*)::integer from {lulc_read_expr} where id_period = {id_period_final}",
  lulc_read_expr = db$get_read_expr("lulc_data_t"),
  as_atomic = TRUE
)

stopifnot(
  "no simulated cells - did allocation write anything?" = nrow(simulated_cells) > 0L,
  "an allocated run is missing cells" =
    nrow(simulated_cells) == nrow(allocated_runs) &&
      all(simulated_cells$n_cells == n_observed_cells)
)
simulated_cells
```

## Quantity check: did the allocator deliver the prescribed change?

The first thing to establish is that the allocator hit the demand it was given. uPAM draws
patches without replacement from the source pool; when the prescribed rate is large relative to
the pool, or `avoid_aggregation` rejects many candidate patches, the pool empties before the
quota is met and the run silently under-delivers. `070` flags the same failure mode for the
scenario rates, where it is worse because the LP pushes rates above the historic envelope.

```{r}
#| label: backcast-quantity
realised_counts <- rbindlist(lapply(allocated_runs$id_run, function(r) {
  db$id_run <- r
  db$get_obs_trans_rates()[
    id_period %in% backcast_periods,
    .(id_run = r, id_period, id_trans, count_realised = count)
  ]
}))
db$id_run <- 0L

quantity <- merge(
  CJ(
    id_run = allocated_runs$id_run,
    id_trans = viable_trans$id_trans,
    id_period = backcast_periods
  )[
    obs_rates[, .(id_period, id_trans, count_prescribed = count)],
    on = c("id_period", "id_trans"),
    nomatch = NULL
  ],
  realised_counts,
  by = c("id_run", "id_period", "id_trans"),
  all.x = TRUE
)
quantity[is.na(count_realised), count_realised := 0L]
quantity[, shortfall_pct := 100 * (count_prescribed - count_realised) / count_prescribed]

quantity[order(-shortfall_pct)][1:15][
  db$trans_meta_t,
  on = "id_trans",
  nomatch = NULL
][, .(
  id_run, id_period, id_lulc_anterior, id_lulc_posterior,
  count_prescribed, count_realised,
  shortfall_pct = round(shortfall_pct, 1)
)]
```

## (c1) Fuzzy similarity of differences

`calc_transition_similarity()` is evoland's port of Dinamica's "similarity of differences"
(Hagen 2003): for one transition it builds the observed and simulated **change** maps, and gives
each changed cell a membership in the other map's change that decays with the distance to the
nearest change there (`exp(-d / 2)` within the window, 1 in the same cell), so a patch placed one
cell off still scores. Each direction is averaged over that map's own changed cells only, so the
unchanged background does not enter; `similarity` is the minimum of the two directions, so a
simulation cannot score well by placing much more or much less change than was observed.

```{r}
#| label: backcast-fuzzy
db$id_run <- 0L
rast_initial <- db$lulc_data_as_rast(id_period = id_period_initial)
rast_obs_final <- db$lulc_data_as_rast(id_period = id_period_final)

fuzzy_one_run <- function(id_run_member) {
  db$id_run <- id_run_member
  rast_sim <- db$lulc_data_as_rast(id_period = id_period_final)

  res <- mapply(
    FUN = calc_transition_similarity,
    from_class = viable_trans$id_lulc_anterior,
    to_class = viable_trans$id_lulc_posterior,
    MoreArgs = list(
      initial_map = rast_initial,
      observed_map = rast_obs_final,
      simulated_map = rast_sim,
      window_size = fuzzy_window,
      use_exp_decay = TRUE,
      decay_divisor = 2.0
    ),
    SIMPLIFY = FALSE
  )

  data.table(
    id_run = id_run_member,
    id_trans = viable_trans$id_trans,
    similarity = vapply(res, \(x) as.numeric(x$similarity), numeric(1L)),
    n_observed = vapply(res, \(x) as.numeric(x$n_observed), numeric(1L)),
    n_simulated = vapply(res, \(x) as.numeric(x$n_simulated), numeric(1L))
  )
}

fuzzy_dt <- rbindlist(lapply(allocated_runs$id_run, fuzzy_one_run))

db$id_run <- 0L

fuzzy_by_param <- run_param_set[fuzzy_dt, on = "id_run"][, .(
  similarity_sd = sd(similarity, na.rm = TRUE),
  similarity = mean(similarity, na.rm = TRUE),
  n_observed = mean(n_observed),
  n_simulated = mean(n_simulated)
), by = .(param_set, id_trans)]

fuzzy_by_param[db$trans_meta_t, on = "id_trans", nomatch = NULL][
  order(param_set, similarity)
][, .(
  param_set, id_lulc_anterior, id_lulc_posterior,
  similarity = round(similarity, 3),
  n_observed, n_simulated = round(n_simulated)
)]
```

## (c2) Cross-tabulation and figure of merit

The cross-tabulation of observed against simulated end class, restricted to the cells that
*should* have changed, decomposes into the four quantities of Pontius' figure of merit:

| | simulated persistence | simulated a change |
| --- | --- | --- |
| **observed a change to _v_** | miss (A) | **hit (B)** if simulated to _v_, wrong class (C) otherwise |
| **observed persistence** | correct null | false alarm (D) |

$\mathrm{FoM} = B / (A + B + C + D)$ — the intersection of observed and simulated change over
their union. It ignores the correct-null cell, which is why it is the right statistic here.

`db$figure_of_merit_v()` computes it in the database: the initial and observed maps from run
`0`, each allocated run's simulated map through its own lineage, one row per run, and with
`by_transition = TRUE` one row per run and transition.

**Glacier cells are excluded.** Their change is imposed from the record (see (b)), so a cell
that starts or ends as glacier is a hit by construction and says nothing about the model;
`exclude_id_lulc` leaves out every cell whose initial or observed class is glacier.

```{r}
#| label: backcast-fom
fom_args <- list(
  id_period_anterior = id_period_initial,
  id_period_post = id_period_final,
  id_run_reference = 0L,
  id_run_simulated = allocated_runs$id_run,
  exclude_id_lulc = id_lulc_glacier
)
fom_by_run <- do.call(db$figure_of_merit_v, fom_args)
fom_by_trans_run <- do.call(db$figure_of_merit_v, c(fom_args, by_transition = TRUE))

fom_by_run[order(-figure_of_merit), .(
  id_run, hits, wrong_hits, misses, false_alarms,
  figure_of_merit = round(figure_of_merit, 4),
  figure_of_merit_null = round(figure_of_merit_null, 4),
  producers_accuracy = round(producers_accuracy, 3),
  users_accuracy = round(users_accuracy, 3)
)]
```

The null — the FoM of a model that gets the quantity right but the location purely by chance —
is available in closed form, which is cheaper and less arbitrary than running a
scrambled-potential control. Within an anterior class $u$ of $N_u$ cells, of which the model
converts $S_{uv}$ to $v$ and the record converts $O_{uv}$, random placement gives
$\mathbb{E}[B_{uv}] = S_{uv} O_{uv} / N_u$, and the expected union of change over the class is
$S_u + O_u - S_u O_u / N_u$; `figure_of_merit_v()` returns it as `figure_of_merit_null`.

Per-transition FoM uses the same decomposition, so it is directly comparable with its null:
hits are cells the record and the model both moved from `u` to `v`, and the denominator is the
union of the two flows.

```{r}
#| label: backcast-fom-by-trans
fom_by_trans <- run_param_set[fom_by_trans_run, on = "id_run"][
  ,
  .(
    observed = mean(observed),
    simulated = mean(simulated),
    hits = mean(hits),
    figure_of_merit_null = mean(figure_of_merit_null)
  ),
  by = .(param_set, id_lulc_anterior, id_lulc_posterior)
][, figure_of_merit := hits / (observed + simulated - hits)][, at_chance := is.finite(figure_of_merit) & is.finite(figure_of_merit_null) &
  figure_of_merit <= figure_of_merit_null]

fom_by_trans[order(param_set, figure_of_merit)][, .(
  param_set, id_lulc_anterior, id_lulc_posterior,
  observed = round(observed), simulated = round(simulated),
  figure_of_merit = round(figure_of_merit, 3),
  figure_of_merit_null = round(figure_of_merit_null, 3),
  at_chance
)]
```

## (c3) Expected figure of merit over the replicate ensemble

A single realisation is one draw from the allocator. Averaging the counts over the
`n_replicates` replicates of a parameter set gives the expected hits, misses, wrong hits and
false alarms, and the FoM computed from those is the ensemble expectation rather than one
sample of it (the mean of the per-run ratios would not be). The null is averaged over the
replicates as it is: it depends on the simulated quantities only, which barely differ between
replicates of one parameter set.

```{r}
#| label: backcast-ensemble-fom
ensemble_fom <- run_param_set[fom_by_run, on = "id_run"][, .(
  n_runs = .N,
  hits = mean(hits),
  wrong_hits = mean(wrong_hits),
  misses = mean(misses),
  false_alarms = mean(false_alarms),
  fom_null = mean(figure_of_merit_null)
), by = param_set][, fom_expected := hits / (hits + wrong_hits + misses + false_alarms)]

ensemble_fom[order(-fom_expected)]
```

## Acceptance criteria

Three, in the order they should be read. The first is a hard gate on the machinery; the second
is the substantive one; the third governs how the results may be reported.

1. **Quantity fidelity.** For every viable transition with `count_prescribed > 1000`, the
   replicate-mean shortfall must be below 5 %. A larger shortfall means the allocator could not
   deliver a demand taken straight from the observed record, and nothing downstream —
   least of all `090`, where the LP pushes rates *above* that record — can be trusted until it is
   understood.
2. **Allocation skill.** The ensemble FoM must beat the random-within-class null by a factor of
   two. This is a deliberately modest bar: published land-change models mostly land between a
   few per cent and ~15 % absolute FoM (Pontius et al. 2008, *Ann. Reg. Sci.* 42, comparing 13
   applications), and absolute FoM scales with how much change there is to get right, so the
   only meaningful comparison is against a null on this same landscape and period.
3. **Per-transition honesty.** Transitions flagged `at_chance` above must be reported as
   spatially uninformative. Their quantity is still prescribed and still lands, but their
   *pattern* is noise, and `090d` must not present maps of them as model output.

```{r}
#| label: backcast-acceptance
quantity_gate <- quantity[
  count_prescribed > 1000,
  .(mean_shortfall_pct = mean(shortfall_pct)),
  by = .(id_trans, id_period)
][mean_shortfall_pct > 5]

skill_gate <- ensemble_fom[, .(param_set, fom_expected, fom_null, ratio = fom_expected / fom_null)]

if (nrow(quantity_gate) > 0L) {
  warning(
    "Quantity gate failed for ", nrow(quantity_gate),
    " (transition, period) pairs - see quantity_gate"
  )
}
if (all(skill_gate$ratio < 2)) {
  warning(
    "No parameter set reaches 2x the random-allocation null (best ratio ",
    signif(max(skill_gate$ratio), 3),
    "). The predictor set, not the allocation parameters, is the thing to revisit."
  )
}
list(quantity_gate = quantity_gate, skill_gate = skill_gate)
```

## Select the parameter set and promote it

The winner is chosen on the ensemble FoM, with the mean `similarity` as the tie-break: FoM answers
"are the right cells changing", the fuzzy similarity answers "is the change the right shape",
and only the first is comparable against a null.

The chosen set is written to run `0` so that every SSP run in `090` inherits it through the
lineage without further plumbing.

```{r}
#| label: backcast-select
param_score <- merge(
  ensemble_fom[, .(param_set, fom_expected, fom_null)],
  fuzzy_by_param[, .(similarity = mean(similarity, na.rm = TRUE)), by = param_set],
  by = "param_set"
)[order(-fom_expected, -similarity)]

best_param <- param_score$param_set[1]
message("Selected parameter set ", best_param)
param_score
```

```{r}
#| label: backcast-commit-params
# Record each set's measured similarity alongside its parameters (the `similarity` column
# exists for exactly this and is what eval_alloc_params_t() would have filled).
for (p in param_sets) {
  db$alloc_params_t <- as_alloc_params_t(merge(
    copy(alloc_params[id_run == p])[, id_run := param_run_id(p)][, -"similarity"],
    fuzzy_by_param[param_set == p, .(id_trans, similarity)],
    by = "id_trans",
    all.x = TRUE
  ))
}

db$id_run <- 0L
db$alloc_params_t <- as_alloc_params_t(merge(
  copy(alloc_params[id_run == best_param])[, id_run := 0L][, -"similarity"],
  fuzzy_by_param[param_set == best_param, .(id_trans, similarity)],
  by = "id_trans",
  all.x = TRUE
))
db$alloc_params_t
```

> **Why not `eval_alloc_params_t()`?** The upstream helper does exactly this loop —
> allocate the observed periods per parameter set, then score with
> `calc_transition_similarity()` — but it calls `alloc_dinamica()`, so it needs a
> DinamicaConsole installation, it does not use the `runs_t` hierarchy for the parameter
> sets, it runs one realisation per set (no ensemble), and it reports no figure of merit.
> Folding the CLUMPY backend and an `n_replicates` argument into it would let this step
> collapse back to one call.

```{r}
#| label: backcast-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key = , value = ,
  "backcast_periods", paste(backcast_periods, collapse = ","),
  "backcast_initial_period", as.character(id_period_initial),
  "backcast_n_param_sets", as.character(length(param_sets)),
  "backcast_n_replicates", as.character(n_replicates),
  "backcast_perturb_sd", as.character(perturb_sd),
  "backcast_fuzzy_window", as.character(fuzzy_window),
  "backcast_selected_param_set", as.character(best_param),
  "backcast_fom_expected", as.character(signif(param_score$fom_expected[1], 4)),
  "backcast_fom_null", as.character(signif(param_score$fom_null[1], 4)),
  "backcast_similarity", as.character(signif(param_score$similarity[1], 4)),
  "backcast_unmodellable_pct", as.character(signif(max(unmodellable$unmodellable_pct), 3))
))
```
