---
title: "09 — Stochastic extrapolation"
---

Run the forward projection: allocate the SSP transition rates `07` solved from the demand
targets, using the transition-potential models from `06` and the allocation parameters `08`
selected, over the extrapolated periods out to 2060.

Everything this step needs is already in the database. It registers one realisation run per
(scenario × replicate), calls the allocator, and checks that what came out matches what `07`
asked for. Reporting figures, maps and tables are `090d`'s job; this step mutates state and
verifies it.

**First iteration.** The `ssp_subset` / `climate_subset` parameters below restrict the run to a
couple of SSPs under one climate framing. The full cross registered in `001` is
5 SSPs × 5 climate framings = 25 leaves, and each leaf allocation is as expensive as one
backcast replicate, so the subset is a deliberate first pass rather than a limitation of the
code — widening it is a parameter change.

**No interventions.** The SSP narratives in `NCCS-SSP-scenarios` are realised not only by the
demand curves (which `07` carries) but also by spatial interventions per SSP. None of them are
applied here; the comments in the allocation section say what each kind would take.

## Parameters

```{r}
#| label: extrapolate-params
library(data.table)
library(evoland)

ssp_subset <- c("SSP1", "SSP3") # widen once the runtime per leaf is known
climate_subset <- "current" # "current" | "gwl"; see 001-setup-db.qmd
n_replicates <- 3L # stochastic realisations per scenario run
seed <- 42L
select_score <- "classif.auc" # must match 06 and 08
select_maximize <- TRUE

db <- evoland_db$new(path = "ssp-ch.evolanddb")
db$id_run <- 0L

# Summaries derived from an evoland table inherit its class and print with the wrong header
# ("Run metadata table" over an area table); strip it. Same helper as in 080.
plain <- function(x) {
  data.table::setattr(data.table::copy(x), "class", c("data.table", "data.frame"))[]
}

anchor_period <- max(db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period])
lulc_anchor <- db$lulc_data_t[id_period == anchor_period]

stopifnot(
  "no allocation parameters on run 0 - run 080-validate-backcasting.qmd first" =
    nrow(db$alloc_params_t) > 0L,
  "no viable transitions - run 04 and 06" =
    nrow(db$trans_meta_t[is_viable == TRUE]) > 0L,
  "no fitted models - run 06-transition-modelling.qmd first" =
    nrow(db$trans_models_t) > 0L
)
```

## Which runs are being allocated

`07` writes its solved rates to every leaf run of an SSP — the leaves differ by climate framing,
which changes the predictors but not the demand. Selecting a subset here therefore selects
(demand × climate) pairings, not demand alone.

```{r}
#| label: extrapolate-runs
runs <- db$runs_t
scenario_runs <- runs[
  !is.na(ssp) & ssp %in% ssp_subset & climate_framing %in% climate_subset
]

# Two families have to be filtered out again, and both are tested outside the `[` because
# data.table evaluates the whole `i` expression whether or not the column exists:
#   * realisation runs written by an earlier execution of this step inherit `ssp` and
#     `climate_framing` from their parent, but they are outputs, not scenarios;
#   * 071 registers legacy-rate siblings of every SSP run.
if ("kind" %in% names(scenario_runs)) {
  scenario_runs <- scenario_runs[is.na(kind) | kind != "extrapolation realisation"]
}
if ("rate_solver" %in% names(scenario_runs)) {
  scenario_runs <- scenario_runs[is.na(rate_solver) | rate_solver == "solver"]
}
scenario_runs <- scenario_runs[order(id_run)]

stopifnot(
  "no scenario runs match the subset - check ssp_subset / climate_subset" =
    nrow(scenario_runs) > 0L
)
scenario_runs[, .(id_run, parent_id_run, ssp, climate_framing, gwl_trajectory)]
```

## Which periods carry rates

The allocator walks contiguous periods, and every period it walks must have a rate for every
viable transition, or `adjusted_trans_pot_v()` silently drops that transition's potentials for
that period (it inner-joins potentials to rates) and the transition simply does not happen.
`07` solves up to the period containing 2060, so the sequence normally ends there.

```{r}
#| label: extrapolate-periods
db$id_run <- NULL # fetch() is lineage-filtered; NULL reads across runs
rates_available <- db$fetch(
  "trans_rates_t",
  where = glue::glue("id_run in ({toString(scenario_runs$id_run)})")
)
db$id_run <- 0L

extrap_periods <- sort(db$periods_t[is_extrapolated == TRUE, id_period])
viable_trans <- db$trans_meta_t[is_viable == TRUE]

coverage <- rates_available[
  id_period %in% extrap_periods & id_trans %in% viable_trans$id_trans,
  .(n_trans = uniqueN(id_trans)),
  by = .(id_run, id_period)
]
alloc_periods <- sort(unique(coverage$id_period))

stopifnot(
  "no extrapolated rates - run 070-transition-rates-solver.qmd first" =
    length(alloc_periods) > 0L,
  "rate periods are not contiguous" = all(diff(alloc_periods) == 1L),
  "rates do not start at the first extrapolated period" =
    min(alloc_periods) == min(extrap_periods),
  "not every selected run carries rates for every period" =
    nrow(coverage) == nrow(scenario_runs) * length(alloc_periods)
)

incomplete <- coverage[n_trans < nrow(viable_trans)]
if (nrow(incomplete) > 0L) {
  warning(
    "Rate table covers fewer than the ", nrow(viable_trans),
    " viable transitions for ", nrow(incomplete),
    " (run, period) pairs; those transitions cannot be allocated there."
  )
}

message("Allocating periods ", paste(alloc_periods, collapse = " -> "))
dcast(coverage, id_run ~ id_period, value.var = "n_trans")
```

## Register the realisation runs

Each realisation is a child of its scenario run, so it inherits the rates, the projected
predictors and the allocation parameters, and differs only in RNG state — the same pattern `08`
uses for its replicates and the evoland stochastic-allocation vignette uses for its ensemble.
Keeping the realisations off the scenario run also means re-running this step with a different
`n_replicates` never overwrites an earlier ensemble member.

```{r}
#| label: extrapolate-register
realisation_runs <- scenario_runs[, .(replicate = seq_len(n_replicates)), by = .(
  parent = id_run, ssp, climate_framing
)][, .(
  id_run = parent * 100L + replicate,
  parent_id_run = parent,
  description = paste0(ssp, " / ", climate_framing, " realisation ", replicate),
  kind = "extrapolation realisation",
  ssp,
  climate_framing,
  replicate,
  seed = seed + replicate
)]

# As in 080: re-running replaces this step's own realisations, but never anything else.
mine <- if ("kind" %in% names(runs)) {
  runs[kind == "extrapolation realisation", id_run]
} else {
  integer(0)
}
clash <- setdiff(intersect(realisation_runs$id_run, runs$id_run), mine)
stopifnot(
  "realisation run ids collide with runs registered elsewhere" = length(clash) == 0L,
  "n_replicates must be below 100 for the id_run scheme" = n_replicates < 100L
)

db$commit(
  as_runs_t(rbind(
    runs[!id_run %in% realisation_runs$id_run],
    realisation_runs,
    fill = TRUE
  )),
  "runs_t",
  "overwrite" # adding columns
)
realisations <- plain(realisation_runs)[, .(id_run, parent_id_run, ssp, replicate)]
realisation_runs[, .(id_run, parent_id_run, ssp, replicate, seed)]
```

## Allocate

```{r}
#| label: extrapolate-allocate
#| output: false
for (i in seq_len(nrow(realisation_runs))) {
  db$id_run <- realisation_runs$id_run[i]

  # db$alloc_clumpy() advertises a `seed` argument that alloc_clumpy() does not accept at the
  # pinned commit; see the note in 080. CLUMPY draws through R's RNG, so this is equivalent.
  set.seed(realisation_runs$seed[i])

  db$alloc_clumpy(
    id_periods = alloc_periods,
    select_score = select_score,
    select_maximize = select_maximize
  )
}
db$id_run <- 0L
```

<!--
--- interventions (not implemented) -------------------------------------------------------
The SSP interventions come in three stages (see the header of any
NCCS-SSP-scenarios/Tools/SSP*_interventions.yml). Where each would attach:

* Pre-allocation -- patch-geometry adjustments. Already reachable: write a modified
  alloc_params_t row set onto the scenario run instead of inheriting run 0's. Recipe and the
  two conversion traps (Relative vs Absolute; isometry -> elongation) are in 080, next to where
  the parameters are estimated.

* Allocation -- per-cell transition-probability adjustment inside a mask. The evoland analogue
  is trans_pot_t: predict potentials, rewrite the masked cells, then allocate without
  re-predicting. The per-SSP rules are all of the form

      Prob_adjust_type: Absolute   -> set the potential inside/outside the mask to a value
      Prob_adjust_type: Relative   -> raise cells above the Prob_adjust_intervention_percentile
                                      of the in-mask potential distribution and/or lower those
                                      below the non-intervention percentile outside it,
                                      subject to Prob_adjust_threshold and Prob_adjust_valency

  and would run per period, ordered by Intervention_ranking, restricted to
  Transition_target_classes (Urban in every file we have) and optionally From_lulc_filter.

  **This is currently not possible without an upstream change.** alloc_clumpy_one_period()
  calls predict_trans_pot() unconditionally as its first act and that write upserts
  trans_pot_t, so any hand-edited potentials are overwritten before they are read -- including
  under use_parent_trans_pot, which only redirects *which* run is predicted for. What is needed
  is either a `skip_prediction` flag or a hook applied between prediction and
  adjusted_trans_pot_v(). Note also that the adjustment interacts with the column scaling in
  adjusted_trans_pot_v(): that view rescales every transition's potentials so their mean matches
  the target rate, so an intervention that raises potentials inside a mask does not raise the
  amount of change, it only moves it -- which is the intended semantics, but means "Absolute
  100" cannot be read as "these cells convert".

* Post-allocation -- direct edit of the simulated map. Trivial here: rewrite lulc_data_t for the
  run and period after allocation, before the next period is allocated (so the edit propagates
  through the chain). No upstream change needed.

Masks. intrv_meta_t / intrv_masks_t exist upstream to hold exactly this (metadata + the
id_coord set per intervention), and nothing in the allocator reads them yet. Before they can be
filled, the mask sources need reproducible replacements: the YAML points at ValPar-local paths
(Data/Spat_prob_perturb_layers/Bulding_zones/BZ_raster.grd, .../Municipality_typology/...),
which is the same provenance problem the 02x ingestion steps solved for the predictors.

Conservation interventions are a further step again: SSP1's Conservation_expansion_and_
preservation is not a static mask but a spatial optimisation (Ca_expansion_target,
Ca_prioritization, Ca_patch_preference, Ca_expansion_rate), implemented in the original as
Scripts/Functions/identify_CAs_by_target_and_configuration.R and friends. That is a port, not a
translation, and should be scoped separately.
-->

## Did the allocation deliver the scenario?

`trans_rate_areas()` replays the rate table forward from the observed anchor state analytically —
the trajectory `07` intended. Comparing it against the realised areas isolates the allocator's
contribution: any gap is the allocator failing to place the demanded change, not the LP or the
demand.

This is the check `07` explicitly defers to this step. Its concern is concrete: the LP is allowed
to exceed the historic rate envelope, `adjusted_trans_pot_v()` scales potentials by an unbounded
`rate / mean_value` factor, the row-closure branch then clips the result, and uPAM's pool empties
before the quota is met. The scenario under-delivers its own demand, silently.

```{r}
#| label: extrapolate-verify
db$id_run <- NULL
realised_areas <- db$fetch(
  "lulc_data_t",
  where = glue::glue(
    "id_run in ({toString(realisation_runs$id_run)}) ",
    "and id_period in ({toString(alloc_periods)})"
  )
)[, .(area_realised = as.numeric(.N)), by = .(id_run, id_period, id_lulc)]
db$id_run <- 0L

predicted_areas <- rbindlist(lapply(scenario_runs$id_run, function(r) {
  replay <- trans_rate_areas(
    lulc_data = lulc_anchor,
    rates = as_trans_rates_t(rates_available[id_run == r & id_period %in% alloc_periods]),
    trans_meta = db$trans_meta_t
  )
  as.data.table(replay)[, parent_id_run := r][]
}))

area_check <- merge(
  realisations[realised_areas, on = "id_run"],
  predicted_areas[, .(parent_id_run, id_period, id_lulc, area_predicted = area)],
  by = c("parent_id_run", "id_period", "id_lulc"),
  all.x = TRUE
)
area_check[, `:=`(
  drift = area_realised - area_predicted,
  drift_pct = 100 * (area_realised - area_predicted) / area_predicted
)]

# Terminal period is where the drift has compounded the most; that is the number to report.
area_check[id_period == max(alloc_periods)][
  db$lulc_meta_t,
  on = "id_lulc",
  nomatch = NULL
][order(-abs(drift_pct))][1:15][, .(
  ssp, replicate, name,
  area_realised, area_predicted = round(area_predicted),
  drift = round(drift), drift_pct = round(drift_pct, 2)
)]
```

```{r}
#| label: extrapolate-verify-summary
drift_summary <- area_check[, .(
  max_abs_drift_pct = max(abs(drift_pct), na.rm = TRUE),
  mean_abs_drift_pct = mean(abs(drift_pct), na.rm = TRUE)
), by = .(ssp, id_period)]

if (max(drift_summary$max_abs_drift_pct) > 5) {
  warning(
    "Realised areas depart from the solved trajectory by more than 5 % ",
    "for some (scenario, period, class). The demand is not being delivered; ",
    "check pct_flow_above_max in 070 before interpreting any map."
  )
}
dcast(drift_summary, ssp ~ id_period, value.var = "max_abs_drift_pct")
```

## Trajectories

A compact numeric trace of where each scenario ends up, so a mis-specified rate table is visible
here rather than in `090d`'s maps. Observed periods come from run `0`, the projected ones from
the realisations.

```{r}
#| label: extrapolate-trajectory
observed_areas <- db$fetch(
  "lulc_data_t",
  where = glue::glue("id_run = 0 and id_period <= {anchor_period} and id_period > 0")
)[, .(area = as.numeric(.N)), by = .(id_period, id_lulc)][, scenario := "observed"]

projected_areas <- realisations[realised_areas, on = "id_run"][
  ,
  .(area = mean(area_realised)),
  by = .(id_period, id_lulc, scenario = ssp)
]

trajectory <- rbind(observed_areas, projected_areas)[
  db$lulc_meta_t,
  on = "id_lulc",
  nomatch = NULL
]

dcast(trajectory, name + scenario ~ id_period, value.var = "area")
```

## Ensemble spread

Where the realisations of one scenario disagree, the allocator — not the scenario — is deciding.
`090d` turns this into the change-frequency map; the number here is the per-class coefficient of
variation across replicates at the terminal period, which is enough to see whether
`n_replicates` is large enough to be worth averaging.

```{r}
#| label: extrapolate-spread
spread <- realisations[realised_areas, on = "id_run"][
  id_period == max(alloc_periods),
  .(
    mean_area = mean(area_realised),
    sd_area = sd(area_realised),
    cv_pct = 100 * sd(area_realised) / mean(area_realised)
  ),
  by = .(ssp, id_lulc)
][db$lulc_meta_t, on = "id_lulc", nomatch = NULL]

spread[order(-cv_pct)][, .(ssp, name, mean_area = round(mean_area), cv_pct = round(cv_pct, 3))]
```

```{r}
#| label: extrapolate-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key = , value = ,
  "extrapolate_ssps", paste(sort(unique(scenario_runs$ssp)), collapse = ","),
  "extrapolate_climate_framings",
  paste(sort(unique(scenario_runs$climate_framing)), collapse = ","),
  "extrapolate_periods", paste(alloc_periods, collapse = ","),
  "extrapolate_n_scenario_runs", as.character(nrow(scenario_runs)),
  "extrapolate_n_replicates", as.character(n_replicates),
  "extrapolate_seed", as.character(seed),
  "extrapolate_alloc_params_from", "run 0 (selected in 080-validate-backcasting.qmd)",
  "extrapolate_interventions", "none applied",
  "extrapolate_max_area_drift_pct",
  as.character(signif(max(drift_summary$max_abs_drift_pct), 4))
))
```
