---
title: "091 — Stochastic allocation to 2030"
---

One extrapolated period, allocated **20 times per SSP**, gathered into a per-cell **change
intensity**: the share of the ensemble in which a cell leaves its last observed class. Five
scenarios, five maps.

This is deliberately the smallest useful forward experiment. `090` walks the whole trajectory to
2060 and is dominated by compounding; here the anterior state is the same observed map for every
member of every scenario, so the only things that differ are the SSP's demand (from `070`) and
the allocator's RNG. What the maps show is therefore *where each scenario's demand insists on
change*, with the allocator's arbitrariness averaged out rather than hidden inside one draw.

Two properties of the design are worth stating up front, because they are what make 100
allocations affordable:

- **The transition potentials are computed once per scenario, not once per member.** The
  anterior state is identical across an SSP's members, so the potential surface is too.
  `use_parent_trans_pot = TRUE` (exposed upstream in `ca4d681`) points every member at its
  parent scenario run's `trans_pot_t`, so the 20 members share one prediction — the pattern the
  `stochastic-allocation-sensitivity` vignette now uses.
- **The single-period allocator is called directly**, bypassing the multi-period
  `alloc_clumpy()` wrapper. See the allocation section for why; it is the difference between
  this step being cheap and it being unrunnable.

```{r}
#| label: alloc2030-params
library(data.table)
library(evoland)

climate_framing_sel <- "current" # the leaves under id_run 100; see 001-setup-db.qmd
n_members <- 20L # stochastic allocations per SSP
seed_base <- 2030L
select_score <- "classif.auc" # must match what 060 stored in crossval_score
select_maximize <- TRUE
n_workers <- 20L # future workers for the prediction phase
member_id_factor <- 1000L # id_run = scenario * factor + member
member_kind <- "stochastic single-period member"

db <- evoland_db$new(path = "ssp-ch.evolanddb")
db$id_run <- 0L

anchor_period <- max(db$periods_t[is_extrapolated == FALSE & id_period > 0, id_period])
target_period <- min(db$periods_t[is_extrapolated == TRUE, id_period])
target_label <- format(as.Date(db$periods_t[id_period == target_period, end_date]), "%Y")

stopifnot(
  "the target period must follow the anchor directly" = target_period == anchor_period + 1L,
  "no viable transitions - run 040 and 060" =
    nrow(db$trans_meta_t[is_viable == TRUE]) > 0L,
  "no fitted models - run 060-transition-modelling.qmd first" =
    db$row_count("trans_models_t") > 0L
)

message(
  "Allocating period ", target_period, " (to ", target_label, ") ",
  n_members, " times per scenario, from observed period ", anchor_period
)
```

Why this period is "2030": `001` puts the first extrapolated period at 2020–2029, immediately
after the last Arealstatistik survey. `target_label` is read off `periods_t` rather than
hard-coded, so the title is wrong rather than the code if the period table changes.

## The five scenario runs

One leaf per SSP under the chosen climate framing. `current` inherits the `id_period 0` climate
baseline, so this step does not depend on `051-ingest-preds-ch2025-3-gwl.qmd` having run. Over a
single 2020s step that costs little: every GWL trajectory in `001` sits at 1.5–2.0 °C in `p5`,
so the climate signal between the leaves is near-nil at this horizon. Switching to the per-SSP
`default_for_ssp` trajectories is a change to `climate_framing_sel` — but check that the
projected predictors actually exist for those runs first, because a missing projection falls
back to the baseline silently.

```{r}
#| label: alloc2030-runs
runs <- db$runs_t

scenario_runs <- runs[!is.na(ssp) & climate_framing == climate_framing_sel]

# Members and realisations written by this step or by 090 inherit `ssp` / `climate_framing`
# from their parent; they are outputs, not scenarios. Tested outside the `[` because
# data.table evaluates the whole `i` expression whether or not the column exists.
if ("kind" %in% names(scenario_runs)) {
  scenario_runs <- scenario_runs[
    is.na(kind) | !kind %in% c("extrapolation realisation", member_kind)
  ]
}
scenario_runs <- scenario_runs[order(ssp)]

stopifnot(
  "no scenario runs for this climate framing" = nrow(scenario_runs) > 0L,
  "expected exactly one run per SSP" = uniqueN(scenario_runs$ssp) == nrow(scenario_runs)
)

scenario_runs[, .(id_run, parent_id_run, ssp, climate_framing, description)]
```

## Allocation parameters: the unperturbed estimate from `080`

`080` registers one run per parameter set, the lowest-numbered of which is the unperturbed
estimate straight out of `create_alloc_params_t()` — the perturbed siblings exist only to be
compared against it. We take that set and write it onto each scenario run, so the members
inherit it through the lineage.

Writing to the scenario runs rather than to run `0` is deliberate: it leaves whatever `080`
selected on run `0` alone, and it keeps this step's parameter choice visible in the run tree
instead of implicit in a shared ancestor.

```{r}
#| label: alloc2030-alloc-params
stopifnot(
  "no backcast parameter runs - run 080-validate-backcasting.qmd first" =
    "kind" %in% names(runs) && nrow(runs[kind == "backcast parameters"]) > 0L
)

alloc_param_run <- min(runs[kind == "backcast parameters", id_run])
db$id_run <- alloc_param_run
unperturbed_params <- db$alloc_params_t
db$id_run <- 0L

stopifnot(
  "the unperturbed parameter run carries no rows" = nrow(unperturbed_params) > 0L,
  "allocation parameters are missing for some viable transition" =
    all(db$trans_meta_t[is_viable == TRUE, id_trans] %in% unperturbed_params$id_trans)
)

for (r in scenario_runs$id_run) {
  db$alloc_params_t <- as_alloc_params_t(copy(unperturbed_params)[, id_run := r])
}
db$id_run <- 0L

message("Using allocation parameters from run ", alloc_param_run)
unperturbed_params[, .(id_trans, mean_patch_size, patch_elongation, frac_expander)]
```

## Rates for the single step

Each scenario needs a rate for every viable transition at the target period, or
`adjusted_trans_pot_v()` drops that transition (it inner-joins potentials to rates) and it
simply does not happen. `070` writes these.

```{r}
#| label: alloc2030-rates
db$id_run <- NULL # fetch() is lineage-filtered; NULL reads across runs
rates <- db$fetch(
  "trans_rates_t",
  where = glue::glue(
    "id_run in ({toString(scenario_runs$id_run)}) and id_period = {target_period}"
  )
)
db$id_run <- 0L

viable_trans <- db$trans_meta_t[is_viable == TRUE]

# The demanded quantity is derived from `rate` x anterior-class size, not from the stored
# `count`. `count` means different things depending on which producer wrote the row --
# extrapolate_trans_rates() multiplies the rate by the *total* cell count, the LP solver writes
# actual flows -- whereas `rate` is what the allocator consumes: adjusted_trans_pot_v() scales
# the potentials so their mean over the anterior class equals it.
anchor_class_sizes <- db$get_query(glue::glue(
  "select id_lulc as id_lulc_anterior, count(*) as n_ant
     from {db$get_read_expr('lulc_data_t')}
     where id_period = {anchor_period}
     group by id_lulc"
))

rate_coverage <- Reduce(
  function(x, y) merge(x, y, by = "id_run"),
  list(
    scenario_runs[, .(id_run, ssp)],
    merge(
      merge(
        rates[id_trans %in% viable_trans$id_trans],
        viable_trans[, .(id_trans, id_lulc_anterior)],
        by = "id_trans"
      ),
      anchor_class_sizes,
      by = "id_lulc_anterior"
    )[, .(n_trans = uniqueN(id_trans), demand_cells = sum(rate * n_ant)), by = id_run]
  )
)

stopifnot(
  "no rates at the target period - run 070-transition-rates-solver.qmd first" =
    nrow(rate_coverage) > 0L,
  "some scenario run has no rates at the target period" =
    nrow(rate_coverage) == nrow(scenario_runs)
)

if (any(rate_coverage$n_trans < nrow(viable_trans))) {
  warning(
    "Some scenario carries rates for fewer than the ", nrow(viable_trans),
    " viable transitions; the missing ones cannot be allocated."
  )
}
rate_coverage[order(ssp)]
```

`demand_cells` is the number of cells each scenario asks to convert in this one step. It is the
quantity the change-intensity maps below should integrate to, and the comparison at the end is
the check that the allocator delivered it.

## Register the members

20 children per scenario, differing only in RNG state. They hold nothing but their simulated
`lulc_data_t` for the target period; rates, predictors, potentials and allocation parameters all
resolve to the parent.

```{r}
#| label: alloc2030-register
members <- scenario_runs[, .(member = seq_len(n_members)), by = .(parent = id_run, ssp)][, .(
  id_run = parent * member_id_factor + member,
  parent_id_run = parent,
  description = paste0(
    ssp, " single-period stochastic member ", member, " (to ", target_label, ")"
  ),
  kind = member_kind,
  ssp,
  member
)][, seed := seed_base + .I][]

# Re-running replaces this step's own members; anything else with these ids is a hard error.
mine <- if ("kind" %in% names(runs)) runs[kind == member_kind, id_run] else integer(0)
clash <- setdiff(intersect(members$id_run, runs$id_run), mine)
stopifnot(
  "member run ids collide with runs registered elsewhere" = length(clash) == 0L,
  "n_members must be below the id_run factor" = n_members < member_id_factor,
  "member ids must stay within integer range" = max(members$id_run) < .Machine$integer.max
)

db$commit(
  as_runs_t(rbind(runs[!id_run %in% members$id_run], members, fill = TRUE)),
  "runs_t",
  "overwrite" # adding columns
)
members[, .N, by = .(ssp, parent_id_run)]
```


## Allocate

`use_parent_trans_pot = TRUE` makes each member read its scenario's potentials; the call to
`predict_trans_pot()` inside the allocator then finds them present and returns without work after the first iteration.

`alloc_clumpy_one_period()` is currently internal, hence the `:::`. May want to change upstream: either an
`update_neighbors` argument on `alloc_clumpy()`, or skipping the upsert after the last period of
the requested sequence, since nothing consumes it. Recorded in `TODO.md`.

```{r}
#| label: alloc2030-allocate
#| output: false
options("evoland.use_prefetch_predict" = TRUE)
for (i in seq_len(nrow(members))) {
  db$id_run <- members$id_run[i]

  # The `seed` argument was removed upstream in abc8bca; CLUMPY draws through R's RNG
  # (src/alloc_clumpy.cpp), so seeding here is exact and per-member reproducible.
  set.seed(members$seed[i])

  lulc_member <- evoland:::alloc_clumpy_one_period(
    self = db,
    id_period_post = target_period,
    select_score = select_score,
    select_maximize = select_maximize,
    use_parent_trans_pot = TRUE
  )

  db$commit(lulc_member, "lulc_data_t", method = "upsert")
}
db$id_run <- 0L
```

## Gather the ensemble into change intensity

The gathering happens in DuckDB rather than in R: the member maps are 20 × 5 × 4.1 M rows, which
is not something to pull into memory to compute a mean over. The query joins every member cell
against the observed anchor and returns one row per (SSP, cell).

```{r}
#| label: alloc2030-gather
db$id_run <- NULL # read across runs

intensity <- db$get_query(glue::glue(
  r"{
  with sim as (
    select id_run, id_coord, id_lulc
    from {db$get_read_expr('lulc_data_t')}
    where id_period = {target_period}
      and id_run in ({toString(members$id_run)})
  ),
  anchor as (
    select id_coord, id_lulc
    from {db$get_read_expr('lulc_data_t')}
    where id_period = {anchor_period} and id_run = 0
  ),
  member_ssp as (
    select id_run, ssp
    from {db$get_read_expr('runs_t')}
    where id_run in ({toString(members$id_run)})
  )
  select
    m.ssp,
    s.id_coord,
    count(*) as n_members,
    sum(case when s.id_lulc <> a.id_lulc then 1 else 0 end) as n_changed
  from sim s
  join anchor a on a.id_coord = s.id_coord
  join member_ssp m on m.id_run = s.id_run
  group by m.ssp, s.id_coord
  }"
))
db$id_run <- 0L

intensity[, p_changed := n_changed / n_members]

stopifnot(
  "a cell is missing members - did every allocation write?" =
    all(intensity$n_members == n_members),
  "expected one row per (SSP, cell)" =
    nrow(intensity) == uniqueN(intensity$ssp) * uniqueN(intensity$id_coord)
)

intensity[, .(
  cells = .N,
  mean_intensity = round(mean(p_changed), 5),
  cells_ever = sum(p_changed > 0),
  cells_always = sum(p_changed == 1),
  cells_majority = sum(p_changed >= 0.5)
), by = ssp][order(ssp)]
```

`cells_ever` against `cells_always` is the useful contrast. Their ratio is how much larger the
*envelope* of possible change is than its robust core: a scenario where the two are close is one
whose demand can only be satisfied in essentially one place, while a wide gap means the
potential surface is flat enough over the candidate pool that the allocator's draw decides.

## Did the ensemble deliver the demanded quantity?

Summing `p_changed` over all cells gives the expected number of cells that change, which should
match the demand `070` solved for this step. This is the same under-delivery check `090` runs
against `trans_rate_areas()`, in its cheapest form.

```{r}
#| label: alloc2030-quantity
delivered <- merge(
  intensity[, .(expected_changed = sum(p_changed)), by = ssp],
  rate_coverage[, .(ssp, demand_cells)],
  by = "ssp"
)
delivered[, `:=`(
  shortfall = demand_cells - expected_changed,
  shortfall_pct = round(100 * (demand_cells - expected_changed) / demand_cells, 2)
)][]

if (any(abs(delivered$shortfall_pct) > 5)) {
  warning(
    "Expected change departs from the solved demand by more than 5 % for some scenario. ",
    "Check pct_flow_above_max in 070 before reading the maps."
  )
}
delivered[order(-abs(shortfall_pct))]
```

The demand counts and the realised change are not the same quantity when a cell is the target of
two transitions in one step, or when a transition's anterior class is itself being converted, so
a small gap is expected even from a perfect allocator. A large one is the pool-exhaustion failure
`070` warns about.

## The maps

`tabular_to_raster()` gathers the long table into one layer per SSP. All five panels share a
`0–1` scale so they are readable against each other, which is the whole point of putting them
side by side.

```{r}
#| label: alloc2030-maps
#| fig-width: 14
#| fig-height: 8
intensity_rast <- tabular_to_raster(
  intensity[, .(id_coord, ssp, p_changed)],
  coords = db$coords_minimal,
  value_col = "p_changed"
)
# tabular_to_raster names layers after the grouping column ("ssp_SSP1"); the panel titles
# should just be the scenario.
names(intensity_rast) <- sub("^ssp_", "", names(intensity_rast))

# workaround: write out so i can fetch it from HPC node to do plotting locally for now
terra::writeRaster(
  intensity_rast,
  "2026-05-ssp-ch/091-change-intensity-2030.tif",
  gdal = c("COMPRESS=DEFLATE"),
  overwrite = TRUE
)
intensity_rast <- terra::rast("2026-05-ssp-ch/091-change-intensity-2030.tif")

# TODO the terra built-in plots are not nice to look at, use ggplot2 with ggspatial added.
# use a white background, also for NAs and 0 intensity. use a white-to-black color scale that shows
# intensity on a scale of (0, 0.65] with every intensity above clipping to the darkest colour.
# because the patterns need to be visible on an a4 print in a 18cm wide 9cm high plot with 5 panels comparing the scenarios, only show a box of roughly
# 2'658'159.81, 1'301'204.30
# 2'755'025.95, 1'077'643.13
# so we can see enough detail. the ensuing plot should be returned in 300dpi png
intensity_palette <- grDevices::hcl.colors(64, "Inferno", rev = TRUE)

# terra::panel() rather than terra::plot(): it draws one shared legend for the whole stack
# instead of repeating an identical one per panel, which is the difference between five maps
# that can be compared and five maps that merely sit next to each other.
plot_intensity <- function() {
  terra::panel(
    intensity_rast,
    col = intensity_palette,
    # `type` must be forced: p_changed takes only n_members + 1 distinct values, and terra
    # otherwise falls back to a categorical legend, which mislabels the scale and (with few
    # members) fails to shade the panels at all.
    type = "continuous",
    range = c(0, 1),
    nc = 3,
    axes = FALSE,
    loc.main = "topleft",
    plg = list(title = "p(change)")
  )
}

plot_intensity()

# Paths are project-root relative: _quarto.yml sets execute-dir: project.
grDevices::png(
  "2026-05-ssp-ch/091-change-intensity-2030.png",
  width = 2400,
  height = 1400,
  res = 150
)
plot_intensity()
invisible(grDevices::dev.off())
```

A cell reads `0` where no member changed it and `1` where all 20 did. Because the anterior map is
the same everywhere, the differences between panels are attributable to the SSP demand alone —
`070`'s rate vectors are the only scenario-varying input at this horizon.

## Where the scenarios disagree

The five panels answer "where does each scenario change things". The complementary question is
where they *differ*, which is a cheap derived layer worth having next to them: the spread across
the five scenario means at each cell.

```{r}
#| label: alloc2030-disagreement
#| fig-width: 6
#| fig-height: 5
# Cell-wise across layers via terra's Summary methods, not app(): app() with an R closure
# evaluates per cell and is far too slow at 4.1 M.
disagreement <- max(intensity_rast) - min(intensity_rast)
terra::writeRaster(
  disagreement,
  "2026-05-ssp-ch/091-change-intensity-2030-disagreement.tif",
  gdal = c("COMPRESS=DEFLATE"),
  overwrite = TRUE
)
disagreement <- terra::rast("2026-05-ssp-ch/091-change-intensity-2030-disagreement.tif")
terra::plot(
  disagreement,
  col = intensity_palette,
  type = "continuous",
  range = c(0, 1),
  axes = FALSE,
  main = paste0("Between-scenario range in change intensity, ", target_label)
)

terra::global(disagreement, c("mean", "max"), na.rm = TRUE)
```

```{r}
#| label: alloc2030-report
db$reporting_t <- as_reporting_t(data.table::rowwiseDT(
  key = , value = ,
  "alloc2030_target_period", as.character(target_period),
  "alloc2030_target_label", target_label,
  "alloc2030_anchor_period", as.character(anchor_period),
  "alloc2030_climate_framing", climate_framing_sel,
  "alloc2030_scenarios", paste(scenario_runs$ssp, collapse = ","),
  "alloc2030_n_members", as.character(n_members),
  "alloc2030_seed_base", as.character(seed_base),
  "alloc2030_alloc_param_run", as.character(alloc_param_run),
  "alloc2030_max_shortfall_pct", as.character(max(abs(delivered$shortfall_pct))),
  "alloc2030_interventions", "none applied"
))
```
